---
title: "Annotation membership QC — sf point-in-polygon vs Out_of_annotation flag"
author: "sceriff0"
date: "2026-07-07"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(sf)
})

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 8, fig.height = 5,
                      dev = c("png", "pdf"))   # PNG for the site, PDF exported to output/

# Loaders (gives neoplastic_data + the helpers). test.R prints colnames(); silence it.
# validation_helpers.R also sources code/plot_theme.R, which applies the house
# figure style (theme + palettes + geom defaults) for every plot below.
invisible(capture.output(suppressMessages(source(here("code", "test.R")))))
source(here("code", "validation_helpers.R"))
source(here("code", "pdf_export.R"))          # export_pdf_figures(): plots -> output/figures/
```

## Starting data and notation

Every value below is derived from two inputs, compared **per cell**:

- `<patient>_a<k>.csv` — FlowPath per-annotation cell exports (one row per cell)
  under the project's `data/flowpath/per_annotation`, carrying **both** the cell
  `centroid` and the precomputed `Out_of_annotation` flag for the same cells. Only
  some patients have these.
- the pathologist annotation **geojson** polygons (`ANNOTATION_k`, in pixels),
  loaded via `load_annotations()` for the patients in `neoplastic_data$SAMPLE`.

The main pipeline decides "inside the annotation" by **sf point-in-polygon** on the
pathologist geojson, mapping cell centroids into the pixel frame with the fixed
`centroid / 0.325`. This report cross-checks that against the shipped
`Out_of_annotation` flag. Because the per-annotation csv carries *both* the
centroid and the flag for the **same cells**, agreement is a direct per-cell
comparison (no cross-file matching):

```
flag inside = Out_of_annotation is FALSE
sf inside   = st_within( centroid / 0.325 , the ANNOTATION_k geojson polygon )
```

Per `(patient, annotation)` we report the two set sizes, their overlap (Jaccard and
% of cells that agree), and the resulting tumour fraction under each method. A large
flag-vs-sf gap on a slide means the fixed `0.325` conversion is wrong **for that
slide** — sf is landing on the wrong cells.

## Compute agreement

```{r qc}
# per-annotation flag csvs (only some patients have these); matches test.R's path style.
PER_ANNOT_DIR <- here("data", "flowpath", "per_annotation")

annots <- load_annotations(patient_ids = neoplastic_data$SAMPLE)

if (dir.exists(PER_ANNOT_DIR) && length(fs::dir_ls(PER_ANNOT_DIR, glob = "*.csv"))) {
  qc <- annotation_membership_qc(PER_ANNOT_DIR, annots, um_per_px = 0.325)
} else {
  qc <- tibble::tibble()
  cat("No per-annotation csvs found at", PER_ANNOT_DIR, "\n")
}

if (nrow(qc)) {
  cat("annotations checked:", nrow(qc),
      "| mean cell-level agreement:",
      sprintf("%.1f%%", 100 * mean(qc$pct_cells_agree, na.rm = TRUE)),
      "| mean Jaccard:", sprintf("%.2f", mean(qc$jaccard, na.rm = TRUE)), "\n")
}
```

## Per-annotation agreement table

`pct_cells_agree` = fraction of cells given the same inside/outside call by both
methods. `jaccard` = |flag ∩ sf| / |flag ∪ sf| of the inside sets. The last two
columns are the tumour-fraction-inside each method would feed into the concordance
plots — they should be close if membership agrees.

```{r qc-table}
if (nrow(qc)) {
  qc |>
    mutate(across(c(jaccard, pct_cells_agree, tumor_frac_flag, tumor_frac_sf),
                  ~ round(.x, 3))) |>
    arrange(jaccard) |>
    knitr::kable(caption = paste("sf vs Out_of_annotation flag, per annotation.",
                                 "Low jaccard / pct_cells_agree = the /0.325 sf",
                                 "disagrees with the flag on that slide."))
}
```

## Tumour fraction: flag vs sf

Each point is one annotation; the dashed line is perfect agreement. Points off the
line are annotations where the two membership definitions give a different tumour
fraction — those are the slides to distrust the sf conversion on.

```{r qc-scatter}
if (nrow(qc) && sum(is.finite(qc$tumor_frac_flag) & is.finite(qc$tumor_frac_sf)) > 1) {
  d  <- qc |> filter(is.finite(tumor_frac_flag), is.finite(tumor_frac_sf))
  cc <- paired_cor3(d$tumor_frac_flag, d$tumor_frac_sf)   # Pearson + Spearman + Kendall
  ggplot(d, aes(tumor_frac_flag, tumor_frac_sf, colour = patient_id)) +
    geom_abline(slope = 1, linetype = "dashed", colour = REF_LINE) +
    geom_point(size = 3, alpha = 0.85) +
    coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
    annotate("text", x = 0, y = 1, hjust = 0, vjust = 1,
             label = sprintf("Pearson r = %.2f\nSpearman rho = %.2f\nKendall tau = %.2f\n(n = %d)",
                             cc$pearson, cc$spearman, cc$kendall, cc$n)) +
    labs(title = "Tumour fraction inside: flag vs sf point-in-polygon",
         x = "tumour fraction (Out_of_annotation flag)",
         y = "tumour fraction (sf / 0.325)", colour = "patient")
}
```

## Notes

- Only patients with per-annotation csvs **and** a matching geojson are checked; a
  `note` column flags the ones skipped (missing geojson or columns).
- High agreement here validates the fixed `0.325` conversion for those slides; a
  slide with low agreement should either be excluded or given its own conversion.
- This compares *membership*, not the pathologist — that concordance is in
  `clinical_data.Rmd`.

```{r export-figures, include=FALSE}
# Copy every PDF knitr emitted for this analysis into output/figures/<slug>/.
export_pdf_figures("annotation_membership_qc")
```
