---
title: "benchmarks"
author: "sceriff0"
date: "2026-07-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: tidyverse + scales + here (in the lockfile). Figure logic is vendored from
# the mirage benchmarking branch (code/benchmark_plots.R).
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
})
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      dev = c("png", "pdf"))   # PNG for the site, PDF exported to output/
source(here("code", "pdf_export.R"))          # export_pdf_figures(): plots -> output/figures/

bench_dir  <- here("data", "benchmark")
have_bench <- file.exists(file.path(bench_dir, "measurements.csv"))
```

## Starting data and notation

Mirage image-processing **pipeline benchmark** — resource scaling, registration
cost/accuracy, and segmentation-method comparison from the SLURM sweep. Every
figure is built by sourcing `code/benchmark_plots.R` (vendored from
`benchmarks/analysis/plots.R`) and rendered **inline from the sweep CSVs** — no
PNGs on disk. This report is a formula index for those figures: for each one it
states the raw columns consumed and the exact reduction to the plotted value.

The core input is **`data/benchmark/measurements.csv`** — one row per
run-and-process. `proc` is the `process` string with everything up to the last
`:` stripped. Per row we read three quantities:

```
peak_rss_gb = peak resident memory of that process   (GiB)
realtime_s  = wall time of that process              (s)
input_gb    = input size of that run                 (GiB)
```

plus every swept parameter and the label `varied_axis` naming which knob that run
varied off baseline. Size-scaling figures keep only the runs whose `varied_axis`
is one of the `size_axes`:

```
size_axes = baseline, scaling_grid, registration_grid,
            distributed_grid, target_px, n_channels
```

Registration-stage RAM is aggregated over the registration **leaf** processes:

```
reg_leaves = REGISTER, REG_PREP, REG_TILE, REG_NONRIGID,
             REG_MICRO_PREP, REG_FINALIZE, and similar
```

and boolean sweep flags are read as "truthy" when the lower-cased value is one of
`true`, `1`, or `yes`.

**Optional CSVs** (a figure needing an absent file is skipped): `resource_stats.csv`,
`quality.csv`, `run_cost.csv`, `segmentation_agreement.csv`,
`classic_vs_distributed_registration.csv`, `registration_drift.csv`. Accuracy
columns (`reg_tre_median_px`, `max_abs_delta`, `pct_pixels_diff`,
`rss_saving_gb`, `peak_rss_gb_mean`/`peak_rss_gb_std`, `instance_f1`,
`cell_count_ratio`) are **precomputed upstream** and only read here — see *Notes*.

```{r run-benchmark}
bench_figs <- list()
if (have_bench) {
  # sourcing builds `bench_figs` (a named list of ggplots) from the sweep CSVs
  source(here("code", "benchmark_plots.R"))
  cat(length(bench_figs), "figure(s) built from", basename(bench_dir), "data.\n")
} else {
  cat("No `data/benchmark/measurements.csv` found — add the sweep CSVs and re-knit.")
}
```

## Metrics

Every reduction below lives in `code/benchmark_plots.R`; the cited function/object
name lets a reader trace raw column -> plotted value.

**Power-law scaling** (`01_memory_scaling_per_process`, `02_time_scaling_per_process`;
`powerlaw` + `powerlaw_plot`). Over rows whose `varied_axis` is in `size_axes`,
with `input_gb` > 0 and y > 0, per process an OLS line is fit in log-log space:

```
fit  log10(y) = a + exponent * log10(input_gb)   (OLS)
report exponent (slope) and R^2

where:
- y = peak_rss_gb for fig 01, realtime_s for fig 02
- the drawn line is  y = 10^(a + exponent * log10(input_gb))
                     across the min..max of input_gb
```

Reading: exponent = 1 linear, exponent > 1 super-linear, exponent < 1 sub-linear.

**Classic vs distributed registration RAM** (`03_classic_vs_distributed_ram`,
`03b_distributed_ram_saving`; from `classic_vs_distributed_registration.csv`).
Columns `reg_peak_rss_gb_classic` and `reg_peak_rss_gb_distributed` are plotted
directly vs `target_px`, faceted by `n_channels`. The saving is the precomputed
column:

```
rss_saving_gb = reg_peak_rss_gb_classic - reg_peak_rss_gb_distributed   (GiB)
```

**N-image registration RAM** (`04_nimage_registration_ram`). `REGISTER` rows on
`registration_grid`/`baseline`/`scaling_grid`, replicate mean per configuration:

```
value = mean of peak_rss_gb over REGISTER runs
        grouped by (target_px, n_channels, n_register_images)
```

**OFAT knob effects** (`05_ofat_knob_effects`). For each knob in `knob_targets`
(an `(axis, proc, metric)` triple), keep rows whose `varied_axis` is the knob's
axis or `baseline` and whose `proc` matches, then average the metric per distinct
knob value:

```
value = mean of realtime_s per distinct knob value
```

**Replicate variance** (`06_replicate_variance`; from `resource_stats.csv`, rows
with `n_reps` > 1). Bars are the precomputed `peak_rss_gb_mean`; error bars span
`peak_rss_gb_mean` minus/plus `peak_rss_gb_std`.

**Stage memory heatmap** (`07_stage_memory_heatmap`). Rows on `size_axes` with
`n_channels` = 2 and `n_register_images` = 2; tile value is the replicate mean:

```
tile = mean of peak_rss_gb per (proc, target_px)   (log colour scale)
```

**Channel-count effect** (`08_channel_effect`). Points of `input_gb` vs
`peak_rss_gb` for `proc` in REGISTER, PREPROCESS, SEGMENT, QUANTIFY on
`size_axes`, `input_gb` > 0; the overlaid trend is `geom_smooth(method="lm")` fit
on the log10-vs-log10 axes, coloured by `n_channels`. No scalar emitted.

**Segmentation methods** (`09_segmentation_methods`, `09b_stardist_tile_grid`).
On `segmentation_grid*` / `proc = SEGMENT`: fig 09 is the raw `realtime_s`
distribution (box = IQR) per `seg_method`. Fig 09b, `stardist` only, tiles the
mean of `peak_rss_gb` per `(seg_n_tiles_x, seg_n_tiles_y)`.

**Registration knobs, both paths** (`10_registration_params_both_paths`). On
`registration_param_grid` / `proc` in `reg_leaves`, the per-run registration peak
is the **max over leaves**, and the path is set from a flag:

```
reg_peak_gb per run = max of peak_rss_gb over the reg_leaves processes
path = distributed if reg_distributed_tiling is truthy, else classic

value = mean of reg_peak_gb
        per (memory_mode, skip_micro_registration, path)
```

**Registration accuracy vs cost** (`11_accuracy_vs_cost`). The accuracy metric,
`reg_tre_median_px` from `quality.csv` (feature-based **target registration
error**, median px per run — precomputed, see *Notes*), plotted vs
`cpu_hours` from `run_cost.csv`, joined on `run_id`; lower-left is better.

**Segmentation cell counts** (`12_segmentation_cell_counts`). Raw `n_cells`
(max mask label) from `quality.csv`, distribution per `seg_method`.

**Segmentation cross-method agreement** (`12b_segmentation_agreement`; from
`segmentation_agreement.csv`). Bar = `instance_f1` (IoU-matched per-cell F1 between
`method_a` and `method_b`, 1 = agree on every cell); text label = `cell_count_ratio`.
Both precomputed.

**End-to-end cost vs size** (`13_end_to_end_cost`; from `run_cost.csv`, `size_axes`).
Per `target_px`:

```
mean cpu_hours = mean of cpu_hours
mean wall clock hours = mean of wall_clock_s / 3600
```

**Bottleneck by size** (`14_bottleneck_by_size`; from `run_cost.csv`). Count runs
per `(target_px, bottleneck_stage)`, drawn `position="fill"`, so each bar is the
**share** of that stage at that size:

```
bar share = count(stage) / total count   at that target_px
```

**Tiled path granularity** (`15_tiled_path_granularity`). On
`distributed_tiling_grid` / `reg_leaves`, per run:

```
reg_peak_gb per run = max of peak_rss_gb over the reg_leaves processes
reg_time_s per run  = sum of realtime_s over the reg_leaves processes

value = mean of reg_peak_gb per (reg_dist_tile_wh, reg_dist_tile_buffer)
```

**Tiled path pixel drift** (`16_tiled_drift_from_classic`; from
`registration_drift.csv`, `path == "tiled"`). Mean of the precomputed
`max_abs_delta` (max absolute intensity difference vs the classic slide; 0 =
identical) and `pct_pixels_diff` per `(tile_wh, tile_buffer)`.

**Registration error by path** (`17_registration_error_by_path`; from `quality.csv`).
`reg_tre_median_px` distribution per `path`, where:

```
path = classic    if reg_distributed_tiling is not truthy
       tiled       if reg_dist_force_tiling is truthy
       separated   otherwise
```

## Figures

```{r figures, results="asis", fig.width=9.5, fig.height=6}
if (length(bench_figs)) {
  for (nm in names(bench_figs)) {
    title <- gsub("_", " ", sub("^[0-9]+[a-z]?_", "", nm))   # "01_memory_scaling" -> "memory scaling"
    cat("\n\n### ", title, "\n\n", sep = "")
    print(bench_figs[[nm]])
    cat("\n\n")
  }
} else if (have_bench) {
  cat("Benchmark data present but no figures were produced — check `measurements.csv` columns.")
}
```

## Notes

- **Accuracy metrics are read, not recomputed.** `reg_tre_median_px` (TRE),
  `max_abs_delta`/`pct_pixels_diff` (drift), `instance_f1`/`cell_count_ratio`
  (segmentation agreement), `rss_saving_gb`, and `peak_rss_gb_mean`/`peak_rss_gb_std`
  all arrive precomputed in their CSVs. The only reductions `benchmark_plots.R`
  itself performs are the log-log OLS (exponent, R^2), the per-run max/sum over
  `reg_leaves`, and replicate means/shares. The subtitles call TRE a "feature TRE
  proxy"; treat it as a proxy, not a ground-truth landmark error.
- **Reductions are means over replicates** unless noted; SLURM isolates per-process
  resources, so `peak_rss_gb` is one process's peak, and registration-stage RAM is
  the max across its leaf processes within a run.
- Any figure whose optional CSV (or required column) is missing is silently
  skipped; the rendered set depends on which sweep outputs were dropped into
  `data/benchmark/`.

```{r export-pdfs, include=FALSE}
# Copy every plot from this analysis (each rendered as a PDF via the
# dev = c("png", "pdf") chunk option) into output/figures/benchmarks/. Runs last so
# that all figures from the chunks above already exist on disk.
export_pdf_figures("benchmarks")
```
