---
title: "benchmarks"
author: "sceriff0"
date: "2026-07-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: tidyverse + here (in the lockfile). Optional: diptest for Hartigan's dip
# test (a bimodality-coefficient fallback runs if it is absent).
#   renv::install("diptest"); renv::snapshot()
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
})

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6)

invisible(capture.output(suppressMessages(source(here("code", "test.R")))))
source(here("code", "validation_helpers.R"))
theme_set(theme_paper)   # house style (mirage benchmarks)
CAP <- "ihc_method internal QC · FlowPath single-cell IHC"
```

## Purpose

**Internal QC of the IHC readout** — validation that needs no external reference,
only `ihc_data`. Traces a fault backwards through the pipeline:

1. **Phenotype ↔ marker concordance** — do cells labelled a type express the
   marker that defines it? (validates the phenotyping/gating)
2. **Per-channel usability** — is each marker's `_sign` a clean threshold on a
   bimodal signal? (validates the individual stains)
3. **Morphology by phenotype** — do cell shapes match their assigned type?
   (validates that segmentation captures real cells)

Lineage **leakage & exclusivity** (segmentation / optical spillover) now lives in
its own report — see [exclusivity.html](exclusivity.html).

```{r marker-long}
mk_long <- ihc_marker_long(ihc_data) |>
  filter(!is.na(phenotype_clean), phenotype_clean != "Unknown")
cat("cells:", dplyr::n_distinct(mk_long$cell_id),
    "| markers:", dplyr::n_distinct(mk_long$marker),
    "| phenotypes:", dplyr::n_distinct(mk_long$phenotype_clean), "\n")
```

## 1. Phenotype ↔ marker concordance

Mean marker z-score per phenotype. A trustworthy readout shows a block/diagonal
structure — each cell type lights up its defining marker(s) and little else.

```{r concordance-matrix}
conc <- mk_long |>
  group_by(phenotype_clean, marker) |>
  summarise(mean_z = mean(zscore, na.rm = TRUE),
            pct_pos = mean(pos,  na.rm = TRUE), .groups = "drop")

ggplot(conc, aes(marker, phenotype_clean, fill = mean_z)) +
  geom_tile(colour = "white") +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b",
                       midpoint = 0, name = "mean z") +
  labs(title = "Phenotype vs marker: mean z-score",
       subtitle = "Expect each phenotype to peak on its defining marker(s); off-diagonal signal = mislabelled cells.",
       x = "marker", y = "phenotype (phenotype_clean)", caption = CAP) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())
```

```{r concordance-expected}
# Encode the expected defining marker(s) per phenotype and check it is the top
# (or near-top) signal. Flags phenotypes whose defining marker is not their max.
expected <- tibble::tribble(
  ~phenotype_clean,           ~marker,
  "PANCK+Tumor",              "PANCK",
  "VIM+Tumor",                "VIMENTIN",
  "T helper",                 "CD4",
  "T cytotoxic",              "CD8",
  "Activated T cytotoxic",    "CD8",
  "CD4+ Treg",                "FOXP3",
  "CD8+ T reg",               "FOXP3",
  "Natural Killer",           "CD56",
  "Activated Natural Killer", "CD56",
  "Immune",                   "CD45",
  "Stroma",                   "SMA"
)

check <- conc |>
  group_by(phenotype_clean) |>
  mutate(rank_z = rank(-mean_z)) |>
  inner_join(expected, by = c("phenotype_clean", "marker")) |>
  transmute(phenotype_clean, defining_marker = marker,
            mean_z = round(mean_z, 2), rank_of_defining = as.integer(rank_z),
            pass = rank_z <= 2) |>
  arrange(desc(rank_of_defining))
knitr::kable(check, caption = "Is each phenotype's defining marker among its top-2 signals?")
```

## 2. Per-channel usability

A usable marker has a **bimodal** z-score (a negative and a positive population)
and a `_sign` call that cleanly separates them. Bimodality: Hartigan's dip test
(if `diptest` is installed) plus Sarle's bimodality coefficient (BC > 0.555
suggests bimodality). Separation: are the `_sign+` z-scores disjoint from
`_sign−`, and what cutoff sits between them.

```{r channel-qc}
bimodality_coef <- function(x) {                       # Sarle's BC
  x <- x[is.finite(x)]; n <- length(x)
  if (n < 4) return(NA_real_)
  m <- mean(x); s <- sd(x); if (s == 0) return(NA_real_)
  g <- mean(((x - m) / s)^3)                           # skewness
  k <- mean(((x - m) / s)^4) - 3                        # excess kurtosis
  (g^2 + 1) / (k + 3 * (n - 1)^2 / ((n - 2) * (n - 3)))
}
dip_p <- function(x) {
  x <- x[is.finite(x)]
  if (!requireNamespace("diptest", quietly = TRUE) || length(x) < 4) return(NA_real_)
  suppressWarnings(diptest::dip.test(x)$p.value)
}

channel_qc <- ihc_marker_long(ihc_data) |>
  filter(!is.na(zscore)) |>
  group_by(marker) |>
  summarise(
    n        = n(),
    bc       = bimodality_coef(zscore),
    dip_p    = dip_p(zscore),
    neg_max  = suppressWarnings(max(zscore[!pos], na.rm = TRUE)),
    pos_min  = suppressWarnings(min(zscore[pos],  na.rm = TRUE)),
    .groups  = "drop"
  ) |>
  mutate(
    separable = is.finite(neg_max) & is.finite(pos_min) & pos_min > neg_max,
    cutoff    = if_else(separable, (neg_max + pos_min) / 2, NA_real_),
    likely_bimodal = bc > 0.555 | (!is.na(dip_p) & dip_p < 0.05)
  ) |>
  arrange(desc(bc))
knitr::kable(channel_qc, digits = 3,
             caption = "Per-channel bimodality and sign-vs-zscore separability")
```

```{r channel-qc-plot, fig.height=8}
ihc_marker_long(ihc_data) |>
  filter(!is.na(zscore), !is.na(sign)) |>
  mutate(call = if_else(pos, "sign +", "sign -")) |>
  ggplot(aes(zscore, fill = call)) +
    geom_density(alpha = 0.55, colour = NA) +
    facet_wrap(~ marker, scales = "free") +
    scale_fill_manual(values = c("sign +" = oi[1], "sign -" = oi[2]), name = NULL) +
    labs(title = "Marker z-score by sign call",
         subtitle = "Clean channels show two separated humps; failed channels overlap or are unimodal.",
         x = "z-score", y = "density", caption = CAP)
```

## 3. Morphology by phenotype

Segmentation should produce cells whose shape matches their type: tumour cells
larger, lymphocytes small and round (high solidity), stroma elongated (high
eccentricity). Includes circularity = 4·π·area / perimeter².

```{r morphology}
morph_feats <- intersect(c("area", "perimeter", "eccentricity", "solidity"),
                         names(ihc_data))

morph <- ihc_data |>
  filter(!is.na(phenotype_clean), phenotype_clean != "Unknown") |>
  mutate(circularity = if (all(c("area", "perimeter") %in% names(ihc_data)))
           4 * pi * area / perimeter^2 else NA_real_) |>
  select(phenotype_clean, any_of(c(morph_feats, "circularity"))) |>
  pivot_longer(-phenotype_clean, names_to = "feature", values_to = "value") |>
  filter(is.finite(value))

ggplot(morph, aes(reorder(phenotype_clean, value, FUN = median), value,
                  fill = phenotype_clean)) +
  geom_boxplot(outlier.shape = NA, width = .6, colour = "grey30") +
  facet_wrap(~ feature, scales = "free") +
  coord_flip() +
  scale_fill_viridis_d(option = "D", guide = "none") +
  labs(title = "Cell morphology by phenotype",
       subtitle = "Shapes should track cell type: tumour largest, lymphocytes small/round, stroma elongated.",
       x = NULL, y = NULL, caption = CAP)
```

## Notes

- All three checks use only `ihc_data` — no external reference, so they hold even
  for slides with no pathologist annotation or matched RNA.
- Concordance (#1) reads the gating; channel QC (#2) reads the individual stains;
  morphology (#3) reads the segmentation. A failure in #1 is often explained by a
  culprit surfaced in #2 (a bad channel) or by spillover (see
  [exclusivity.html](exclusivity.html)).
- `separable = FALSE` in the channel table means `_sign` is not a pure threshold on
  `_zscore` (likely thresholded on `_raw` instead) — inspect that channel's density
  panel before trusting its positivity calls.
