---
title: "benchmarks"
author: "sceriff0"
date: "2026-07-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: tidyverse + scales + here (in the lockfile). Figure logic is vendored from
# the mirage benchmarking branch (code/benchmark_plots.R).
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
})
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

bench_dir  <- here("data", "benchmark")
have_bench <- file.exists(file.path(bench_dir, "measurements.csv"))
```

## Purpose

Mirage image-processing **pipeline benchmark** — resource scaling, registration
cost/accuracy, and segmentation-method comparison from the SLURM sweep. Every
figure is built by sourcing `code/benchmark_plots.R` (vendored from
`benchmarks/analysis/plots.R`) and rendered **inline from the sweep CSVs** — no
PNGs on disk. This report is a formula index for those figures: for each one it
states the raw columns consumed and the exact reduction to the plotted value.

## Starting data and notation

The core input is **`data/benchmark/measurements.csv`** — one row per
$(\text{run}\times\text{process})$. Write $r$ for a run and $s$ for a process
(`proc`, the `process` string with everything up to the last `:` stripped:
$\texttt{proc}=\texttt{str\_replace(process, ".*:", "")}$). Per row:

$$
R_{r,s}=\texttt{peak\_rss\_gb}\ \text{(GiB)},\quad
T_{r,s}=\texttt{realtime\_s}\ \text{(s)},\quad
x_r=\texttt{input\_gb}\ \text{(GiB)},
$$

plus every swept parameter and the label `varied_axis` naming which knob that run
varied off baseline. Size-scaling figures keep only

$$
\texttt{size\_axes}=\{\texttt{baseline},\texttt{scaling\_grid},\texttt{registration\_grid},
\texttt{distributed\_grid},\texttt{target\_px},\texttt{n\_channels}\}.
$$

Registration-stage RAM is aggregated over the registration **leaf** processes

$$
\texttt{reg\_leaves}=\{\texttt{REGISTER},\texttt{REG\_PREP},\texttt{REG\_TILE},
\texttt{REG\_NONRIGID},\texttt{REG\_MICRO\_PREP},\texttt{REG\_FINALIZE},\dots\},
$$

and boolean sweep flags are read with $\texttt{truthy}(x)=\big[\operatorname{lower}(x)\in\{\texttt{true},1,\texttt{yes}\}\big]$.

**Optional CSVs** (a figure needing an absent file is skipped): `resource_stats.csv`,
`quality.csv`, `run_cost.csv`, `segmentation_agreement.csv`,
`classic_vs_distributed_registration.csv`, `registration_drift.csv`. Accuracy
columns (`reg_tre_median_px`, `max_abs_delta`, `pct_pixels_diff`,
`rss_saving_gb`, `peak_rss_gb_mean/std`, `instance_f1`, `cell_count_ratio`) are
**precomputed upstream** and only read here — see *Notes*.

```{r run-benchmark}
bench_figs <- list()
if (have_bench) {
  # sourcing builds `bench_figs` (a named list of ggplots) from the sweep CSVs
  source(here("code", "benchmark_plots.R"))
  cat(length(bench_figs), "figure(s) built from", basename(bench_dir), "data.\n")
} else {
  cat("No `data/benchmark/measurements.csv` found — add the sweep CSVs and re-knit.")
}
```

## Metrics

Every reduction below lives in `code/benchmark_plots.R`; the cited function/object
name lets a reader trace raw column $\to$ plotted value.

**Power-law scaling** (`01_memory_scaling_per_process`, `02_time_scaling_per_process`;
`powerlaw` + `powerlaw_plot`). Over rows with `varied_axis` $\in\texttt{size\_axes}$,
$x_r>0$ and $y>0$, per process $s$ an OLS line is fit in log-log space:

$$
\log_{10} y \;=\; a_s + \beta_s\,\log_{10} x_r
\qquad(\texttt{lm(log10(y) \textasciitilde{} log10(input\_gb))}),
$$

reporting the **scaling exponent** $\beta_s$ (slope) and $R^2_s$
(`summary(f)$r.squared`); the drawn line is $y=10^{\,a_s+\beta_s\log_{10}x}$ across
$[\min x,\max x]$. $y=R_{r,s}$ for fig 01, $y=T_{r,s}$ for fig 02.
Reading: $\beta=1$ linear, $\beta>1$ super-linear, $\beta<1$ sub-linear.

**Classic vs distributed registration RAM** (`03_classic_vs_distributed_ram`,
`03b_distributed_ram_saving`; from `classic_vs_distributed_registration.csv`).
Columns `reg_peak_rss_gb_classic` and `reg_peak_rss_gb_distributed` are plotted
directly vs `target_px`, faceted by `n_channels`. The saving is the precomputed
column $\texttt{rss\_saving\_gb}=\texttt{classic}-\texttt{distributed}$ (GiB).

**N-image registration RAM** (`04_nimage_registration_ram`). `REGISTER` rows on
`registration_grid`/`baseline`/`scaling_grid`, replicate mean per configuration:

$$
\overline{R}(\texttt{target\_px},\texttt{n\_channels},\texttt{n\_register\_images})
=\operatorname{mean}(R_{r,\texttt{REGISTER}}).
$$

**OFAT knob effects** (`05_ofat_knob_effects`). For each knob in `knob_targets`
(an `(axis, proc, metric)` triple), keep rows with `varied_axis` $\in\{\text{axis},
\texttt{baseline}\}$ and $\texttt{proc}=\text{proc}$, then average the metric
(`realtime_s`) per distinct knob value: $\overline{T}(\text{value})=\operatorname{mean}(T)$.

**Replicate variance** (`06_replicate_variance`; from `resource_stats.csv`, rows
with `n_reps` $>1$). Bars are the precomputed `peak_rss_gb_mean`; error bars span
$\texttt{peak\_rss\_gb\_mean}\pm\texttt{peak\_rss\_gb\_std}$.

**Stage memory heatmap** (`07_stage_memory_heatmap`). Rows on `size_axes` with
`n_channels` $=2$ and `n_register_images` $=2$; tile value is the replicate mean
$\operatorname{mean}(R_{r,s})$ per $(\texttt{proc},\texttt{target\_px})$
(log colour scale).

**Channel-count effect** (`08_channel_effect`). Points $(x_r,R_{r,s})$ for
$s\in\{$REGISTER, PREPROCESS, SEGMENT, QUANTIFY$\}$ on `size_axes`, $x_r>0$; the
overlaid trend is `geom_smooth(method="lm")` fit on the $\log_{10}$–$\log_{10}$
axes, coloured by `n_channels`. No scalar emitted.

**Segmentation methods** (`09_segmentation_methods`, `09b_stardist_tile_grid`).
On `segmentation_grid*` / `proc = SEGMENT`: fig 09 is the raw $T_{r,\texttt{SEGMENT}}$
distribution (box = IQR) per `seg_method`. Fig 09b, `stardist` only, tiles
$\operatorname{mean}(R)$ per $(\texttt{seg\_n\_tiles\_x},\texttt{seg\_n\_tiles\_y})$.

**Registration knobs, both paths** (`10_registration_params_both_paths`). On
`registration_param_grid` / `proc` $\in\texttt{reg\_leaves}$, the per-run
registration peak is the **max over leaves**

$$
\texttt{reg\_peak\_gb}_r=\max_{s\in\texttt{reg\_leaves}} R_{r,s},
\qquad \texttt{path}_r=\begin{cases}\texttt{distributed}&\texttt{truthy(reg\_distributed\_tiling)}\\\texttt{classic}&\text{else,}\end{cases}
$$

then averaged: $\operatorname{mean}(\texttt{reg\_peak\_gb})$ per
$(\texttt{memory\_mode},\texttt{skip\_micro\_registration},\texttt{path})$.

**Registration accuracy vs cost** (`11_accuracy_vs_cost`). The accuracy metric,
`reg_tre_median_px` from `quality.csv` (feature-based **target registration
error**, median px per run — precomputed, see *Notes*), plotted vs
`cpu_hours` from `run_cost.csv`, joined on `run_id`; lower-left is better.

**Segmentation cell counts** (`12_segmentation_cell_counts`). Raw `n_cells`
(max mask label) from `quality.csv`, distribution per `seg_method`.

**Segmentation cross-method agreement** (`12b_segmentation_agreement`; from
`segmentation_agreement.csv`). Bar = `instance_f1` (IoU-matched per-cell F1 between
`method_a` and `method_b`, $1=$ agree on every cell); text label = `cell_count_ratio`.
Both precomputed.

**End-to-end cost vs size** (`13_end_to_end_cost`; from `run_cost.csv`, `size_axes`).
Per `target_px`:

$$
\overline{\texttt{cpu\_hours}}=\operatorname{mean}(\texttt{cpu\_hours}),\qquad
\overline{\texttt{wall\_clock\_h}}=\frac{\operatorname{mean}(\texttt{wall\_clock\_s})}{3600}.
$$

**Bottleneck by size** (`14_bottleneck_by_size`; from `run_cost.csv`). Count runs
per $(\texttt{target\_px},\texttt{bottleneck\_stage})$, drawn `position="fill"`, so
each bar is the **share** $n_{\text{stage}}/\sum_{\text{stage}} n$ at that size.

**Tiled path granularity** (`15_tiled_path_granularity`). On
`distributed_tiling_grid` / `reg_leaves`, per run
$\texttt{reg\_peak\_gb}_r=\max_s R_{r,s}$ and
$\texttt{reg\_time\_s}_r=\sum_s T_{r,s}$; then $\operatorname{mean}(\texttt{reg\_peak\_gb})$
per $(\texttt{reg\_dist\_tile\_wh},\texttt{reg\_dist\_tile\_buffer})$.

**Tiled path pixel drift** (`16_tiled_drift_from_classic`; from
`registration_drift.csv`, `path == "tiled"`). Mean of the precomputed
`max_abs_delta` (max $|\Delta\text{intensity}|$ vs the classic slide; $0=$ identical)
and `pct_pixels_diff` per $(\texttt{tile\_wh},\texttt{tile\_buffer})$.

**Registration error by path** (`17_registration_error_by_path`; from `quality.csv`).
`reg_tre_median_px` distribution per `path`, where

$$
\texttt{path}=\begin{cases}
\texttt{classic}   & \lnot\texttt{truthy(reg\_distributed\_tiling)}\\
\texttt{tiled}     & \texttt{truthy(reg\_dist\_force\_tiling)}\\
\texttt{separated} & \text{otherwise.}
\end{cases}
$$

## Figures

```{r figures, results="asis", fig.width=9.5, fig.height=6}
if (length(bench_figs)) {
  for (nm in names(bench_figs)) {
    title <- gsub("_", " ", sub("^[0-9]+[a-z]?_", "", nm))   # "01_memory_scaling" -> "memory scaling"
    cat("\n\n### ", title, "\n\n", sep = "")
    print(bench_figs[[nm]])
    cat("\n\n")
  }
} else if (have_bench) {
  cat("Benchmark data present but no figures were produced — check `measurements.csv` columns.")
}
```

## Notes

- **Accuracy metrics are read, not recomputed.** `reg_tre_median_px` (TRE),
  `max_abs_delta`/`pct_pixels_diff` (drift), `instance_f1`/`cell_count_ratio`
  (segmentation agreement), `rss_saving_gb`, and `peak_rss_gb_mean/std` all arrive
  precomputed in their CSVs. The only reductions `benchmark_plots.R` itself
  performs are the log-log OLS ($\beta$, $R^2$), the per-run
  $\max$/$\sum$ over `reg_leaves`, and replicate means/shares. The subtitles call
  TRE a "feature TRE proxy"; treat it as a proxy, not a ground-truth landmark error.
- **Reductions are means over replicates** unless noted; SLURM isolates per-process
  resources, so $R_{r,s}$ is one process's peak, and registration-stage RAM is the
  max across its leaf processes within a run.
- Any figure whose optional CSV (or required column) is missing is silently
  skipped; the rendered set depends on which sweep outputs were dropped into
  `data/benchmark/`.
