---
title: "benchmarks"
author: "sceriff0"
date: "2026-07-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: tidyverse + here (in the lockfile). Optional: diptest for Hartigan's dip
# test (a bimodality-coefficient fallback runs if it is absent).
#   renv::install("diptest"); renv::snapshot()
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
})

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6)

invisible(capture.output(suppressMessages(source(here("code", "test.R")))))
source(here("code", "validation_helpers.R"))
```

## Purpose

**Internal QC of the IHC readout** — validation that needs no external reference,
only `ihc_data`. Traces a fault backwards through the pipeline:

1. **Phenotype ↔ marker concordance** — do cells labelled a type express the
   marker that defines it? (validates the phenotyping/gating)
2. **Lineage exclusivity** — a panel of literature-backed mutually-exclusive
   marker pairs (tumour/immune, T/NK, lymphoid/myeloid, …); excess co-positivity
   signals segmentation or optical spillover between neighbours.
3. **Per-channel usability** — is each marker's `_sign` a clean threshold on a
   bimodal signal? (validates the individual stains)
4. **Morphology by phenotype** — do cell shapes match their assigned type?
   (validates that segmentation captures real cells)

```{r marker-long}
mk_long <- ihc_marker_long(ihc_data) |>
  filter(!is.na(phenotype_clean), phenotype_clean != "Unknown")
cat("cells:", dplyr::n_distinct(mk_long$cell_id),
    "| markers:", dplyr::n_distinct(mk_long$marker),
    "| phenotypes:", dplyr::n_distinct(mk_long$phenotype_clean), "\n")
```

## 1. Phenotype ↔ marker concordance

Mean marker z-score per phenotype. A trustworthy readout shows a block/diagonal
structure — each cell type lights up its defining marker(s) and little else.

```{r concordance-matrix}
conc <- mk_long |>
  group_by(phenotype_clean, marker) |>
  summarise(mean_z = mean(zscore, na.rm = TRUE),
            pct_pos = mean(pos,  na.rm = TRUE), .groups = "drop")

ggplot(conc, aes(marker, phenotype_clean, fill = mean_z)) +
  geom_tile(colour = "grey90") +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b",
                       midpoint = 0, name = "mean z") +
  labs(title = "Phenotype vs marker: mean z-score",
       subtitle = "expect each phenotype to peak on its defining marker(s)",
       x = "marker", y = "phenotype (phenotype_clean)") +
  theme_minimal(base_size = 11) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())
```

```{r concordance-expected}
# Encode the expected defining marker(s) per phenotype and check it is the top
# (or near-top) signal. Flags phenotypes whose defining marker is not their max.
expected <- tibble::tribble(
  ~phenotype_clean,           ~marker,
  "PANCK+Tumor",              "PANCK",
  "VIM+Tumor",                "VIMENTIN",
  "T helper",                 "CD4",
  "T cytotoxic",              "CD8",
  "Activated T cytotoxic",    "CD8",
  "CD4+ Treg",                "FOXP3",
  "CD8+ T reg",               "FOXP3",
  "Natural Killer",           "CD56",
  "Activated Natural Killer", "CD56",
  "Immune",                   "CD45",
  "Stroma",                   "SMA"
)

check <- conc |>
  group_by(phenotype_clean) |>
  mutate(rank_z = rank(-mean_z)) |>
  inner_join(expected, by = c("phenotype_clean", "marker")) |>
  transmute(phenotype_clean, defining_marker = marker,
            mean_z = round(mean_z, 2), rank_of_defining = as.integer(rank_z),
            pass = rank_z <= 2) |>
  arrange(rank_of_defining |> desc())
knitr::kable(check, caption = "Is each phenotype's defining marker among its top-2 signals?")
```

## 3. Lineage exclusivity (segmentation / staining spillover)

Cell lineages that biology keeps separate should not co-stain in the same cell.
Excess co-positivity is a direct signal of **optical spillover between touching
cells** or **mis-segmentation** (one mask capturing two cells). This generalises
the single PANCK⁺/CD45⁺ check into a panel of literature-backed mutually-exclusive
marker pairs, each with the rare biological exception noted so the QC flags
*excess* co-positivity, not the known baseline.

### 3a. Phenotype-level leakage

CD45⁺ cells among tumour-typed cells, and PANCK⁺ cells among immune-typed cells —
both should be ~0 if the phenotyping respects the tumour/immune divide.

```{r exclusivity-phenotype}
immune_ph <- c("Immune", "T helper", "T cytotoxic", "Activated T cytotoxic",
               "CD4+ Treg", "CD8+ T reg", "Natural Killer", "Activated Natural Killer")

excl <- ihc_data |>
  mutate(patient_id  = norm_id(patient_id),
         is_tumor_ph = str_detect(replace_na(phenotype_clean, ""), "Tumor"),
         is_immune_ph = phenotype_clean %in% immune_ph,
         panck_pos = is_pos(PANCK_sign),
         cd45_pos  = is_pos(CD45_sign))

spill <- excl |>
  group_by(patient_id) |>
  summarise(
    n_cells            = n(),
    cd45pos_in_tumorph = if (any(is_tumor_ph))  mean(cd45_pos[is_tumor_ph],  na.rm = TRUE) else NA_real_,
    panckpos_in_immune = if (any(is_immune_ph)) mean(panck_pos[is_immune_ph], na.rm = TRUE) else NA_real_,
    .groups = "drop"
  )
knitr::kable(spill, digits = 3,
             caption = "Per-patient phenotype-level leakage (want ~0)")
```

### 3b. Marker-pair exclusivity panel

Each pair is two markers biology assigns to distinct lineages. `p_b_given_a` /
`p_a_given_b` are the conditional co-positivity rates — the fraction of one
marker's positive cells that also call the other. These should sit near the
"expected co-positivity" column (the rare true double-positive population); values
well above it flag spillover for that pair.

```{r exclusivity-pairs}
# Per-cell marker positivity matrix (one logical column per marker).
sign_pos <- ihc_data |>
  transmute(patient_id = norm_id(patient_id),
            across(all_of(paste0(ihc_markers, "_sign")), is_pos, .names = "{.col}")) |>
  rename_with(~ sub("_sign$", "", .x), ends_with("_sign"))

# Literature-backed mutually-exclusive lineage marker pairs (+ known exception).
exclusivity_pairs <- tibble::tribble(
  ~a,      ~b,      ~pair_label,                              ~expected_copos, ~reference,
  "PANCK", "CD45",  "epithelial/tumour vs leukocyte",         "~0 (rare hybrid CTCs)",     "Nandedkar 1998",
  "CD4",   "CD8",   "helper vs cytotoxic T",                  "~3% peripheral DP-T",       "Overgaard 2023 (Biomedicines)",
  "CD3",   "CD56",  "T cell vs NK cell",                      "rare CD3+CD56+ NKT",        "std NK/NKT gating",
  "CD3",   "CD14",  "lymphoid vs monocyte/macrophage",        "~0 distinct lineages",      "flow-panel convention",
  "CD3",   "CD163", "lymphoid vs M2 macrophage",              "~0 (lymphocytes CD163-)",   "Roberts 2004",
  "CD45",  "SMA",   "leukocyte vs smooth-muscle/myofibroblast","~0 distinct lineages",     "stromal-marker convention"
) |>
  # keep only pairs whose markers are both present
  filter(a %in% names(sign_pos), b %in% names(sign_pos))

pair_stats <- exclusivity_pairs |>
  rowwise() |>
  mutate(
    n_a          = sum(sign_pos[[a]], na.rm = TRUE),
    n_b          = sum(sign_pos[[b]], na.rm = TRUE),
    n_both       = sum(sign_pos[[a]] & sign_pos[[b]], na.rm = TRUE),
    pct_dp_all   = n_both / nrow(sign_pos),
    p_b_given_a  = ifelse(n_a > 0, n_both / n_a, NA_real_),
    p_a_given_b  = ifelse(n_b > 0, n_both / n_b, NA_real_)
  ) |>
  ungroup() |>
  select(pair = pair_label, a, b, expected_copos, reference,
         n_both, pct_dp_all, p_b_given_a, p_a_given_b)

knitr::kable(pair_stats, digits = 4,
             caption = "Mutually-exclusive marker pairs: observed vs expected co-positivity")
```

```{r exclusivity-pairs-plot}
per_pat <- exclusivity_pairs |>
  select(a, b, pair_label) |>
  purrr::pmap_dfr(function(a, b, pair_label) {
    sign_pos |>
      group_by(patient_id) |>
      summarise(dp = mean(.data[[a]] & .data[[b]], na.rm = TRUE), .groups = "drop") |>
      mutate(pair = pair_label)
  })

if (nrow(per_pat) > 0) {
  ggplot(per_pat, aes(patient_id, pair, fill = dp)) +
    geom_tile(colour = "grey90") +
    scale_fill_viridis_c(option = "magma", direction = -1, name = "double-pos\nfraction") +
    labs(title = "Marker-pair co-positivity per patient",
         subtitle = "bright = high double-positive rate = likely spillover",
         x = "patient", y = NULL) +
    theme_minimal(base_size = 10) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1), panel.grid = element_blank())
}
```

**Pairs deliberately excluded** (this panel's phenotypes acknowledge their
overlap, so co-positivity is expected, not an artifact): PANCK/VIMENTIN (the
`VIM+Tumor` EMT phenotype), and CD8/FOXP3 (the `CD8+ T reg` phenotype).

**References for the exclusivity claims:**

- PANCK/CD45 — Nandedkar MA *et al.* (1998), CD45 immunoreactivity as a diagnostic
  pitfall in carcinoma: keratin and CD45 are normally exclusive.
  [PubMed 9872653](https://pubmed.ncbi.nlm.nih.gov/9872653/)
- CD4/CD8 — "Legend or Truth: Mature CD4⁺CD8⁺ Double-Positive T Cells in the
  Periphery," *Biomedicines* 2023: DP-T are a real but ~3% minority.
  [MDPI](https://www.mdpi.com/2227-9059/11/10/2702)
- CD3/CD56 — CD3⁻CD56⁺ = NK, CD3⁺CD56⁺ = rare NKT; standard immunophenotyping.
- CD3/CD163 — Roberts ES *et al.* (2004): lymphocytes are CD163-negative.
  [PMC2258269](https://pmc.ncbi.nlm.nih.gov/articles/PMC2258269/)

## 4. Per-channel usability

A usable marker has a **bimodal** z-score (a negative and a positive population)
and a `_sign` call that cleanly separates them. Bimodality: Hartigan's dip test
(if `diptest` is installed) plus Sarle's bimodality coefficient (BC > 0.555
suggests bimodality). Separation: are the `_sign+` z-scores disjoint from
`_sign−`, and what cutoff sits between them.

```{r channel-qc}
bimodality_coef <- function(x) {                       # Sarle's BC
  x <- x[is.finite(x)]; n <- length(x)
  if (n < 4) return(NA_real_)
  m <- mean(x); s <- sd(x); if (s == 0) return(NA_real_)
  g <- mean(((x - m) / s)^3)                           # skewness
  k <- mean(((x - m) / s)^4) - 3                        # excess kurtosis
  (g^2 + 1) / (k + 3 * (n - 1)^2 / ((n - 2) * (n - 3)))
}
dip_p <- function(x) {
  x <- x[is.finite(x)]
  if (!requireNamespace("diptest", quietly = TRUE) || length(x) < 4) return(NA_real_)
  suppressWarnings(diptest::dip.test(x)$p.value)
}

channel_qc <- ihc_marker_long(ihc_data) |>
  filter(!is.na(zscore)) |>
  group_by(marker) |>
  summarise(
    n        = n(),
    bc       = bimodality_coef(zscore),
    dip_p    = dip_p(zscore),
    neg_max  = suppressWarnings(max(zscore[!pos], na.rm = TRUE)),
    pos_min  = suppressWarnings(min(zscore[pos],  na.rm = TRUE)),
    .groups  = "drop"
  ) |>
  mutate(
    separable = is.finite(neg_max) & is.finite(pos_min) & pos_min > neg_max,
    cutoff    = if_else(separable, (neg_max + pos_min) / 2, NA_real_),
    likely_bimodal = bc > 0.555 | (!is.na(dip_p) & dip_p < 0.05)
  ) |>
  arrange(desc(bc))
knitr::kable(channel_qc, digits = 3,
             caption = "Per-channel bimodality and sign-vs-zscore separability")
```

```{r channel-qc-plot, fig.height=8}
ihc_marker_long(ihc_data) |>
  filter(!is.na(zscore), !is.na(sign)) |>
  mutate(call = if_else(pos, "sign +", "sign -")) |>
  ggplot(aes(zscore, fill = call)) +
    geom_density(alpha = 0.5) +
    facet_wrap(~ marker, scales = "free") +
    labs(title = "Marker z-score by sign call",
         subtitle = "clean channels: two separated humps; failed channels: overlapping/unimodal",
         x = "z-score", y = "density", fill = NULL) +
    theme_bw(base_size = 9)
```

## 5. Morphology by phenotype

Segmentation should produce cells whose shape matches their type: tumour cells
larger, lymphocytes small and round (high solidity), stroma elongated (high
eccentricity). Includes circularity = 4·π·area / perimeter².

```{r morphology}
morph_feats <- intersect(c("area", "perimeter", "eccentricity", "solidity"),
                         names(ihc_data))

morph <- ihc_data |>
  filter(!is.na(phenotype_clean), phenotype_clean != "Unknown") |>
  mutate(circularity = if (all(c("area", "perimeter") %in% names(ihc_data)))
           4 * pi * area / perimeter^2 else NA_real_) |>
  select(phenotype_clean, any_of(c(morph_feats, "circularity"))) |>
  pivot_longer(-phenotype_clean, names_to = "feature", values_to = "value") |>
  filter(is.finite(value))

ggplot(morph, aes(reorder(phenotype_clean, value, FUN = median), value,
                  colour = phenotype_clean)) +
  geom_boxplot(outlier.shape = NA) +
  facet_wrap(~ feature, scales = "free") +
  coord_flip() +
  labs(title = "Cell morphology by phenotype", x = NULL, y = NULL) +
  theme_classic(base_size = 10) +
  theme(legend.position = "none")
```

## Notes

- All four checks use only `ihc_data` — no external reference, so they hold even
  for slides with no pathologist annotation or matched RNA.
- Concordance (#1) reads the gating; spillover (#3) reads the segmentation;
  channel QC (#4) reads the individual stains. A failure in #1 is often explained
  by a culprit surfaced in #3 or #4.
- `separable = FALSE` in the channel table means `_sign` is not a pure threshold on
  `_zscore` (likely thresholded on `_raw` instead) — inspect that channel's density
  panel before trusting its positivity calls.
