---
title: "benchmarks"
author: "sceriff0"
date: "2026-07-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: tidyverse + scales + here (in the lockfile). Figure logic is vendored from
# the mirage benchmarking branch (code/benchmark_plots.R).
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
})
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE)

bench_dir  <- here("data", "benchmark")
have_bench <- file.exists(file.path(bench_dir, "measurements.csv"))
```

## Purpose

Mirage image-processing **pipeline benchmark** — resource scaling, registration
cost/accuracy, and segmentation-method comparison from the SLURM sweep. The figure
logic is vendored verbatim from the mirage benchmarking branch
(`benchmarks/analysis/plots.R` → `code/benchmark_plots.R`); this report just runs
it against this project's sweep data and embeds the figures.

Internal IHC QC lives elsewhere: [ihc_qc.html](ihc_qc.html) (phenotyping,
channels, morphology) and [exclusivity.html](exclusivity.html) (lineage spillover).

## Benchmark data

Drop the sweep CSVs into **`data/benchmark/`** (under the gitignored `data/`, so
they are not committed):

- `measurements.csv` — **required**; one row per (run × process) with
  `peak_rss_gb`, `realtime_s`, `input_gb`, and every swept parameter.
- optional: `resource_stats.csv`, `quality.csv`, `run_cost.csv`,
  `segmentation_agreement.csv`, `classic_vs_distributed_registration.csv`,
  `registration_drift.csv` — figures that need an absent file are skipped.

```{r run-benchmark}
if (have_bench) {
  source(here("code", "benchmark_plots.R"))   # writes PNGs to data/benchmark/figures_R/
} else {
  cat("No `data/benchmark/measurements.csv` found — add the sweep CSVs and re-knit.")
}
```

## Figures

```{r embed-figures, results="asis"}
figs <- if (have_bench) {
  sort(list.files(file.path(bench_dir, "figures_R"), pattern = "\\.png$", full.names = TRUE))
} else {
  character(0)
}

if (length(figs)) {
  for (f in figs) {
    cat("\n\n### ", tools::file_path_sans_ext(basename(f)), "\n\n", sep = "")
    cat(sprintf("![](%s)\n\n", f))
  }
} else if (have_bench) {
  cat("Benchmark data present but no figures were produced — check `measurements.csv` columns.")
}
```
