---
title: "benchmarks"
author: "sceriff0"
date: "2026-07-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: tidyverse + scales + here (in the lockfile). Figure logic is vendored from
# the mirage benchmarking branch (code/benchmark_plots.R).
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
})
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE,
                      dev = c("png", "pdf"))   # PNG for the site, PDF exported to output/
source(here("code", "pdf_export.R"))          # export_pdf_figures(): plots -> output/figures/

bench_dir  <- here("data", "benchmark")
have_bench <- file.exists(file.path(bench_dir, "measurements.csv"))
```

## Starting data and notation

Mirage image-processing **pipeline benchmark** — resource scaling (memory, runtime,
**I/O**), registration cost/accuracy, and segmentation-method comparison from the
SLURM sweep. Every figure is built by sourcing `code/benchmark_plots.R` (vendored
from mirage `benchmarks/analysis/plots.R`) and rendered **inline from the sweep
CSVs** — no PNGs on disk.

**Where the numbers come from.** One real source image is rescaled across a size
axis and a channel axis; the Mirage pipeline is run once per sweep cell (and per
swept parameter) under Nextflow tracing. Each run writes a `trace.txt` (per-process
peak RSS, runtime, CPU, and I/O byte counts) and a `size_logs/input_sizes.csv`
(per-process input bytes). `code/benchmark_plots.R` joins those into
**`data/benchmark/measurements.csv`** — **one row per run-and-process** — and the
figures reduce that table. `proc` is the `process` string with everything up to the
last `:` stripped. Per row we read:

```
peak_rss_gb = peak resident memory of that process    (GiB, trace peak_rss)
realtime_s  = wall time of that process               (s,   trace realtime)
read_gb     = bytes read by that process              (GiB, trace rchar)   [I/O volume]
write_gb    = bytes written by that process           (GiB, trace wchar)   [I/O volume]
input_gb    = input size of that run                  (GiB, size_logs)
```

plus every swept parameter and the label `varied_axis` naming which knob that run
varied off baseline. Size-scaling figures keep only the runs whose `varied_axis`
is a size axis, `size_axes = {baseline, scaling_grid, registration_grid,
distributed_grid, target_px, n_channels}`; registration-stage RAM is aggregated over
the registration **leaf** processes (`REGISTER, REG_PREP, REG_TILE, REG_NONRIGID,
REG_MICRO_PREP, REG_FINALIZE, …`); boolean sweep flags are read as "truthy" when the
lower-cased value is one of `true`, `1`, `yes`.

!!! warning "Two caveats that change how these figures read"
    **The sweep images are synthetic.** Extra channels and the moving registration
    panels are *synthesized* from channel 0 (duplicate + intensity jitter + a fixed
    pixel shift), so figures that touch channels or registration *accuracy* on sweep
    cells reflect a known synthetic offset, **not** biological registration
    difficulty. **`reg_tre_median_px` is a feature-distance proxy, not landmark
    TRE** — treat the accuracy figures (11, 17) as relative, cost-vs-error views,
    not absolute accuracy. (Absolute landmark accuracy for the classic VALIS path is
    established separately on the ANHIR/ACROBAT challenges.)

**Axes are linear throughout** (no log-log): the power-law scaling figures fit the
exponent in log space but draw the fitted curve on linear axes, with the exponent
`β` and `R²` shown in each facet strip.

**Optional CSVs** (a figure needing an absent file is skipped): `resource_stats.csv`,
`quality.csv`, `run_cost.csv`, `segmentation_agreement.csv`,
`classic_vs_distributed_registration.csv`, `registration_drift.csv`. Accuracy
columns (`reg_tre_median_px`, `max_abs_delta`, `pct_pixels_diff`, `rss_saving_gb`,
`instance_f1`, `cell_count_ratio`, `peak_rss_gb_mean`/`_std`) are **precomputed
upstream** and only read here.

```{r run-benchmark}
bench_figs <- list()
if (have_bench) {
  # sourcing builds `bench_figs` (a named list of ggplots) from the sweep CSVs
  source(here("code", "benchmark_plots.R"))
  cat(length(bench_figs), "figure(s) built from", basename(bench_dir), "data.\n")
} else {
  cat("No `data/benchmark/measurements.csv` found — add the sweep CSVs and re-knit.")
}
```

```{r fig-notes}
# Derivation note per figure: what is shown, which raw columns, the exact reduction,
# and any interpretation caveat. The figures chunk prints each note IMMEDIATELY above
# its figure, so in the knitted page every plot carries its provenance in place.
# Keyed by the names in `bench_figs`; a figure with no entry falls back to no note.
fig_notes <- c(
  "01_memory_scaling_per_process" = paste(
    "Peak resident memory `peak_rss_gb` (trace `peak_rss`) vs run input size `input_gb`,",
    "one panel per process, over size-varying runs only (`varied_axis` in `size_axes`).",
    "Per process an OLS power law `log10(peak_rss) ~ log10(input_gb)` is fit; the strip",
    "shows the exponent β (1 = linear, >1 super-linear) and R². Linear axes, curve = the fit."),
  "02_time_scaling_per_process" = paste(
    "Same construction as the memory-scaling figure but for wall time `realtime_s`",
    "(trace `realtime`): OLS power law `log10(realtime) ~ log10(input_gb)` per process,",
    "β and R² in the strip, linear axes."),
  "02b_io_volume_scaling" = paste(
    "Total I/O **volume** per process, `read_gb + write_gb` (trace `rchar`/`wchar`,",
    "cumulative bytes moved through read/write syscalls) vs `input_gb`, same power-law fit.",
    "This is transferred volume, not throughput (no per-second rate)."),
  "03_classic_vs_distributed_ram" = paste(
    "Registration-stage peak RAM, classic vs distributed path, from",
    "`classic_vs_distributed_registration.csv`: columns `reg_peak_rss_gb_classic` and",
    "`reg_peak_rss_gb_distributed` plotted vs `target_px`, faceted by `n_channels`.",
    "The classic path holds the BioFormats JVM heap; the JVM-free distributed path stays bounded."),
  "03b_distributed_ram_saving" = paste(
    "The precomputed RAM saving `rss_saving_gb = reg_peak_rss_gb_classic -",
    "reg_peak_rss_gb_distributed` (GiB) vs `target_px`, coloured by channel count."),
  "04_nimage_registration_ram" = paste(
    "`REGISTER` peak RSS vs the number of co-registered slides. Rows on",
    "`registration_grid`/`baseline`/`scaling_grid`; value = mean `peak_rss_gb` grouped by",
    "`(target_px, n_channels, n_register_images)` (1 reference + N-1 moving)."),
  "05_ofat_knob_effects" = paste(
    "One-factor-at-a-time cost knobs. For each knob (preproc iters/overlap/workers, seg GPU,",
    "quantify compartments/expansion), rows whose `varied_axis` is that knob or `baseline`;",
    "each panel = mean `realtime_s` per distinct knob value, free y-scale."),
  "06_replicate_variance" = paste(
    "Replicate spread from `resource_stats.csv` (rows with `n_reps > 1`): bars are the",
    "precomputed `peak_rss_gb_mean` per process, error bars span ± `peak_rss_gb_std` across",
    "the repeats sharing a `config_id`. This is the variance a single-shot sweep cannot give."),
  "07_stage_memory_heatmap" = paste(
    "Where the memory goes: mean `peak_rss_gb` per `(proc, target_px)` on `size_axes` at",
    "`n_channels = 2`, `n_register_images = 2`. Colour is log-scaled (fill only, not an axis)",
    "so small and large stages are both visible; darker = the memory bottleneck at that size."),
  "07b_stage_io_split" = paste(
    "I/O **volume** by stage: mean `read_gb` and `write_gb` per process (trace `rchar`/`wchar`)",
    "on `size_axes`, stacked read + write. Identifies the I/O-heavy stages and the read/write split."),
  "08_channel_effect" = paste(
    "Does 2 vs 4 channels shift memory scaling? `peak_rss_gb` vs `input_gb` for REGISTER,",
    "PREPROCESS, SEGMENT, QUANTIFY on `size_axes`, coloured by `n_channels`, with a linear trend.",
    "**Caveat:** the extra channels are synthesized (duplicate + jitter), so this is the cost of",
    "*more channel bytes*, not of independent biological channels."),
  "09_segmentation_methods" = paste(
    "Segmentation runtime by backend: `SEGMENT` `realtime_s` distribution per `seg_method`",
    "on the `segmentation_grid` runs. Box = IQR across each method's own parameter sweep;",
    "points = individual configs."),
  "09b_stardist_tile_grid" = paste(
    "StarDist only: mean `peak_rss_gb` per `(seg_n_tiles_x, seg_n_tiles_y)` tile grid —",
    "how tiling trades memory."),
  "10_registration_params_both_paths" = paste(
    "Registration knobs measured in both paths. On `registration_param_grid`, per run the",
    "registration peak = **max** `peak_rss_gb` over the `reg_leaves`; path = distributed if",
    "`reg_distributed_tiling` is truthy else classic; value = mean per",
    "`(memory_mode, skip_micro_registration, path)`."),
  "11_accuracy_vs_cost" = paste(
    "Accuracy vs cost (lower-left is better). `reg_tre_median_px` from `quality.csv` vs",
    "`cpu_hours` from `run_cost.csv`, joined on `run_id`, one point per config.",
    "**`reg_tre_median_px` is a feature-distance proxy, not landmark TRE** — read this as a",
    "relative cost/error trade-off, not absolute registration accuracy."),
  "12_segmentation_cell_counts" = paste(
    "Cells detected per method: raw `n_cells` (max mask label, from `quality.csv`) distribution",
    "per `seg_method`. Large gaps between methods mean they disagree on how many cells exist."),
  "12b_segmentation_agreement" = paste(
    "Cross-method **agreement**, from `segmentation_agreement.csv`: bar = `instance_f1`",
    "(IoU-matched per-cell F1 between two methods, 1 = agree on every cell); text = `cell_count_ratio`.",
    "**This is inter-method concordance, not accuracy vs a ground truth** — two methods wrong the",
    "same way would still agree."),
  "13_end_to_end_cost" = paste(
    "Whole-pipeline cost vs image size, from `run_cost.csv` on `size_axes`: per `target_px`,",
    "mean `cpu_hours` (total compute) and mean `wall_clock_s / 3600` (wall-clock per slide)."),
  "14_bottleneck_by_size" = paste(
    "Which stage dominates, from `run_cost.csv`: count of runs per `(target_px, bottleneck_stage)`",
    "drawn as a filled bar, so each bar is the **share** of runs whose slowest single process is",
    "that stage — the bottleneck shifts with image size."),
  "15_tiled_path_granularity" = paste(
    "Distributed tiled path only (`distributed_tiling_grid`): per run the registration peak =",
    "**max** `peak_rss_gb` over `reg_leaves`; value = mean per `(reg_dist_tile_wh,",
    "reg_dist_tile_buffer)`. How tile size and overlap trade RAM."),
  "16_tiled_drift_from_classic" = paste(
    "Tiled path pixel drift from `registration_drift.csv` (`path == \"tiled\"`): mean of the",
    "precomputed `max_abs_delta` (max absolute intensity difference vs the classic slide; 0 =",
    "identical) per `(tile_wh, tile_buffer)`. Drift, not a failure — tiled is a different algorithm."),
  "17_registration_error_by_path" = paste(
    "Registration error by path, from `quality.csv`: `reg_tre_median_px` distribution per path",
    "(classic / separated / tiled, decided from `reg_distributed_tiling` and `reg_dist_force_tiling`).",
    "Separated should match classic; tiled shows any accuracy cost of tiling.",
    "Same **feature-distance-proxy** caveat as figure 11.")
)
```

## Figures

Each figure below is preceded by its derivation note — the raw columns it consumes
and the exact reduction to the plotted value. Figures whose optional CSV was not
present in `data/benchmark/` are silently absent.

```{r figures, results="asis", fig.width=9.5, fig.height=6}
if (length(bench_figs)) {
  for (nm in names(bench_figs)) {
    title <- gsub("_", " ", sub("^[0-9]+[a-z]?_", "", nm))   # "01_memory_scaling" -> "memory scaling"
    cat("\n\n### ", title, "\n\n", sep = "")
    if (!is.na(fig_notes[nm])) cat(fig_notes[[nm]], "\n\n")   # derivation note, directly above the plot
    print(bench_figs[[nm]])
    cat("\n\n")
  }
} else if (have_bench) {
  cat("Benchmark data present but no figures were produced — check `measurements.csv` columns.")
}
```

## Notes

- **Accuracy metrics are read, not recomputed.** `reg_tre_median_px` (a
  feature-distance proxy), `max_abs_delta`/`pct_pixels_diff` (drift),
  `instance_f1`/`cell_count_ratio` (segmentation agreement), `rss_saving_gb`, and the
  replicate mean/sd all arrive precomputed in their CSVs. The reductions
  `benchmark_plots.R` performs itself are the log-space OLS (β, R²), the per-run
  max/sum over `reg_leaves`, and replicate means/shares.
- **Reductions are means over replicates** unless noted; SLURM isolates per-process
  resources, so `peak_rss_gb` is one process's peak and registration-stage RAM is the
  max across its leaf processes within a run. I/O (`read_gb`/`write_gb`) is transferred
  **volume**, not throughput.
- Any figure whose optional CSV (or required column) is missing is silently skipped;
  the rendered set depends on which sweep outputs were dropped into `data/benchmark/`.

```{r export-pdfs, include=FALSE}
# Copy every plot from this analysis (each rendered as a PDF via the
# dev = c("png", "pdf") chunk option) into output/figures/benchmarks/. Runs last so
# that all figures from the chunks above already exist on disk.
export_pdf_figures("benchmarks")
```
