---
title: "bulkRna"
author: "sceriff0"
date: "2026-07-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: tidyverse + here (already in the lockfile).
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
})

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6,
                      dev = c("png", "pdf"))   # PNG for the site, PDF exported to output/

invisible(capture.output(suppressMessages(source(here("code", "test.R")))))
source(here("code", "validation_helpers.R"))
source(here("code", "pdf_export.R"))          # export_pdf_figures(): plots -> output/figures/
```

## Starting data and notation

Every value below is derived from three inputs:

- `ihc_data` — FlowPath single cells, deduplicated to **one copy per patient**
  (whole slide, no annotation boundary). Carries per-marker `<marker>_sign` and
  `<marker>_zscore` columns and `patient_id`.
- `counts_data` — DESeq2-normalised **bulk RNA**, one row per RNA sample, one
  column per gene, plus the CRF-space sample id `Sample`.
- `clinical_data` — the clinical CRF, used only for the id crosswalk
  `ID PATIENT` (slide space) <-> `ID CRF PRESERVE` (CRF space).

Notation and mappings used throughout:

```
cells of a patient p = all of p's cells in `ihc_data`
marker-positive cell (marker <marker>) = is_pos(<marker>_sign) is TRUE
  (is_pos accepts +, pos, positive, yes, true, 1)
crf2slide = crf_to_slide_map(clinical_data): a named lookup from
            norm_id(`ID CRF PRESERVE`) (CRF space) to slide_key(`ID PATIENT`) (slide space)
slide_key(id) = the canonical slide key (strip non-digits,
                apply the verified typo fix `15879` -> `15897`)
genes of a marker = marker_gene_map[<marker>]: a comma-separated gene list
  (e.g. CD3 -> CD3D, CD3E, CD3G; PANCK -> EPCAM, KRT5, KRT8, KRT14, KRT18, KRT19);
  multi-gene markers are averaged
```

Paired associations are Spearman correlations via `paired_spearman`.

## Per-patient bulk expression of each marker's gene(s)

RNA samples are keyed in CRF space; `crf2slide` bridges them to slide space.
`counts_data` is melted long to one row per (sample, gene), with columns
`Sample`, `crf_id = norm_id(Sample)`, `gene`, and `expr`. `gene_rows` expands
`marker_gene_map` to one row per (marker, gene) over each marker's gene list
(comma-split, trimmed); `marker_genes` keeps only the (marker, gene) rows whose
gene appears in the RNA matrix:

```
available genes of a marker = its genes that are also columns of `counts_data`
```

Per marker and CRF sample, expression is the mean over that marker's available
genes:

```
rna_expr(<marker>, s) = (sum over the marker's available genes of expr(s, gene))
                        / (number of the marker's available genes)
where:
  expr(s, gene) = the DESeq2-normalised value for gene in sample s
```

The CRF id is then mapped to slide space with
`patient_id = crf2slide[crf_id(s)]`, and rows with no crosswalk match (NA) are
dropped.

```{r marker-expression}
# RNA samples are CRF-id space; bridge to IHC slide-id space via clinical
# (base-R named-vector lookup, not a join).
crf2slide <- crf_to_slide_map(clinical_data)

expr_long <- counts_data |>
  mutate(crf_id = norm_id(Sample)) |>
  pivot_longer(-c(Sample, crf_id), names_to = "gene", values_to = "expr")

# marker -> its genes, one row per (marker, gene). Explicit $-access + strsplit
# (no NSE) so it is robust to the base pipe / tidyr separate_rows() quirks.
gene_rows <- do.call(rbind, Map(function(m, g)
  data.frame(marker = m, gene = trimws(strsplit(g, ",")[[1]]), stringsAsFactors = FALSE),
  marker_gene_map$marker, marker_gene_map$genes))

missing <- setdiff(gene_rows$gene, unique(expr_long$gene))
if (length(missing)) message("genes absent from RNA matrix: ", paste(missing, collapse = ", "))

# keep only genes present in the RNA matrix
marker_genes <- gene_rows |> filter(gene %in% unique(expr_long$gene))

# Per patient, average expression across a marker's available genes, then map the
# CRF sample id to the IHC slide id via the crosswalk.
marker_expr <- expr_long |>
  inner_join(marker_genes, by = "gene", relationship = "many-to-many") |>
  group_by(marker, crf_id) |>
  summarise(rna_expr = mean(expr), .groups = "drop") |>
  mutate(patient_id = unname(crf2slide[crf_id])) |>   # crf id -> slide id
  filter(!is.na(patient_id)) |>
  select(marker, patient_id, rna_expr)
```

## Per-patient IHC positive-cell fraction

`ihc_marker_fraction(ihc_data)` groups cells by `slide_key(patient_id)` and, for
each marker, returns the whole-slide positive fraction (mean of `is_pos` over
non-NA sign values):

```
<marker>_posfrac(p) = (cells of p positive for <marker>) / (cells of p with a non-NA <marker>_sign)
where:
  positive = is_pos(<marker>_sign): one of +, pos, positive, yes, true, 1
```

These `_posfrac` columns are reshaped long to `patient_id`, `marker`, and
`ihc_posfrac`.

```{r ihc-fraction}
ihc_frac <- ihc_marker_fraction(ihc_data) |>
  select(patient_id, ends_with("_posfrac")) |>
  pivot_longer(-patient_id, names_to = "marker", values_to = "ihc_posfrac") |>
  mutate(marker = sub("_posfrac$", "", marker))

cat("IHC patients:", dplyr::n_distinct(ihc_frac$patient_id),
    "| RNA patients:", dplyr::n_distinct(marker_expr$patient_id), "\n")
```

## Marker-by-marker concordance

Inner-join `marker_expr` and `ihc_frac` on `marker` and `patient_id`, keeping
finite pairs. Per marker, over its shared patients, the statistic is the Spearman
correlation between `ihc_posfrac` and `rna_expr` across patients (`paired_spearman`,
via `cor.test(method = "spearman")`), reported with its p-value and the pair count
`n`; it is `NA` when `n < 3` or either vector is constant. Rows are sorted by
descending correlation.

```{r concordance}
paired <- marker_expr |>
  inner_join(ihc_frac, by = c("marker", "patient_id")) |>
  filter(is.finite(rna_expr), is.finite(ihc_posfrac))

cat("shared patients (RNA + IHC):",
    dplyr::n_distinct(paired$patient_id), "\n")

summary_tbl <- paired |>
  group_by(marker) |>
  group_modify(~ paired_spearman(.x$ihc_posfrac, .x$rna_expr)) |>
  ungroup() |>
  arrange(desc(rho))

knitr::kable(summary_tbl, digits = 3,
             caption = "Spearman: IHC positive-cell fraction vs bulk-RNA expression")
```

Scatter per marker: `ihc_posfrac` against `log1p(rna_expr)` (`log1p` keeps zeros),
with an OLS fit line.

```{r concordance-plot, fig.height=8}
if (nrow(paired) > 0) {
  # log-scale expression (counts span orders of magnitude); +1 to keep zeros.
  ggplot(paired, aes(ihc_posfrac, log1p(rna_expr))) +
    geom_point(alpha = 0.6, size = 1.8) +
    geom_smooth(method = "lm", se = FALSE, colour = "steelblue", formula = y ~ x) +
    facet_wrap(~ marker, scales = "free") +
    labs(title = "IHC marker+ fraction vs bulk-RNA expression",
         x = "IHC fraction of cells positive", y = "log1p(bulk expression)") +
    theme_bw(base_size = 10)
}
```

Sensitivity: replace the hard-threshold positive fraction with the per-patient
mean staining intensity (mean z-score over non-NA cells):

```
<marker>_z(p) = (sum over p's cells with a non-NA <marker>_zscore of that zscore)
                / (cells of p with a non-NA <marker>_zscore)     (ihc_marker_fraction)
```

Then correlate `<marker>_z` against `rna_expr` per marker with the same statistic
(Spearman correlation across patients).

```{r zscore-sensitivity}
ihc_z <- ihc_marker_fraction(ihc_data) |>
  select(patient_id, ends_with("_z")) |>
  pivot_longer(-patient_id, names_to = "marker", values_to = "ihc_z") |>
  mutate(marker = sub("_z$", "", marker))

paired_z <- marker_expr |>
  inner_join(ihc_z, by = c("marker", "patient_id")) |>
  filter(is.finite(rna_expr), is.finite(ihc_z))

if (nrow(paired_z) > 0) {
  paired_z |>
    group_by(marker) |>
    group_modify(~ paired_spearman(.x$ihc_z, .x$rna_expr)) |>
    ungroup() |>
    arrange(desc(rho)) |>
    knitr::kable(digits = 3,
                 caption = "Spearman: IHC mean z-score vs bulk-RNA expression")
}
```

## Notes

- Both IHC reductions use `na.rm = TRUE`: `<marker>_posfrac` and `<marker>_z`
  average over the non-NA sign / z-score cells of a patient, so the effective
  denominator is that count, not necessarily all of the patient's cells.
- `rna_expr` averages only genes actually present in `counts_data` (a marker's
  available genes); genes in a marker's list that are absent from the matrix are
  reported by the `missing` message and excluded from the mean. A marker with no
  available genes contributes no RNA row.
- The gene join is `relationship = "many-to-many"`, but each (sample, gene) has a
  single `expr`, so the per-(marker, sample) mean is over distinct genes.
- `PANCK` averages `EPCAM` + keratins; `SMA` -> `ACTA2`, `CD56` -> `NCAM1`,
  `PDL1` -> `CD274`, `PD1` -> `PDCD1`, `FSP1` -> `S100A4`. `wrongL1CAM` is
  excluded by construction (not in `marker_gene_map`).
- Correlations are cross-patient and small-n; every row carries its `n`.

```{r export-pdfs, include=FALSE}
# Copy every plot from this analysis (each rendered as a PDF via the
# dev = c("png", "pdf") chunk option) into output/figures/bulkRna/. Runs last so
# that all figures from the chunks above already exist on disk.
export_pdf_figures("bulkRna")
```
