---
title: "bulkRna"
author: "sceriff0"
date: "2026-07-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: tidyverse + here (already in the lockfile).
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
})

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6)

invisible(capture.output(suppressMessages(source(here("code", "test.R")))))
source(here("code", "validation_helpers.R"))
```

## Starting data and notation

Every value below is derived from three inputs:

- `ihc_data` — FlowPath single cells, deduplicated to **one copy per patient**
  (whole slide, no annotation boundary). Carries per-marker `<marker>_sign` and
  `<marker>_zscore` columns and `patient_id`.
- `counts_data` — DESeq2-normalised **bulk RNA**, one row per RNA sample, one
  column per gene, plus the CRF-space sample id `Sample`.
- `clinical_data` — the clinical CRF, used only for the id crosswalk
  `ID PATIENT` (slide space) $\leftrightarrow$ `ID CRF PRESERVE` (CRF space).

Notation used throughout. $C_p$ = all cells of patient $p$ in `ihc_data`. A cell
$c$ is **marker-positive** for marker $m$ iff `is_pos(`$m$`_sign`$(c))$, i.e.
$m$`_sign`$(c)\in\{+,\text{pos},\text{positive},\text{yes},\text{true},1\}$.
Ids live in two spaces bridged through the clinical CRF:
`crf_to_slide_map(clinical_data)` is the named vector
$\texttt{crf2slide}:\ \texttt{norm\_id}(\texttt{ID CRF PRESERVE})\mapsto
\texttt{slide\_key}(\texttt{ID PATIENT})$, and `slide_key()` is the canonical
slide key (strip non-digits, apply the verified typo fix `15879`$\to$`15897`).
`marker_gene_map` sends each IHC marker $m$ to a comma-separated gene list
$G_m$ (e.g. `CD3`$\to$`{CD3D,CD3E,CD3G}`, `PANCK`$\to$
`{EPCAM,KRT5,KRT8,KRT14,KRT18,KRT19}`); multi-gene markers are averaged. Paired
associations are Spearman $\rho$ via `paired_spearman`.

## Per-patient bulk expression of each marker's gene(s)

RNA samples are keyed in CRF space; `crf2slide` bridges them to slide space.
`counts_data` is melted to long triples $(\texttt{Sample},\ \texttt{crf\_id}=
\texttt{norm\_id}(\texttt{Sample}),\ \texttt{gene},\ \texttt{expr})$, one row per
(sample, gene). `gene_rows` expands `marker_gene_map` to one row per $(m,g)$ for
$g\in G_m$ (comma-split, trimmed); `marker_genes` keeps only $(m,g)$ whose gene
appears in the RNA matrix, i.e. $g\in G_m^{\,\text{RNA}}=G_m\cap
\texttt{genes}(\texttt{counts\_data})$.

Per marker $m$ and CRF sample $s$, expression is the mean over that marker's
available genes:

$$
\texttt{rna\_expr}_{m,s}=\frac{1}{|G_m^{\,\text{RNA}}|}\sum_{g\in G_m^{\,\text{RNA}}}\texttt{expr}(s,g),
$$

then the CRF id is mapped to slide space via
$\texttt{patient\_id}=\texttt{crf2slide}[\texttt{crf\_id}(s)]$ and rows with no
crosswalk match (`NA`) are dropped.

```{r marker-expression}
# RNA samples are CRF-id space; bridge to IHC slide-id space via clinical
# (base-R named-vector lookup, not a join).
crf2slide <- crf_to_slide_map(clinical_data)

expr_long <- counts_data |>
  mutate(crf_id = norm_id(Sample)) |>
  pivot_longer(-c(Sample, crf_id), names_to = "gene", values_to = "expr")

# marker -> its genes, one row per (marker, gene). Explicit $-access + strsplit
# (no NSE) so it is robust to the base pipe / tidyr separate_rows() quirks.
gene_rows <- do.call(rbind, Map(function(m, g)
  data.frame(marker = m, gene = trimws(strsplit(g, ",")[[1]]), stringsAsFactors = FALSE),
  marker_gene_map$marker, marker_gene_map$genes))

missing <- setdiff(gene_rows$gene, unique(expr_long$gene))
if (length(missing)) message("genes absent from RNA matrix: ", paste(missing, collapse = ", "))

# keep only genes present in the RNA matrix
marker_genes <- gene_rows |> filter(gene %in% unique(expr_long$gene))

# Per patient, average expression across a marker's available genes, then map the
# CRF sample id to the IHC slide id via the crosswalk.
marker_expr <- expr_long |>
  inner_join(marker_genes, by = "gene", relationship = "many-to-many") |>
  group_by(marker, crf_id) |>
  summarise(rna_expr = mean(expr), .groups = "drop") |>
  mutate(patient_id = unname(crf2slide[crf_id])) |>   # crf id -> slide id
  filter(!is.na(patient_id)) |>
  select(marker, patient_id, rna_expr)
```

## Per-patient IHC positive-cell fraction

`ihc_marker_fraction(ihc_data)` groups cells by `slide_key(patient_id)` and, for
each marker $m$, returns the whole-slide positive fraction (mean of `is_pos`
over non-`NA` sign values):

$$
m\texttt{\_posfrac}_p=\frac{1}{|C_p|}\sum_{c\in C_p}\mathbf{1}\!\left[\texttt{is\_pos}(m\texttt{\_sign}(c))\right],
$$

taken here for all `_posfrac` columns and reshaped long to
$(\texttt{patient\_id},\ \texttt{marker}=m,\ \texttt{ihc\_posfrac})$.

```{r ihc-fraction}
ihc_frac <- ihc_marker_fraction(ihc_data) |>
  select(patient_id, ends_with("_posfrac")) |>
  pivot_longer(-patient_id, names_to = "marker", values_to = "ihc_posfrac") |>
  mutate(marker = sub("_posfrac$", "", marker))

cat("IHC patients:", dplyr::n_distinct(ihc_frac$patient_id),
    "| RNA patients:", dplyr::n_distinct(marker_expr$patient_id), "\n")
```

## Marker-by-marker concordance

Inner-join `marker_expr` and `ihc_frac` on $(\texttt{marker},\
\texttt{patient\_id})$, keeping finite pairs. Per marker $m$, over its shared
patients $P_m$, the statistic is Spearman's
$\rho=1-\dfrac{6\sum d_i^2}{n(n^2-1)}$ on the ranks of the paired vectors
$x=(\texttt{ihc\_posfrac}_p)_{p\in P_m}$ and $y=(\texttt{rna\_expr}_p)_{p\in
P_m}$ (`paired_spearman`, via `cor.test(method="spearman")`), with its p-value
and $n=|P_m|$; `NA` when $n<3$ or either vector is constant. Rows are sorted by
descending $\rho$.

```{r concordance}
paired <- marker_expr |>
  inner_join(ihc_frac, by = c("marker", "patient_id")) |>
  filter(is.finite(rna_expr), is.finite(ihc_posfrac))

cat("shared patients (RNA + IHC):",
    dplyr::n_distinct(paired$patient_id), "\n")

summary_tbl <- paired |>
  group_by(marker) |>
  group_modify(~ paired_spearman(.x$ihc_posfrac, .x$rna_expr)) |>
  ungroup() |>
  arrange(desc(rho))

knitr::kable(summary_tbl, digits = 3,
             caption = "Spearman: IHC positive-cell fraction vs bulk-RNA expression")
```

Scatter per marker: $x=\texttt{ihc\_posfrac}_p$ against
$y=\log(1+\texttt{rna\_expr}_p)$ (`log1p`, keeping zeros), with an OLS fit line.

```{r concordance-plot, fig.height=8}
if (nrow(paired) > 0) {
  # log-scale expression (counts span orders of magnitude); +1 to keep zeros.
  ggplot(paired, aes(ihc_posfrac, log1p(rna_expr))) +
    geom_point(alpha = 0.6, size = 1.8) +
    geom_smooth(method = "lm", se = FALSE, colour = "steelblue", formula = y ~ x) +
    facet_wrap(~ marker, scales = "free") +
    labs(title = "IHC marker+ fraction vs bulk-RNA expression",
         x = "IHC fraction of cells positive", y = "log1p(bulk expression)") +
    theme_bw(base_size = 10)
}
```

Sensitivity: replace the hard-threshold positive fraction with the per-patient
mean staining intensity (mean z-score over non-`NA` cells),

$$
m\texttt{\_z}_p=\frac{1}{|C_p|}\sum_{c\in C_p}m\texttt{\_zscore}(c)
\quad(\texttt{ihc\_marker\_fraction}),
$$

and correlate $m\texttt{\_z}_p$ against $\texttt{rna\_expr}_p$ per marker with the
same Spearman statistic.

```{r zscore-sensitivity}
ihc_z <- ihc_marker_fraction(ihc_data) |>
  select(patient_id, ends_with("_z")) |>
  pivot_longer(-patient_id, names_to = "marker", values_to = "ihc_z") |>
  mutate(marker = sub("_z$", "", marker))

paired_z <- marker_expr |>
  inner_join(ihc_z, by = c("marker", "patient_id")) |>
  filter(is.finite(rna_expr), is.finite(ihc_z))

if (nrow(paired_z) > 0) {
  paired_z |>
    group_by(marker) |>
    group_modify(~ paired_spearman(.x$ihc_z, .x$rna_expr)) |>
    ungroup() |>
    arrange(desc(rho)) |>
    knitr::kable(digits = 3,
                 caption = "Spearman: IHC mean z-score vs bulk-RNA expression")
}
```

## Notes

- Both IHC reductions use `na.rm = TRUE`: $m\texttt{\_posfrac}_p$ and
  $m\texttt{\_z}_p$ average over the non-`NA` sign / z-score cells of $C_p$, so
  the effective denominator is that count, not necessarily $|C_p|$.
- `rna_expr` averages only genes actually present in `counts_data`
  ($G_m^{\,\text{RNA}}$); genes in $G_m$ absent from the matrix are reported by
  the `missing` message and excluded from the mean. A marker with
  $G_m^{\,\text{RNA}}=\varnothing$ contributes no RNA row.
- The gene join is `relationship = "many-to-many"`, but each $(s,g)$ has a single
  `expr`, so the per-$(m,s)$ mean is over distinct genes.
- `PANCK` averages `EPCAM` + keratins; `SMA`$\to$`ACTA2`, `CD56`$\to$`NCAM1`,
  `PDL1`$\to$`CD274`, `PD1`$\to$`PDCD1`, `FSP1`$\to$`S100A4`. `wrongL1CAM` is
  excluded by construction (not in `marker_gene_map`).
- Correlations are cross-patient and small-n; every row carries its $n$.
