---
title: "bulkRna"
author: "sceriff0"
date: "2026-07-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: tidyverse + here (already in the lockfile).
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
})

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6)

invisible(capture.output(suppressMessages(source(here("code", "test.R")))))
source(here("code", "validation_helpers.R"))
```

## Purpose

Cross-modality validation of the IHC readout against **bulk-RNA marker genes**: if
IHC protein positivity is real, the fraction of cells positive for a marker should
track the bulk-RNA expression of that marker's gene across patients. This is the
most direct check — one protein against its own transcript — and needs no
deconvolution model.

For each IHC marker we take its canonical gene(s) (`marker_gene_map`), average
multi-gene markers, and correlate bulk expression vs the per-patient fraction of
`<marker>_sign == "+"` cells.

## Per-patient bulk expression of each marker's gene(s)

```{r marker-expression}
# RNA samples are CRF-id space; bridge to IHC slide-id space via clinical.
xwalk <- id_crosswalk(clinical_data)

expr_long <- counts_data |>
  mutate(crf_id = norm_id(Sample)) |>
  pivot_longer(-c(Sample, crf_id), names_to = "gene", values_to = "expr")

# marker -> its genes, one row per (marker, gene). Explicit $-access + strsplit
# (no NSE) so it is robust to the base pipe / tidyr separate_rows() quirks.
gene_rows <- do.call(rbind, Map(function(m, g)
  data.frame(marker = m, gene = trimws(strsplit(g, ",")[[1]]), stringsAsFactors = FALSE),
  marker_gene_map$marker, marker_gene_map$genes))

missing <- setdiff(gene_rows$gene, unique(expr_long$gene))
if (length(missing)) message("genes absent from RNA matrix: ", paste(missing, collapse = ", "))

# keep only genes present in the RNA matrix
marker_genes <- gene_rows |> filter(gene %in% unique(expr_long$gene))

# Per patient, average expression across a marker's available genes, then map the
# CRF sample id to the IHC slide id via the crosswalk.
marker_expr <- expr_long |>
  inner_join(marker_genes, by = "gene", relationship = "many-to-many") |>
  group_by(marker, crf_id) |>
  summarise(rna_expr = mean(expr), .groups = "drop") |>
  inner_join(xwalk, by = "crf_id") |>
  transmute(marker, patient_id = slide_id, rna_expr)
```

## Per-patient IHC positive-cell fraction

```{r ihc-fraction}
ihc_frac <- ihc_marker_fraction(ihc_data) |>
  select(patient_id, ends_with("_posfrac")) |>
  pivot_longer(-patient_id, names_to = "marker", values_to = "ihc_posfrac") |>
  mutate(marker = sub("_posfrac$", "", marker))

cat("IHC patients:", dplyr::n_distinct(ihc_frac$patient_id),
    "| RNA patients:", dplyr::n_distinct(marker_expr$patient_id), "\n")
```

## Marker-by-marker concordance

```{r concordance}
paired <- marker_expr |>
  inner_join(ihc_frac, by = c("marker", "patient_id")) |>
  filter(is.finite(rna_expr), is.finite(ihc_posfrac))

cat("shared patients (RNA + IHC):",
    dplyr::n_distinct(paired$patient_id), "\n")

summary_tbl <- paired |>
  group_by(marker) |>
  group_modify(~ paired_spearman(.x$ihc_posfrac, .x$rna_expr)) |>
  ungroup() |>
  arrange(desc(rho))

knitr::kable(summary_tbl, digits = 3,
             caption = "Spearman: IHC positive-cell fraction vs bulk-RNA expression")
```

```{r concordance-plot, fig.height=8}
if (nrow(paired) > 0) {
  # log-scale expression (counts span orders of magnitude); +1 to keep zeros.
  ggplot(paired, aes(ihc_posfrac, log1p(rna_expr))) +
    geom_point(alpha = 0.6, size = 1.8) +
    geom_smooth(method = "lm", se = FALSE, colour = "steelblue", formula = y ~ x) +
    facet_wrap(~ marker, scales = "free") +
    labs(title = "IHC marker+ fraction vs bulk-RNA expression",
         x = "IHC fraction of cells positive", y = "log1p(bulk expression)") +
    theme_bw(base_size = 10)
}
```

Sensitivity: the mean marker z-score (continuous staining intensity) instead of the
positive-cell fraction, in case a hard `_sign` threshold loses signal.

```{r zscore-sensitivity}
ihc_z <- ihc_marker_fraction(ihc_data) |>
  select(patient_id, ends_with("_z")) |>
  pivot_longer(-patient_id, names_to = "marker", values_to = "ihc_z") |>
  mutate(marker = sub("_z$", "", marker))

paired_z <- marker_expr |>
  inner_join(ihc_z, by = c("marker", "patient_id")) |>
  filter(is.finite(rna_expr), is.finite(ihc_z))

if (nrow(paired_z) > 0) {
  paired_z |>
    group_by(marker) |>
    group_modify(~ paired_spearman(.x$ihc_z, .x$rna_expr)) |>
    ungroup() |>
    arrange(desc(rho)) |>
    knitr::kable(digits = 3,
                 caption = "Spearman: IHC mean z-score vs bulk-RNA expression")
}
```

## Notes

- Expect strong positive correlation for lineage-specific markers (PTPRC/CD45,
  CD3, CD8A, CD163, NCAM1/CD56) where protein and transcript co-vary; weaker or
  absent for markers regulated post-transcriptionally or with poor antibody
  specificity — a low `rho` flags an IHC channel worth auditing.
- `PANCK` maps to `EPCAM` + keratin genes (averaged); `SMA`→`ACTA2`,
  `CD56`→`NCAM1`, `PDL1`→`CD274`, `PD1`→`PDCD1`, `FSP1`→`S100A4`. `wrongL1CAM` is
  excluded by construction.
- Correlations are cross-patient and small-n; every row carries its `n`.
