---
title: "clinical_data"
author: "sceriff0"
date: "2026-07-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: needs sf + jsonlite for the geojson point-in-polygon (+ tidyverse, here).
#   renv::install(c("sf", "jsonlite")); renv::snapshot()
#   Survival QC additionally uses survival + survminer (guarded at use):
#   renv::install(c("survival", "survminer")); renv::snapshot()
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(sf)
})

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6)

# Loaders (dds, clinical_data, neoplastic_data, counts_data, ihc_data) and the
# shared validation helpers. test.R prints colnames() on load; silence it.
invisible(capture.output(suppressMessages(source(here("code", "test.R")))))
source(here("code", "validation_helpers.R"))
```

## Purpose

Validate the FlowPath IHC readout against the **pathologist**. Two independent
checks:

1. **Neoplastic cellularity** — IHC tumour fraction inside each annotation vs the
   pathologist's `neoplastic_data` estimate. Cell membership in the annotation is
   recomputed from the raw geojson with `sf` (not the CSV `Out_of_annotation`
   flag), and reported both per-annotation and over the dissolved union.
2. **Immune infiltrate** — IHC CD45/lineage fractions inside the tumour region vs
   the clinical `HOT score` / `Immuno-phenotype`.
3. **Biological-validity QC** — do the IHC-derived metrics behave the way real
   biology predicts against the wider clinical table? Three cross-checks:
   peripheral-blood immunity (lymphocytes / NLR / PNI), induction-treatment
   `RESPONSE`, and **survival** (`OS`, `PFS`). A genuine immune readout should
   track peripheral lymphocytes, better response, and longer survival.

Tumour cell = `phenotype_clean` containing "Tumor" (`PANCK+Tumor`, `VIM+Tumor`).

Checks 1–2 use the annotation region (sf); check 3 uses **whole-slide** IHC immune
metrics so it covers every imaged patient, not only the handful with pathologist
annotations.

## Load annotations and assign cells (sf point-in-polygon)

```{r annotations}
# IHC, neoplastic and geojson all live in the SLIDE-id space (numeric codes).
neo_ids <- slide_key(neoplastic_data$SAMPLE)
annots  <- load_annotations(patient_ids = neoplastic_data$SAMPLE)

cat("annotation polygons loaded:", nrow(annots),
    "for", dplyr::n_distinct(annots$patient_id), "patients\n")

ihc_ids       <- unique(slide_key(ihc_data$patient_id))
have_geojson  <- unique(slide_key(annots$patient_id))
no_geojson    <- setdiff(intersect(neo_ids, ihc_ids), have_geojson)
cat("patients with BOTH IHC cells and pathologist score:",
    length(intersect(neo_ids, ihc_ids)), "\n")
cat("  of which via sf polygons:", length(intersect(neo_ids, have_geojson)),
    "| via CSV Out_of_annotation fallback:", length(no_geojson), "\n")
if (length(intersect(neo_ids, ihc_ids)) == 0)
  warning("no overlap between ihc_data patients and neoplastic_data — check IDs")
```

## Tumour content — IHC vs pathologist

The pathologist scores tumour content per annotation (`ANNOTATION_1..3`). Compare
against the IHC tumour fraction computed inside the matching polygon.

```{r tumor-content}
ihc_tumor_per_ann <- ihc_annotation_metrics(ihc_data, annots, scope = "per_annotation")
ihc_tumor_union   <- ihc_annotation_metrics(ihc_data, annots, scope = "union")

path_long <- neoplastic_data |>
  mutate(patient_id = slide_key(SAMPLE)) |>
  pivot_longer(starts_with("ANNOTATION_"),
               names_to = "annotation", values_to = "path_pct") |>
  filter(!is.na(path_pct)) |>
  mutate(path_frac = path_pct / 100)

# Per-annotation concordance (join on patient + annotation index).
cc_ann <- ihc_tumor_per_ann |>
  inner_join(path_long, by = c("patient_id", "annotation")) |>
  filter(!is.na(tumor_over_inside), !is.na(path_frac))

# Union concordance: pathologist reference = mean of that patient's annotations.
path_union <- path_long |>
  group_by(patient_id) |>
  summarise(path_frac = mean(path_frac), .groups = "drop")
cc_union <- ihc_tumor_union |>
  inner_join(path_union, by = "patient_id") |>
  filter(!is.na(tumor_over_inside), !is.na(path_frac))

cat("per-annotation pairs:", nrow(cc_ann), " | union (per-patient) pairs:", nrow(cc_union), "\n")
```

```{r tumor-content-plot}
plot_concordance <- function(df, subtitle) {
  st <- paired_spearman(df$tumor_over_inside, df$path_frac)
  ggplot(df, aes(tumor_over_inside, path_frac)) +
    geom_abline(slope = 1, linetype = "dashed", colour = "red") +
    geom_point(alpha = 0.7, size = 2) +
    coord_equal(xlim = c(0, 1), ylim = c(0, 1)) +
    annotate("text", x = 0, y = 1, hjust = 0, vjust = 1,
             label = sprintf("Spearman rho = %.2f (p = %.3g, n = %d)",
                             st$rho, st$p, st$n)) +
    labs(title = "Neoplastic cellularity: IHC vs pathologist", subtitle = subtitle,
         x = "IHC tumour fraction inside annotation",
         y = "Pathologist tumour fraction") +
    theme_classic(base_size = 12)
}

if (nrow(cc_ann)   > 0) print(plot_concordance(cc_ann,   "per annotation"))
if (nrow(cc_union) > 0) print(plot_concordance(cc_union, "union of a patient's annotations"))
```

Bland–Altman agreement (both quantities are on the same 0–1 scale, so the
difference vs mean shows systematic bias, not just rank correlation).

```{r tumor-content-bland-altman}
if (nrow(cc_union) > 2) {
  ba <- cc_union |>
    transmute(mean_val = (tumor_over_inside + path_frac) / 2,
              diff_val = tumor_over_inside - path_frac)
  bias <- mean(ba$diff_val); loa <- 1.96 * sd(ba$diff_val)
  ggplot(ba, aes(mean_val, diff_val)) +
    geom_hline(yintercept = bias, colour = "blue") +
    geom_hline(yintercept = bias + c(-loa, loa), linetype = "dashed", colour = "grey40") +
    geom_point(alpha = 0.7, size = 2) +
    labs(title = "Bland–Altman: IHC − pathologist tumour fraction",
         subtitle = sprintf("bias = %.3f, limits of agreement +/- %.3f", bias, loa),
         x = "mean of the two", y = "IHC − pathologist") +
    theme_classic(base_size = 12)
}
```

## Immune infiltrate — IHC vs clinical immune phenotype

IHC immune infiltrate = CD45+ (and lineage) fractions inside the tumour region
(union of annotations). Compare against the clinician's `HOT score` (continuous)
and `Immuno-phenotype` (categorical).

```{r infiltrate-join}
# IHC metrics are slide-space; clinical is keyed to slides via `ID PATIENT`.
clin <- clinical_data |>
  mutate(patient_id = slide_key(`ID PATIENT`)) |>
  select(patient_id,
         hot_score  = any_of("HOT score"),
         immuno_phe = any_of("Immuno-phenotype"))

infiltrate <- ihc_tumor_union |>
  select(patient_id, cd45_over_inside, frac_CD8T, frac_CD4T, frac_Treg, frac_NK) |>
  inner_join(clin, by = "patient_id")

cat("patients with IHC infiltrate + clinical immune phenotype:", nrow(infiltrate), "\n")
```

```{r infiltrate-hotscore}
if ("hot_score" %in% names(infiltrate) &&
    sum(is.finite(suppressWarnings(as.numeric(infiltrate$hot_score)))) > 2) {
  df <- infiltrate |> mutate(hot_score = suppressWarnings(as.numeric(hot_score)))
  st <- paired_spearman(df$cd45_over_inside, df$hot_score)
  ggplot(df, aes(cd45_over_inside, hot_score)) +
    geom_point(alpha = 0.7, size = 2) +
    geom_smooth(method = "lm", se = FALSE, colour = "steelblue", formula = y ~ x) +
    annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
             label = sprintf("Spearman rho = %.2f (p = %.3g, n = %d)",
                             st$rho, st$p, st$n)) +
    labs(title = "Immune infiltrate vs clinical HOT score",
         x = "IHC CD45+ fraction inside tumour region", y = "clinical HOT score") +
    theme_classic(base_size = 12)
}
```

```{r infiltrate-phenotype}
if ("immuno_phe" %in% names(infiltrate) &&
    dplyr::n_distinct(na.omit(infiltrate$immuno_phe)) > 1) {
  df <- infiltrate |>
    pivot_longer(c(cd45_over_inside, frac_CD8T, frac_CD4T, frac_Treg, frac_NK),
                 names_to = "ihc_metric", values_to = "frac") |>
    filter(!is.na(immuno_phe), !is.na(frac))
  ggplot(df, aes(immuno_phe, frac, colour = immuno_phe)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.2, alpha = 0.6) +
    facet_wrap(~ ihc_metric, scales = "free_y") +
    labs(title = "IHC infiltrate metrics by clinical immuno-phenotype",
         subtitle = sprintf("%d patients with both IHC infiltrate and immuno-phenotype",
                            dplyr::n_distinct(df$patient_id)),
         x = NULL, y = "fraction inside tumour region", colour = "Immuno-phenotype") +
    theme_classic(base_size = 11) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
}
```

## Biological-validity QC — IHC immune readout vs the wider clinical table

Whole-slide IHC immune metrics per patient (covers every imaged patient, not just
the annotated ones): lineage fractions over all cells plus CD45/CD8/FOXP3
positive-cell fractions.

```{r whole-slide-immune}
ihc_lin <- ihc_lineage_fraction(ihc_data) |>
  filter(lineage %in% comparable_lineages) |>
  select(patient_id, lineage, frac) |>
  pivot_wider(names_from = lineage, values_from = frac, values_fill = 0)

ihc_mk <- ihc_marker_fraction(ihc_data) |>
  select(patient_id, any_of(c("CD45_posfrac", "CD8_posfrac", "FOXP3_posfrac")))

ihc_immune <- full_join(ihc_lin, ihc_mk, by = "patient_id")

# Full clinical table on the shared key (slide space via `ID PATIENT`).
clin_full <- clinical_data |>
  mutate(patient_id = slide_key(`ID PATIENT`))

ihc_clin <- ihc_immune |> inner_join(clin_full, by = "patient_id")
cat("patients with whole-slide IHC immune metrics + clinical row:", nrow(ihc_clin), "\n")

# Primary IHC immune summary used throughout: CD45+ fraction if present, else CD8T.
# Guarded because a lineage that never occurs would leave no pivoted column.
score_col <- intersect(c("CD45_posfrac", "CD8T"), names(ihc_clin))[1]
ihc_clin$ihc_immune_score <- if (!is.na(score_col)) ihc_clin[[score_col]] else NA_real_
```

### Tissue vs peripheral-blood immunity

A real tissue immune infiltrate should track the patient's circulating lymphocytes
and run opposite to the neutrophil-to-lymphocyte ratio (NLR); the prognostic
nutritional index (PNI) is a related systemic-immunity/nutrition score.

```{r blood-immunity}
blood <- ihc_clin |>
  transmute(
    patient_id, ihc_immune_score,
    lymphocytes = suppressWarnings(as.numeric(`BASAL VALUE OF LYMPHOCYTES`)),
    neutrophils = suppressWarnings(as.numeric(`BASAL VALUE OF NEUTROPHILS`)),
    pni         = suppressWarnings(as.numeric(`PROGNOSTIC NUTRITIONAL INDEX`))
  ) |>
  mutate(nlr = neutrophils / lymphocytes)

blood_long <- blood |>
  pivot_longer(c(lymphocytes, nlr, pni), names_to = "blood_metric", values_to = "value") |>
  filter(is.finite(ihc_immune_score), is.finite(value))

if (nrow(blood_long) > 2) {
  labs_tbl <- blood_long |>
    group_by(blood_metric) |>
    group_modify(~ paired_spearman(.x$ihc_immune_score, .x$value)) |>
    mutate(lab = sprintf("rho = %.2f (p = %.3g, n = %d)", rho, p, n))
  ggplot(blood_long, aes(ihc_immune_score, value)) +
    geom_point(alpha = 0.7, size = 2) +
    geom_smooth(method = "lm", se = FALSE, colour = "steelblue", formula = y ~ x) +
    geom_text(data = labs_tbl, aes(x = -Inf, y = Inf, label = lab),
              hjust = -0.05, vjust = 1.4, inherit.aes = FALSE, size = 3) +
    facet_wrap(~ blood_metric, scales = "free_y") +
    labs(title = "IHC immune score vs peripheral-blood immunity",
         x = "IHC immune score (CD45+ fraction)", y = NULL) +
    theme_classic(base_size = 11)
}
```

### Induction-treatment response

Immunologically "hot" tumours tend to respond better to induction chemotherapy.

```{r response}
if ("RESPONSE" %in% names(ihc_clin)) {
  df <- ihc_clin |>
    filter(!is.na(RESPONSE), is.finite(ihc_immune_score))
  # order response by descending frequency (base R; avoids forcats on this env)
  df$RESPONSE <- factor(as.character(df$RESPONSE),
                        levels = names(sort(table(df$RESPONSE), decreasing = TRUE)))
  if (dplyr::n_distinct(df$RESPONSE) > 1 && nrow(df) > 3) {
    ggplot(df, aes(RESPONSE, ihc_immune_score, colour = RESPONSE)) +
      geom_boxplot(outlier.shape = NA) +
      geom_jitter(width = 0.2, alpha = 0.6) +
      labs(title = "IHC immune score by induction response",
           subtitle = sprintf("%d patients with both IHC score and response", nrow(df)),
           x = NULL, y = "IHC immune score (CD45+ fraction)") +
      theme_classic(base_size = 12) +
      theme(axis.text.x = element_text(angle = 30, hjust = 1), legend.position = "none")
  }
}
```

### Survival — does the IHC immune readout carry prognostic signal?

Kaplan–Meier by median split of the IHC immune score, plus a univariable Cox HR on
the continuous score, for `OS` and `PFS`.

**Assumption to verify:** the time columns are `OS` / `PFS` (numeric months) and the
event columns are `FOLLOW UP` (death) / `PERSISTENCE OR RELAPSE?` (progression).
`coerce_event()` maps common encodings (yes/no, alive/dead, 1/0) to 0/1 — adjust
`surv_cfg` if your coding differs.

```{r survival}
surv_cfg <- list(
  OS  = list(time = "OS",  event = "FOLLOW UP"),
  PFS = list(time = "PFS", event = "PERSISTENCE OR RELAPSE?")
)

coerce_event <- function(x) {
  if (is.numeric(x)) return(as.integer(x > 0))
  v <- tolower(trimws(as.character(x)))
  dplyr::case_when(
    v %in% c("1","yes","y","dead","death","deceased","event","relapse",
             "progression","progressed","true") ~ 1L,
    v %in% c("0","no","n","alive","ned","censored","false") ~ 0L,
    TRUE ~ NA_integer_
  )
}

run_survival <- function(cfg, label) {
  if (!requireNamespace("survival", quietly = TRUE)) {
    message("survival package not installed — skipping ", label); return(invisible())
  }
  if (!all(c(cfg$time, cfg$event) %in% names(ihc_clin))) {
    message("missing ", cfg$time, "/", cfg$event, " — skipping ", label); return(invisible())
  }
  d <- ihc_clin |>
    transmute(ihc_immune_score,
              .time  = suppressWarnings(as.numeric(.data[[cfg$time]])),
              .event = coerce_event(.data[[cfg$event]])) |>
    filter(is.finite(.time), !is.na(.event), is.finite(ihc_immune_score))
  if (nrow(d) < 5 || dplyr::n_distinct(d$.event) < 2) {
    message("too few informative rows for ", label, " (n = ", nrow(d), ")"); return(invisible())
  }
  d <- d |> mutate(immune_group = if_else(ihc_immune_score >= median(ihc_immune_score),
                                          "IHC-immune high", "IHC-immune low"))
  fit  <- survival::survfit(survival::Surv(.time, .event) ~ immune_group, data = d)
  cox  <- survival::coxph(survival::Surv(.time, .event) ~ ihc_immune_score, data = d)
  hr   <- exp(stats::coef(cox))[1]; pcox <- summary(cox)$coefficients[1, "Pr(>|z|)"]
  cat(sprintf("%s: Cox HR per unit IHC immune score = %.2f (p = %.3g, n = %d)\n",
              label, hr, pcox, nrow(d)))
  if (requireNamespace("survminer", quietly = TRUE)) {
    print(survminer::ggsurvplot(fit, data = d, pval = TRUE, risk.table = TRUE,
                                title = paste(label, "by IHC immune score (median split)"),
                                legend.title = "")$plot)
  } else {
    plot(fit, col = c("firebrick", "steelblue"), xlab = "time", ylab = "survival",
         main = paste(label, "by IHC immune score"))
    legend("topright", levels(factor(d$immune_group)),
           col = c("firebrick", "steelblue"), lty = 1, bty = "n")
  }
}

invisible(purrr::iwalk(surv_cfg, ~ run_survival(.x, .y)))
```

## Notes

- Cohorts are small (the pathologist scored a handful of slides); Spearman is used
  throughout and every correlation reports its `n`. Read associations as
  descriptive, not inferential.
- Per-annotation vs union: per-annotation tests each scored region directly; the
  union collapses a patient to one tumour-content value (pathologist reference =
  mean of their annotations).
- Membership is `sf::st_within` on the raw geojson — a cell counts as inside only
  if its centroid falls in the polygon, independent of the CSV flag.
- Biological-validity QC (blood / response / survival) is exploratory and
  underpowered at this cohort size — it checks the *direction* the IHC immune
  readout points (with peripheral lymphocytes, with response, with survival), not
  statistical significance. Confirm the `surv_cfg` time/event columns match your
  data coding before trusting the KM/Cox output.
