---
title: "clinical_data (per-annotation flag, old/ overlay)"
author: "sceriff0"
date: "2026-07-20"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(sf)
})

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6,
                      dev = c("png", "pdf"))   # PNG for the site, PDF exported to output/

# Loaders (dds, clinical_data, neoplastic_data, counts_data, ihc_data) and the shared
# validation helpers. test.R prints colnames() on load; silence it. Cells come from
# the per-annotation flag files via per_annotation_flag.R (sourced after the helpers
# it depends on); the whole-slide `ihc_data` is used ONLY as the fallback for
# patients that have no per-annotation csv.
invisible(capture.output(suppressMessages(source(here("code", "test.R")))))
source(here("code", "validation_helpers.R"))
source(here("code", "aggregation_compare.R"))  # annotation-aggregation sensitivity grid
source(here("code", "per_annotation_flag.R"))  # per-annotation flag loader + metrics
source(here("code", "pdf_export.R"))          # export_pdf_figures(): plots -> output/figures/
```

## Starting data and notation

This is the **per-annotation flag, `old/`-overlay** variant of `clinical_data.Rmd`:
the cells and the tumour-membership come from the FlowPath per-annotation exports,
not the whole-slide files + geojson. It is the counterpart of
`clinical_data_per_annotation.Rmd`, which is identical except that it ignores
`per_annotation/old/` entirely — so every difference between the two reports is
attributable to the overlaid annotations alone. Every value below is derived from
three inputs:

- `ann_cells` — FlowPath single cells, one row per cell, assembled from **three**
  places by `build_flag_cells()`:
    - *primary* — the per-annotation files
      `data/flowpath/per_annotation/<patient>_a<k>.csv` (**one csv per (patient,
      annotation)**; the `new/` subfolder is not read). Each file carries
      `phenotype`, per-marker `<marker>_sign`, `centroid_x`/`y`, and the
      precomputed `Out_of_annotation` flag for that annotation. `patient_id` = the
      stem before `_a`; `annotation` = `ANNOTATION_<k>`.
    - *overlay* — `data/flowpath/per_annotation/old/<patient>_a<k>.csv`. Where the
      overlay provides a `(patient, annotation)`, it **replaces** the top-level file
      for that annotation outright: the two csvs are alternative exports of the same
      annotation, so their cells are swapped, never pooled (pooling would
      double-count every cell). Annotations the overlay does not mention keep their
      top-level file untouched.
    - *fallback* — for a patient with **no** per-annotation csv at all, the
      whole-slide file `data/flowpath/<patient>.csv`. Its single
      `Out_of_annotation` flag was computed over all that patient's annotations at
      once, so it gives **one** region. When the patient has exactly one scored
      annotation that region *is* that annotation and is labelled accordingly
      (`ANNOTATION_1` for both `10338` and `15897`), so the patient reaches every
      panel; otherwise it is labelled `"csv"` — see the note below.
- `neoplastic_data` — pathologist tumour-content percentage per annotation,
  `ANNOTATION_1..3`.
- `clinical_data` — the clinical CRF (`HOT score`, `Immuno-phenotype`, blood counts,
  `RESPONSE`, `OS`/`PFS`).

Membership (which cells count as "inside the tumour annotation"):

```
cell is INSIDE = Out_of_annotation is FALSE   (the precomputed FlowPath flag)

per-annotation patients                    source = "flag"
  per annotation = inside cells of file <patient>_a<k>.csv   (annotation ANNOTATION_k)
  union          = inside cells pooled across a patient's annotation files,
                   deduped to one row per cell (cell_id if present, else centroid)

fallback patients (no per-annotation csv)  source = "csv"
  one region only = inside cells of <patient>.csv
  labelled        = the patient's sole scored ANNOTATION_k, when it has exactly one
                    (true for 10338 and 15897); otherwise "csv"
  union           = that same single region
```

There is no sf point-in-polygon and no geojson anywhere in this analysis — both
sources are the precomputed flag, which is why the fallback is the flag column of
the whole-slide csv rather than `clinical_data.Rmd`'s polygon path. With no polygon,
annotation area is unknown, so every `dens_*_per_mm2` column (and `area_mm2`) is
`NA` — the density panels are empty and only the fraction-based comparisons carry
values.

**Why the fallback region can be labelled with an annotation.** The per-annotation
panels join IHC to the pathologist on `(patient_id, annotation)`, so a region
labelled `"csv"` would never match `ANNOTATION_1..3` and the patient would silently
drop out. That is avoidable exactly when the patient has **one** scored annotation:
the whole-slide flag region and the pathologist's single scored annotation are then
the same region, so labelling it `ANNOTATION_1` compares like with like. Both
fallback patients here (`10338`, `15897`) are in that case, so **all six patients
appear in every panel**.

The guard matters for future data. If a fallback patient ever has two or more scored
annotations, its one flag region spans all of them and cannot be attributed to any
single `ANNOTATION_k` — joining it to one score would compare a whole-slide region
against a part-region score. `load_flag_fallback_cells()` then falls back to the
`"csv"` label **and warns**, and that patient appears in the union panels only (the
union score is the mean over its annotations, which does match). Check the knit log
for such a warning whenever the cohort changes.

Either way `source` stays `"csv"`, so a fallback patient is never disguised as a
per-annotation one — the inventory table and the `source` column on every metrics
frame still tell you which membership each row came from.

Cell labels:

```
tumour cell = phenotype_clean contains "Tumor"
              (phenotype_clean = the text inside the parentheses of `phenotype`)
CD45+ cell  = CD45_sign is "+"  (is_pos accepts +, pos, positive, yes, true, 1)
lineage(cell) = phenotype_clean collapsed via the phenotype_lineage table
```

`slide_key()` maps every id to the shared numeric slide space (strip non-digits,
apply the verified typo fix `15879` -> `15897`).

## Membership counts

For a set of cells (either one annotation's inside cells or the union inside
cells), `region_ratios()` returns:

```
n_inside          = number of cells in the set
n_inside_clean    = cells in the set after dropping Outlier == TRUE and
                    Unknown-phenotype cells (QC-filtered denominator)
n_tumor_inside    = tumour cells in the set
n_tumor_clean     = tumour cells in the set that survive the same clean filter
n_cd45_inside     = CD45+ cells in the set
n_cd3cd45_inside  = CD45+ AND CD3+ cells (marker-gated T cells)
n_gzmb_nk_inside  = NK-lineage cells that are GZMB+ (cytotoxic/activated NK)
n_<lineage>       = cells in the set with that lineage
                    (lineage in: CD8T, CD4T, Treg, NK, Immune_other, Stroma)

tumor_over_inside       = n_tumor_inside / n_inside
tumor_over_inside_clean = n_tumor_clean  / n_inside_clean  (outliers + Unknown removed)
cd45_over_inside    = n_cd45_inside    / n_inside
cd3cd45_over_inside = n_cd3cd45_inside / n_inside
cd3cd45_over_cd45   = n_cd3cd45_inside / n_cd45_inside   (T cells among immune)
gzmb_nk_over_inside = n_gzmb_nk_inside / n_inside
gzmb_nk_over_nk     = n_gzmb_nk_inside / n_NK            (GZMB+ fraction of NK)
frac_<lineage>      = n_<lineage>      / n_inside
```

Any fraction with a zero denominator is `NA`. The marker-gated subsets use
`<marker>_sign` (`is_pos`), not the phenotype label.

```{r annotations}
# Cells for the whole analysis: per-annotation files with old/ layered OVER them
# (old/ wins on collision), plus the whole-slide Out_of_annotation fallback for any
# patient with no per-annotation csv. One long table keyed by (patient_id, annotation).
# old/ overrides the top-level file. neoplastic_data lets a fallback patient with a
# single scored annotation be labelled with that annotation, so it reaches every panel.
ann_cells <- build_flag_cells(ihc_data, overlay = PER_ANNOT_OLD_DIR,
                              neoplastic_data = neoplastic_data)

inv <- flag_cells_inventory(ann_cells)

neo_ids  <- slide_key(neoplastic_data$SAMPLE)
from_old <- inv |> filter(file_origin == "old")
per_ann  <- sort(unique(inv$patient_id[inv$file_origin %in% c("per_annotation", "old")]))
fallback <- sort(unique(inv$patient_id[inv$file_origin == "whole_slide"]))
missing  <- setdiff(neo_ids, unique(inv$patient_id))  # scored, but no cells at all

cat("per-annotation patients:", length(per_ann), "[", paste(per_ann, collapse = ", "), "]",
    "across", sum(inv$file_origin %in% c("per_annotation", "old")), "annotation file(s)\n")
cat("annotations taken from old/ (superseding the top-level file):", nrow(from_old),
    "[", paste(paste0(from_old$patient_id, " ", from_old$annotation), collapse = "; "), "]\n")
cat("whole-slide FALLBACK patients (no per-annotation csv):", length(fallback),
    "[", paste(fallback, collapse = ", "), "]\n")
cat("patients with a pathologist score AND cells:",
    length(intersect(neo_ids, unique(inv$patient_id))), "\n")
if (length(missing))
  cat("pathologist patients with NO csv at all (absent from every IHC metric):",
      paste(sort(missing), collapse = ", "), "\n")

knitr::kable(inv, caption = paste(
  "Cell inventory. file_origin = per_annotation (top-level <patient>_a<k>.csv),",
  "old (that file superseded by per_annotation/old/), or whole_slide (the fallback",
  "<patient>.csv, annotation \"csv\"). n_inside = cells with Out_of_annotation FALSE."))
```

## Tumour content — IHC vs pathologist

The IHC tumour metric is shown under **three normalizations**, all restricted to the
cells inside the annotation:

```
tumor_over_inside       = (tumour cells inside) / (all cells inside)     -> 0..1
tumor_over_inside_clean = (tumour cells inside) / (cells inside)  -> 0..1
                          with the cell set first cleaned: FlowPath Outlier == TRUE
                          cells and Unknown-phenotype cells dropped from BOTH the
                          numerator and the denominator (a QC-filtered version of
                          tumor_over_inside; if a csv has no `Outlier` column no cell
                          is treated as an outlier)
tumor_over_cd45         = (tumour cells inside) / (CD45+ cells inside)   -> can exceed 1
dens_tumor_per_mm2      = (tumour cells inside) / (annotation area in mm^2)
                          NA throughout here — no polygon is read, so no area

pathologist (per annotation) = ANNOTATION_k / 100
pathologist (union)          = mean of that patient's ANNOTATION_k / 100 values
```

`tumor_over_inside_clean` answers the same IHC-vs-pathologist question as
`tumor_over_inside` but removes debris/segmentation outliers and unclassified cells
that dilute the denominator, so it is expected to track the pathologist's tumour
fraction more closely.

`dens_tumor_per_mm2` is the tumour-cell density (cells/mm²). It needs the polygon
area, which this flag-based variant does not have, so it is **`NA` throughout** and
the density panel is empty. It is kept only so the schema matches `clinical_data.Rmd`.

Each normalization is computed both per-annotation (one polygon) and over the union
of a patient's polygons.

```{r tumor-content}
# Membership = the Out_of_annotation flag in each per-annotation file. Same output
# schema as ihc_annotation_metrics(); area_mm2 / dens_*_per_mm2 are NA (no polygon).
ihc_tumor_per_ann <- ihc_flag_metrics_per_annotation(ann_cells)
ihc_tumor_union   <- ihc_flag_metrics_union(ann_cells)

path_long <- neoplastic_data |>
  mutate(patient_id = slide_key(SAMPLE)) |>
  pivot_longer(starts_with("ANNOTATION_"),
               names_to = "annotation", values_to = "path_pct") |>
  filter(!is.na(path_pct)) |>
  mutate(path_frac = path_pct / 100)

cc_ann <- ihc_tumor_per_ann |>
  inner_join(path_long, by = c("patient_id", "annotation")) |>
  filter(!is.na(tumor_over_inside), !is.na(path_frac))

path_union <- path_long |>
  group_by(patient_id) |>
  summarise(path_frac = mean(path_frac), .groups = "drop")
cc_union <- ihc_tumor_union |>
  inner_join(path_union, by = "patient_id") |>
  filter(!is.na(tumor_over_inside), !is.na(path_frac))

cat("per-annotation pairs:", nrow(cc_ann), " | union pairs:", nrow(cc_union), "\n")

# Diagnostic: the ACTUAL pairs that enter the correlation, largest gap first. Read
# the top rows to tell WHY the correlation is what it is:
#   - big gap + tiny n_inside          -> suspect membership/flag, not biology
#   - big gap + healthy cell count     -> genuine IHC/pathologist disagreement
#   - per-annotation: verify each file _a<k> lines up with pathologist ANNOTATION_k
cc_union |>
  transmute(patient_id, source, n_inside,
            area_mm2       = round(area_mm2, 3),
            ihc_tumor_frac = round(tumor_over_inside, 3),
            dens_tumor_mm2 = round(dens_tumor_per_mm2, 1),
            path_frac      = round(path_frac, 3),
            abs_gap        = round(abs(tumor_over_inside - path_frac), 3)) |>
  arrange(desc(abs_gap)) |>
  knitr::kable(caption = "UNION concordance pairs (largest IHC-vs-pathologist gap first).")

cc_ann |>
  transmute(patient_id, annotation, source, n_inside,
            ihc_tumor_frac = round(tumor_over_inside, 3),
            path_frac      = round(path_frac, 3),
            abs_gap        = round(abs(tumor_over_inside - path_frac), 3)) |>
  arrange(desc(abs_gap)) |>
  knitr::kable(caption = "PER-ANNOTATION concordance pairs (largest gap first). Confirm each geojson _a<k> is the pathologist's ANNOTATION_k.")
```

**Concordance statistics.** `paired_cor3(x, y)` reports three correlations between
IHC and pathologist across the shared patients, each `NA` when `n < 3` or either
vector is constant:

```
Pearson r    = linear correlation
Spearman rho = rank correlation (monotonic, robust to outliers)
Kendall tau  = rank concordance (pairwise, best-behaved at small n)
```

```{r tumor-content-plot}
# `metric` selects the IHC normalization; `unit_scale` draws the x=y line and forces
# a 0..1 square (only meaningful for tumor_over_inside — tumor_over_cd45 is unbounded
# and pathologist is 0..1, so their scatter is judged by RANK, not the diagonal).
# Points are coloured by `patient_id` (from cc_ann / cc_union) so a single patient's
# annotations read as one colour — useful in the per-annotation panel where a patient
# contributes several points.
plot_concordance <- function(df, metric, subtitle, xlab, unit_scale = TRUE) {
  d  <- dplyr::filter(df, is.finite(.data[[metric]]), is.finite(path_frac))
  cc <- paired_cor3(d[[metric]], d$path_frac)   # Pearson + Spearman + Kendall
  p <- ggplot(d, aes(.data[[metric]], path_frac)) +
    geom_point(aes(colour = patient_id), alpha = 0.7, size = 2) +
    annotate("text", x = -Inf, y = Inf, hjust = -0.05, vjust = 1.2,
             label = sprintf("Pearson r = %.2f\nSpearman rho = %.2f\nKendall tau = %.2f\n(n = %d)",
                             cc$pearson, cc$spearman, cc$kendall, cc$n)) +
    labs(title = "Neoplastic cellularity: IHC vs pathologist", subtitle = subtitle,
         x = xlab, y = "Pathologist tumour fraction", colour = "Patient") +
    theme_classic(base_size = 12)
  if (unit_scale)
    p <- p + geom_abline(slope = 1, linetype = "dashed", colour = "red") +
             coord_equal(xlim = c(0, 1), ylim = c(0, 1))
  p
}

norm_specs <- list(
  list(metric = "tumor_over_inside",       xlab = "IHC tumour / all cells inside",  unit = TRUE),
  list(metric = "tumor_over_inside_clean", xlab = "IHC tumour / cells inside (excl. outliers + Unknown)", unit = TRUE),
  list(metric = "tumor_over_cd45",         xlab = "IHC tumour / CD45+ cells inside", unit = FALSE),
  list(metric = "dens_tumor_per_mm2",      xlab = "IHC tumour density (cells / mm²) inside", unit = FALSE)
)

for (s in norm_specs) {
  if (nrow(cc_ann) > 0)
    print(plot_concordance(cc_ann, s$metric, paste0(s$metric, " — per annotation"),
                           s$xlab, s$unit))
  if (nrow(cc_union) > 0)
    print(plot_concordance(cc_union, s$metric, paste0(s$metric, " — union"),
                           s$xlab, s$unit))
}
```

**Bland–Altman (union).** Per patient:

```
mean = (IHC + pathologist) / 2
diff = IHC - pathologist

bias   = average of diff over patients
LoA    = 1.96 * standard deviation of diff
limits = bias plus/minus LoA
```

Bland–Altman is strictly interpretable only when both quantities share the 0..1
scale (i.e. `tumor_over_inside`); for `tumor_over_cd45` it is shown for consistency,
but read it as descriptive since the CD45 ratio is unbounded.

```{r tumor-content-bland-altman}
plot_bland_altman <- function(df, metric, title) {
  d <- dplyr::filter(df, is.finite(.data[[metric]]), is.finite(path_frac))
  if (nrow(d) <= 2) return(invisible())
  ba <- d |> transmute(mean_val = (.data[[metric]] + path_frac) / 2,
                       diff_val = .data[[metric]] - path_frac)
  bias <- mean(ba$diff_val); loa <- 1.96 * sd(ba$diff_val)
  print(
    ggplot(ba, aes(mean_val, diff_val)) +
      geom_hline(yintercept = bias, colour = "blue") +
      geom_hline(yintercept = bias + c(-loa, loa), linetype = "dashed", colour = "grey40") +
      geom_point(alpha = 0.7, size = 2) +
      labs(title = title,
           subtitle = sprintf("bias = %.3f, limits of agreement +/- %.3f", bias, loa),
           x = "mean of the two", y = "IHC − pathologist") +
      theme_classic(base_size = 12)
  )
}

for (s in norm_specs)
  plot_bland_altman(cc_union, s$metric,
                    paste0("Bland–Altman (union): ", s$metric, " − pathologist"))
```

## Aggregation strategy — how collapsing the annotations moves the answer

The concordance panels above compare neoplastic cellularity at two fixed levels:
**per annotation** (one point per `ANNOTATION_k`) and **union** (one point per
patient). The union level hides a choice, because a patient with several annotations
has several IHC values *and* several pathologist values, and each side has to be
collapsed to one number before they can be paired. That collapse is not neutral.

**The mismatch this section exists to expose.** The union pair above collapses the
two sides by *different* rules:

```
pathologist   path_union      = mean(ANNOTATION_k / 100)          UNWEIGHTED mean
IHC           ihc_tumor_union = tumour cells / all cells,
                                over the POOLED cells of every
                                annotation of that patient        CELL-WEIGHTED mean
```

For a patient whose annotations differ in size these are different summaries of the
same slide. Take two annotations, 40 000 cells at 80% tumour and 4 000 cells at 50%:
the cell-weighted IHC value is 0.77 while the unweighted pathologist value is 0.65.
That 0.12 gap is **aggregation artefact**, not IHC-vs-pathologist disagreement — and
the existing union panels cannot tell the two apart.

So below, every IHC aggregator is paired against every pathologist aggregator:

```
IHC side (per patient)
  pooled      cells of every annotation pooled and deduped, ratio recomputed
              (exactly what the union panels above use)          cell-weighted
  wmean_ann   per-annotation fractions averaged, weights = that annotation's
              cell count                                          cell-weighted
  mean_ann    per-annotation fractions averaged, every annotation equal
  median_ann  median of the per-annotation fractions (only differs from mean_ann
              at 3+ annotations)

pathologist side (per patient)
  wmean_cells ANNOTATION_k scores averaged, weights = that annotation's IHC cell
              count — the pathologist analogue of pooling cells   cell-weighted
  mean        plain mean over the patient's annotations (used by the union panels)
  median      median over the annotations
  max         the highest-scoring annotation ("worst area governs")
  min         the lowest-scoring annotation (most conservative)
```

The weight is always the denominator of the metric being aggregated: `n_inside` for
`tumor_over_inside`, `n_inside_clean` for `tumor_over_inside_clean`. Weighting a
clean-set metric by the unclean count would over-weight annotations whose cells were
mostly discarded as outliers or Unknown.

One consequence to watch in the tables: a patient whose IHC region is labelled
`"csv"` rather than an `ANNOTATION_k` cannot be matched to a per-annotation
pathologist score, so it contributes no weight and drops out of the `wmean_cells`
rows while still appearing in the `mean` / `median` / `max` / `min` ones. That is why
`n` is printed on every row of the grid — compare rows only at equal `n`.

**Only two of the twenty combinations weight both sides the same way** —
`(pooled, wmean_cells)` and `(mean_ann, mean)`. Those are the pairs whose gap is
interpretable as genuine disagreement; everywhere else part of the gap is the
aggregation mismatch. They are shaded in the scatter grids and boxed in the heatmap.

**A built-in check.** `wmean_ann` should reproduce `pooled` almost exactly, since
pooling cells *is* cell-weighting them. The only thing separating those two rows is
the union's dedup of cells falling inside two overlapping annotations — so a visible
gap between them is a measurement of **annotation overlap**, not an aggregation
error. Read those two rows against each other before trusting anything else here.

```{r aggregation-grid}
# Every IHC aggregator x every pathologist aggregator, for both tumour-fraction
# normalizations. $pairs = the per-patient points; $stats = paired_cor3 + bias per
# combination. `bias` = mean(IHC - pathologist): the systematic offset that the
# correlation coefficients are deliberately blind to (a method can correlate
# perfectly and still read 15 points high).
agg      <- aggregation_grid(ihc_tumor_per_ann, ihc_tumor_union, path_long)
baseline <- per_annotation_baseline(ihc_tumor_per_ann, path_long)

agg$stats |>
  transmute(metric, ihc = ihc_lab, pathologist = path_lab,
            matched = if_else(matched, "yes", ""),
            n, pearson = round(pearson, 2), spearman = round(spearman, 2),
            kendall = round(kendall, 2), bias = round(bias, 3)) |>
  knitr::kable(caption = paste(
    "Aggregation grid, best Spearman first within each metric.",
    "matched = both sides weighted the same way (the interpretable rows).",
    "bias = mean(IHC - pathologist)."))

baseline |>
  transmute(metric, n, pearson = round(pearson, 2), spearman = round(spearman, 2),
            kendall = round(kendall, 2), bias = round(bias, 3)) |>
  knitr::kable(caption = paste(
    "PER-ANNOTATION baseline: no aggregation at all, one point per",
    "(patient, annotation). The reference the aggregated rows above should be read",
    "against — it neither gains aggregation's smoothing nor loses the",
    "within-patient signal."))
```

**Read the correlations with the `n` beside them.** This cohort gives a handful of
patients, so a Spearman rho here has a confidence interval spanning most of the
range, and the *ranking* of the twenty combinations is far less stable than the
numbers suggest. The point of the grid is not to pick the winner by rho — that would
be selecting on noise — but to see **whether the choice matters at all**. A flat grid
means the conclusion is robust to the aggregation; a grid with a large spread means
any single reported correlation is partly an artefact of the collapse rule.

```{r aggregation-heatmap, fig.width=8, fig.height=8}
# Reading a ROW shows how much the pathologist aggregator matters; a COLUMN, how much
# the IHC one does. Flat = the choice is immaterial for this cohort (the hoped-for
# outcome). Boxed cells are the matched-weighting pairs.
plot_aggregation_heatmap(agg$stats, "spearman")
```

```{r aggregation-scatter-raw, fig.width=11, fig.height=9}
# tumour / all cells inside. Dashed line is x = y; vertical distance from it is the
# disagreement. Shaded panels weight both sides the same way.
plot_aggregation_grid(agg$pairs, "tumor_over_inside")
```

```{r aggregation-scatter-clean, fig.width=11, fig.height=9}
# tumour / cells inside, with FlowPath Outlier == TRUE and Unknown-phenotype cells
# removed from both numerator and denominator.
plot_aggregation_grid(agg$pairs, "tumor_over_inside_clean")
```

Finally, which patients the choice actually bites on. A patient with one annotation
has zero spread by construction; the spread is the range a patient's value takes
across the aggregators, so large spreads mark the patients whose annotations
disagree most internally.

```{r aggregation-spread}
aggregation_spread(agg$pairs) |>
  transmute(metric, patient_id, n_pairs,
            ihc_min = round(ihc_min, 3), ihc_max = round(ihc_max, 3),
            ihc_spread = round(ihc_spread, 3),
            path_min = round(path_min, 3), path_max = round(path_max, 3),
            path_spread = round(path_spread, 3)) |>
  knitr::kable(caption = paste(
    "Per-patient spread across the aggregators (largest IHC spread first).",
    "Spread 0 = a single-annotation patient, where every aggregator coincides."))
```

### Recommendation

**For method validation, prefer the per-annotation level over any aggregation.** The
pathologist reports one score per annotation, so the annotation — not the patient —
is the natural unit at which the two methods can be compared like for like. It also
uses every observation instead of averaging most of them away, which matters at this
sample size. Aggregation is a compromise made necessary by the *clinical* joins
(survival, response, blood counts), which are per patient.

**When a patient-level number is required, use a matched pair.** Reporting a
cell-weighted IHC value against an unweighted pathologist value — what the union
panels currently do — mixes a real effect with an artefact. Of the two matched
options:

- `(mean_ann, mean)` — treats each annotation as one observation on both sides.
  Preferred as the primary: it mirrors how the pathologist's own numbers are
  produced (a per-region score, not a cell census), and one very large annotation
  cannot dominate the patient's value.
- `(pooled, wmean_cells)` — treats the patient's tissue as a single specimen and
  asks what fraction of its cells are neoplastic. Report alongside as the
  sensitivity analysis. It is the more natural quantity if the annotations are
  arbitrary subdivisions of one tumour rather than distinct regions.

If those two disagree materially, the disagreement is itself the finding: it means
the patient's annotations are heterogeneous in both size and tumour content, and no
single patient-level number represents that slide well.

## Immune infiltrate — IHC vs clinical immune phenotype

Per patient, over the **union** inside cells:

```
cd45_over_inside    = n_cd45_inside    / n_inside   (all CD45+ cells)
frac_<lineage>      = n_<lineage>      / n_inside   (lineage in CD8T, CD4T, Treg, NK)
cd3cd45_over_inside = n_cd3cd45_inside / n_inside   (marker CD3+CD45+ T cells)
gzmb_nk_over_inside = n_gzmb_nk_inside / n_inside   (GZMB+ NK cells)
```

`cd3cd45` (CD45+ AND CD3+) and `gzmb_nk` (NK lineage AND GZMB+) are **marker-gated**
subsets — defined by `<marker>_sign`, not the phenotype label — so they sit
alongside the phenotype lineages as extra readouts of overall immune (CD45+),
T-cell (CD3+CD45+) and cytotoxic-NK (GZMB+ NK) content.

Clinical references: `HOT score` (continuous) and `Immuno-phenotype` (categorical),
joined on `slide_key(`ID PATIENT`)`.

```{r infiltrate-join}
clin <- clinical_data |>
  mutate(patient_id = slide_key(`ID PATIENT`)) |>
  select(patient_id,
         hot_score  = any_of("HOT score"),
         immuno_phe = any_of("Immuno-phenotype"))

infiltrate <- ihc_tumor_union |>
  select(patient_id, cd45_over_inside, frac_CD8T, frac_CD4T, frac_Treg, frac_NK,
         cd3cd45_over_inside, gzmb_nk_over_inside) |>
  inner_join(clin, by = "patient_id")

cat("patients with IHC infiltrate + clinical immune phenotype:", nrow(infiltrate), "\n")
```

Statistic: Spearman correlation between `cd45_over_inside` and the numeric `HOT score`.

```{r infiltrate-hotscore}
if ("hot_score" %in% names(infiltrate) &&
    sum(is.finite(suppressWarnings(as.numeric(infiltrate$hot_score)))) > 2) {
  df <- infiltrate |> mutate(hot_score = suppressWarnings(as.numeric(hot_score)))
  st <- paired_spearman(df$cd45_over_inside, df$hot_score)
  ggplot(df, aes(cd45_over_inside, hot_score)) +
    geom_point(alpha = 0.7, size = 2) +
    geom_smooth(method = "lm", se = FALSE, colour = "steelblue", formula = y ~ x) +
    annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
             label = sprintf("Spearman rho = %.2f (p = %.3g, n = %d)",
                             st$rho, st$p, st$n)) +
    labs(title = "Immune infiltrate vs clinical HOT score",
         x = "IHC CD45+ fraction inside tumour region", y = "clinical HOT score") +
    theme_classic(base_size = 12)
}
```

Grouped by `Immuno-phenotype` (ordered cold -> intermediate -> hot); each box is
the distribution of the IHC fraction within a phenotype group. Metrics now include
the marker-gated CD45+, CD3+CD45+ and GZMB+ NK readouts.

```{r infiltrate-phenotype}
# The metrics carried into every immuno-phenotype comparison (all are 0..1
# fractions of the cells inside the tumour region, so the arcsin panel below can
# transform them all).
infiltrate_metrics <- c("cd45_over_inside", "cd3cd45_over_inside", "gzmb_nk_over_inside",
                        "frac_CD8T", "frac_CD4T", "frac_Treg", "frac_NK")

# One boxplot-by-phenotype figure; `transform` is applied to the fraction so the
# same code draws the raw and the arcsin-sqrt panels (see next chunk).
plot_infiltrate_phenotype <- function(infiltrate, ylab, transform = identity) {
  if (!("immuno_phe" %in% names(infiltrate)) ||
      dplyr::n_distinct(na.omit(infiltrate$immuno_phe)) <= 1) return(invisible())
  df <- infiltrate |>
    pivot_longer(all_of(infiltrate_metrics), names_to = "ihc_metric", values_to = "frac") |>
    filter(!is.na(immuno_phe), is.finite(frac)) |>
    mutate(immuno_phe = hotcold_order(immuno_phe), value = transform(frac)) |>
    filter(is.finite(value))
  ggplot(df, aes(immuno_phe, value, colour = immuno_phe)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.2, alpha = 0.6) +
    facet_wrap(~ ihc_metric, scales = "free_y") +
    scale_colour_manual(values = hotcold_cols(levels(df$immuno_phe)),
                        name = "Immuno-phenotype") +
    labs(title = "IHC infiltrate metrics by clinical immuno-phenotype",
         x = NULL, y = ylab, colour = "Immuno-phenotype") +
    theme_classic(base_size = 11) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
}

plot_infiltrate_phenotype(infiltrate, "fraction inside tumour region")
```

**Arcsin-transformed.** The same metrics under the arcsine square-root transform
(`asin(sqrt(fraction))`), which stabilises the variance of a proportion by
stretching the crowded 0/1 tails — read the boxes as the transformed scale, not raw
fractions.

```{r infiltrate-phenotype-arcsin}
plot_infiltrate_phenotype(infiltrate, "arcsin(sqrt(fraction))", transform = asin_sqrt)
```

### Composition under three denominators

Same population counts, three denominators (each `NA` when its denominator is 0).
Alongside the phenotype lineages, the marker-gated **CD45+**, **CD3+ CD45+** and
**GZMB+ NK** populations are included:

```
frac_inside = n_<population> / n_inside         (of all cells inside)
frac_tumor  = n_<population> / n_tumor_inside   (of tumour cells inside)
frac_cd45   = n_<population> / n_cd45_inside    (of CD45+ cells inside)

populations = CD8T, CD4T, Treg, NK, Immune_other        (phenotype lineages)
            + CD45+, CD3+ CD45+, GZMB+ NK               (marker-gated subsets)
```

Note `frac_cd45` for the CD45+ population is 1 by construction (a self-normalisation
check), and marker-gated subsets are shown under all three denominators for
consistency with the lineages.

```{r infiltrate-composition, fig.height=5}
# Marker-gated populations to add to the phenotype lineages (label -> count column
# from region_ratios). CD45+ = all immune; CD3+CD45+ = marker T cells; GZMB+ NK.
comp_markers <- c("CD45+"      = "n_cd45_inside",
                  "CD3+ CD45+" = "n_cd3cd45_inside",
                  "GZMB+ NK"   = "n_gzmb_nk_inside")

# One composition figure (population x denominator); `transform` applied to the
# fraction so the same code draws the raw and the arcsin-sqrt panels.
plot_composition <- function(region_metrics, ylab, caption, transform = identity) {
  comp <- region_composition(region_metrics, markers = comp_markers) |>
    pivot_longer(c(frac_inside, frac_tumor, frac_cd45),
                 names_to = "normalization", values_to = "frac") |>
    mutate(normalization = factor(
             c(frac_inside = "of all cells", frac_tumor = "of tumour cells",
               frac_cd45 = "of CD45+ cells")[normalization],
             levels = c("of all cells", "of tumour cells", "of CD45+ cells")),
           value = transform(frac)) |>
    filter(is.finite(value))
  if (nrow(comp) == 0) return(invisible())
  ggplot(comp, aes(lineage, value, fill = lineage)) +
    geom_boxplot(outlier.shape = NA, colour = "grey30") +
    geom_jitter(width = .15, alpha = .5, size = 1) +
    facet_wrap(~ normalization, scales = "free_y") +
    scale_fill_manual(values = oi, guide = "none") +
    labs(title = "IHC immune composition inside the tumour region",
         x = NULL, y = ylab, caption = caption) +
    theme_classic(base_size = 11) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
}

plot_composition(ihc_tumor_union, "fraction",
                 "denominators: all cells / tumour cells / CD45+ cells inside")
```

**Arcsin-transformed.** Same composition under the arcsine square-root transform
of each fraction.

```{r infiltrate-composition-arcsin, fig.height=5}
plot_composition(ihc_tumor_union, "arcsin(sqrt(fraction))",
                 "arcsin-sqrt of: all cells / tumour cells / CD45+ cells inside",
                 transform = asin_sqrt)
```

## Biological-validity QC — in-annotation IHC immune readout vs clinical table

These use the **per-patient union of in-annotation cells** (`union_cells`: the flag-
inside cells pooled across a patient's annotation files, deduped one row per cell) —
the in-annotation analogue of the original's whole-slide readout. For a fallback
patient there is only one region, so its "union" is simply the flag-inside cells of
the whole-slide csv. Per patient p, over that patient's in-annotation cells:

```
lineage fraction:  frac(p, lineage) = (cells of p with that lineage) / (all cells of p)
                                       [ihc_lineage_fraction]

marker pos-fraction: <marker>_posfrac(p) =
     (cells of p positive for <marker>) / (cells of p with a non-NA <marker>_sign)
                                       [ihc_marker_fraction; na.rm]
```

```{r whole-slide-immune}
# In-annotation analogue of whole-slide ihc_data: per-patient union of inside cells.
union_cells <- union_inside_cells(ann_cells)

ihc_lin <- ihc_lineage_fraction(union_cells) |>
  filter(lineage %in% comparable_lineages) |>
  select(patient_id, lineage, frac) |>
  pivot_wider(names_from = lineage, values_from = frac, values_fill = 0)

ihc_mk <- ihc_marker_fraction(union_cells) |>
  select(patient_id, any_of(c("CD45_posfrac", "CD8_posfrac", "FOXP3_posfrac")))

ihc_immune <- full_join(ihc_lin, ihc_mk, by = "patient_id")

clin_full <- clinical_data |>
  mutate(patient_id = slide_key(`ID PATIENT`))

ihc_clin <- ihc_immune |> inner_join(clin_full, by = "patient_id")
cat("patients with whole-slide IHC immune metrics + clinical row:", nrow(ihc_clin), "\n")

# ihc_immune_score = CD45_posfrac if present, else the CD8T lineage fraction.
score_col <- intersect(c("CD45_posfrac", "CD8T"), names(ihc_clin))[1]
ihc_clin$ihc_immune_score <- if (!is.na(score_col)) ihc_clin[[score_col]] else NA_real_
```

### Tissue vs peripheral-blood immunity

```
NLR(p) = BASAL VALUE OF NEUTROPHILS / BASAL VALUE OF LYMPHOCYTES
```

compared against `ihc_immune_score` by Spearman correlation (also for lymphocytes
and PNI).

```{r blood-immunity}
blood <- ihc_clin |>
  transmute(
    patient_id, ihc_immune_score,
    lymphocytes = suppressWarnings(as.numeric(`BASAL VALUE OF LYMPHOCYTES`)),
    neutrophils = suppressWarnings(as.numeric(`BASAL VALUE OF NEUTROPHILS`)),
    pni         = suppressWarnings(as.numeric(`PROGNOSTIC NUTRITIONAL INDEX`))
  ) |>
  mutate(nlr = neutrophils / lymphocytes)

blood_long <- blood |>
  pivot_longer(c(lymphocytes, nlr, pni), names_to = "blood_metric", values_to = "value") |>
  filter(is.finite(ihc_immune_score), is.finite(value))

if (nrow(blood_long) > 2) {
  labs_tbl <- blood_long |>
    group_by(blood_metric) |>
    group_modify(~ paired_spearman(.x$ihc_immune_score, .x$value)) |>
    mutate(lab = sprintf("rho = %.2f (p = %.3g, n = %d)", rho, p, n))
  ggplot(blood_long, aes(ihc_immune_score, value)) +
    geom_point(alpha = 0.7, size = 2) +
    geom_smooth(method = "lm", se = FALSE, colour = "steelblue", formula = y ~ x) +
    geom_text(data = labs_tbl, aes(x = -Inf, y = Inf, label = lab),
              hjust = -0.05, vjust = 1.4, inherit.aes = FALSE, size = 3) +
    facet_wrap(~ blood_metric, scales = "free_y") +
    labs(title = "IHC immune score vs peripheral-blood immunity",
         x = "IHC immune score (CD45+ fraction)", y = NULL) +
    theme_classic(base_size = 11)
}
```

### Induction-treatment response

`ihc_immune_score` distribution grouped by `RESPONSE` (factor ordered by ascending
class value, 1 -> 3); no summary statistic beyond the group boxplots.

```{r response}
if ("RESPONSE" %in% names(ihc_clin)) {
  df <- ihc_clin |>
    filter(!is.na(RESPONSE), is.finite(ihc_immune_score))
  df$RESPONSE <- factor(as.character(df$RESPONSE),
                        levels = as.character(sort(unique(as.numeric(as.character(df$RESPONSE))))))
  if (dplyr::n_distinct(df$RESPONSE) > 1 && nrow(df) > 3) {
    ggplot(df, aes(RESPONSE, ihc_immune_score, colour = RESPONSE)) +
      geom_boxplot(outlier.shape = NA) +
      geom_jitter(width = 0.2, alpha = 0.6) +
      labs(title = "IHC immune score by induction response",
           x = NULL, y = "IHC immune score (CD45+ fraction)") +
      theme_classic(base_size = 12) +
      theme(axis.text.x = element_text(angle = 30, hjust = 1), legend.position = "none")
  }
}
```

### Survival

```
group(p) = "high" if ihc_immune_score(p) >= median, else "low"   -> Kaplan-Meier
HR       = exp( Cox coefficient of ihc_immune_score )            (univariable coxph)
```

Event coding (`coerce_event`): yes/dead/relapse/1 -> 1; no/alive/censored/0 -> 0.
Time = `OS`/`PFS` (months); event = `FOLLOW UP` (death) / `PERSISTENCE OR RELAPSE?`
(progression).

```{r survival}
surv_cfg <- list(
  OS  = list(time = "OS",  event = "FOLLOW UP"),
  PFS = list(time = "PFS", event = "PERSISTENCE OR RELAPSE?")
)

coerce_event <- function(x) {
  if (is.numeric(x)) return(as.integer(x > 0))
  v <- tolower(trimws(as.character(x)))
  dplyr::case_when(
    v %in% c("1","yes","y","dead","death","deceased","event","relapse",
             "progression","progressed","true") ~ 1L,
    v %in% c("0","no","n","alive","ned","censored","false") ~ 0L,
    TRUE ~ NA_integer_
  )
}

run_survival <- function(cfg, label) {
  if (!requireNamespace("survival", quietly = TRUE)) {
    message("survival package not installed — skipping ", label); return(invisible())
  }
  if (!all(c(cfg$time, cfg$event) %in% names(ihc_clin))) {
    message("missing ", cfg$time, "/", cfg$event, " — skipping ", label); return(invisible())
  }
  d <- ihc_clin |>
    transmute(ihc_immune_score,
              .time  = suppressWarnings(as.numeric(.data[[cfg$time]])),
              .event = coerce_event(.data[[cfg$event]])) |>
    filter(is.finite(.time), !is.na(.event), is.finite(ihc_immune_score))
  if (nrow(d) < 5 || dplyr::n_distinct(d$.event) < 2) {
    message("too few informative rows for ", label, " (n = ", nrow(d), ")"); return(invisible())
  }
  d <- d |> mutate(immune_group = if_else(ihc_immune_score >= median(ihc_immune_score),
                                          "IHC-immune high", "IHC-immune low"))
  fit  <- survival::survfit(survival::Surv(.time, .event) ~ immune_group, data = d)
  cox  <- survival::coxph(survival::Surv(.time, .event) ~ ihc_immune_score, data = d)
  hr   <- exp(stats::coef(cox))[1]; pcox <- summary(cox)$coefficients[1, "Pr(>|z|)"]
  cat(sprintf("%s: Cox HR per unit IHC immune score = %.2f (p = %.3g, n = %d)\n",
              label, hr, pcox, nrow(d)))
  if (requireNamespace("survminer", quietly = TRUE)) {
    print(survminer::ggsurvplot(fit, data = d, pval = TRUE, risk.table = TRUE,
                                title = paste(label, "by IHC immune score (median split)"),
                                legend.title = "")$plot)
  } else {
    plot(fit, col = c("firebrick", "steelblue"), xlab = "time", ylab = "survival",
         main = paste(label, "by IHC immune score"))
    legend("topright", levels(factor(d$immune_group)),
           col = c("firebrick", "steelblue"), lty = 1, bty = "n")
  }
}

invisible(purrr::iwalk(surv_cfg, ~ run_survival(.x, .y)))
```

## Notes

- Membership is the FlowPath **`Out_of_annotation` flag** everywhere
  (`inside = flag FALSE`); no geojson, no sf point-in-polygon. `area_mm2` /
  `dens_*_per_mm2` are `NA` throughout (no polygon).
- `source` is `flag` for patients read from a per-annotation `<patient>_a<k>.csv`
  and `csv` for fallback patients read from the whole-slide `<patient>.csv`. Check
  the inventory table above before reading any panel: a `csv` row is one region for
  the whole slide, not a single pathologist annotation.
- The `old/` subfolder **overrides** the top-level file for every `(patient,
  annotation)` it provides — see the `file_origin == "old"` rows of the inventory
  table for exactly which. The companion `clinical_data_per_annotation.Rmd` is
  identical except that it ignores `old/`, so differences between the two reports
  are attributable to those re-exported annotations alone.
- A fallback patient has ONE region, not a per-annotation breakdown. It is labelled
  with the patient's sole scored annotation when it has exactly one — so all six
  patients reach every panel — and `"csv"` (union panels only, with a warning) when
  it has more. Read a fallback patient's per-annotation point as "the whole slide",
  which for a single-annotation patient is the annotation.
- In-annotation marker fractions use `na.rm`, so the real denominator is the count
  of in-annotation cells with a non-`NA` sign, not every in-annotation cell.
- Cohorts are small; every correlation is Spearman with its `n` reported. Read
  associations as descriptive.

```{r export-pdfs, include=FALSE}
# Copy every plot from this analysis (each rendered as a PDF via the
# dev = c("png", "pdf") chunk option) into output/figures/clinical_data_per_annotation_old/.
# Runs last so that all figures from the chunks above already exist on disk.
export_pdf_figures("clinical_data_per_annotation_old")
```
