---
title: "clinical_data"
author: "sceriff0"
date: "2026-07-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(sf)
})

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6,
                      dev = c("png", "pdf"))   # PNG for the site, PDF exported to output/

# Loaders (dds, clinical_data, neoplastic_data, counts_data, ihc_data) and the
# shared validation helpers. test.R prints colnames() on load; silence it.
invisible(capture.output(suppressMessages(source(here("code", "test.R")))))
source(here("code", "validation_helpers.R"))
source(here("code", "pdf_export.R"))          # export_pdf_figures(): plots -> output/figures/
```

## Starting data and notation

Every value below is derived from three inputs:

- `ihc_data` — FlowPath single cells, **one csv per patient** (`<patient>.csv`), one
  row per cell. Carries `phenotype`, per-marker `<marker>_sign`, `centroid_x`/`y`,
  `patient_id`, and (optionally) an `Out_of_annotation` flag used only as a
  membership fallback.
- `neoplastic_data` — pathologist tumour-content percentage per annotation,
  `ANNOTATION_1..3`.
- `clinical_data` — the clinical CRF (`HOT score`, `Immuno-phenotype`, blood counts,
  `RESPONSE`, `OS`/`PFS`).

Membership (which cells count as "inside the tumour annotation"):

```
PRIMARY (patient HAS a geojson):
  cell is INSIDE  = its centroid falls in the pathologist geojson polygon
                    (sf point-in-polygon), with centroids mapped to the geojson
                    pixel frame by  x_px = centroid_x / 0.325  (0.325 microns/pixel)
  per annotation  = inside one polygon (ANNOTATION_k)
  union           = inside the dissolved union of the patient's polygons

FALLBACK (patient has NO geojson):
  cell is INSIDE = Out_of_annotation is FALSE   (a single flag, one row labelled "csv")
```

The `source` column on every row records which was used (`sf` or `csv`).

Cell labels:

```
tumour cell = phenotype_clean contains "Tumor"
              (phenotype_clean = the text inside the parentheses of `phenotype`)
CD45+ cell  = CD45_sign is "+"  (is_pos accepts +, pos, positive, yes, true, 1)
lineage(cell) = phenotype_clean collapsed via the phenotype_lineage table
```

`slide_key()` maps every id to the shared numeric slide space (strip non-digits,
apply the verified typo fix `15879` -> `15897`).

## Membership counts

For a set of cells (either one annotation's inside cells or the union inside
cells), `region_ratios()` returns:

```
n_inside          = number of cells in the set
n_inside_clean    = cells in the set after dropping Outlier == TRUE and
                    Unknown-phenotype cells (QC-filtered denominator)
n_tumor_inside    = tumour cells in the set
n_tumor_clean     = tumour cells in the set that survive the same clean filter
n_cd45_inside     = CD45+ cells in the set
n_cd3cd45_inside  = CD45+ AND CD3+ cells (marker-gated T cells)
n_gzmb_nk_inside  = NK-lineage cells that are GZMB+ (cytotoxic/activated NK)
n_<lineage>       = cells in the set with that lineage
                    (lineage in: CD8T, CD4T, Treg, NK, Immune_other, Stroma)

tumor_over_inside       = n_tumor_inside / n_inside
tumor_over_inside_clean = n_tumor_clean  / n_inside_clean  (outliers + Unknown removed)
cd45_over_inside    = n_cd45_inside    / n_inside
cd3cd45_over_inside = n_cd3cd45_inside / n_inside
cd3cd45_over_cd45   = n_cd3cd45_inside / n_cd45_inside   (T cells among immune)
gzmb_nk_over_inside = n_gzmb_nk_inside / n_inside
gzmb_nk_over_nk     = n_gzmb_nk_inside / n_NK            (GZMB+ fraction of NK)
frac_<lineage>      = n_<lineage>      / n_inside
```

Any fraction with a zero denominator is `NA`. The marker-gated subsets use
`<marker>_sign` (`is_pos`), not the phenotype label.

```{r annotations}
# Pathologist geojson polygons (primary membership source).
annots <- load_annotations(patient_ids = neoplastic_data$SAMPLE)

neo_ids      <- slide_key(neoplastic_data$SAMPLE)
ihc_ids      <- unique(slide_key(ihc_data$patient_id))
have_geojson <- unique(slide_key(annots$patient_id))
missing      <- setdiff(neo_ids, ihc_ids)            # no FlowPath csv at all
csv_only     <- setdiff(intersect(neo_ids, ihc_ids), have_geojson)  # cells but no geojson

cat("FlowPath patients loaded:", length(ihc_ids),
    "[", paste(sort(ihc_ids), collapse = ", "), "]\n")
cat("patients with a pathologist score AND IHC cells:",
    length(intersect(neo_ids, ihc_ids)),
    "| via sf geojson:", length(intersect(intersect(neo_ids, ihc_ids), have_geojson)),
    "| via CSV fallback:", length(csv_only), "\n")
if (length(missing))
  cat("pathologist patients with NO FlowPath csv (absent from every IHC metric):",
      paste(sort(missing), collapse = ", "), "\n")
```

## Tumour content — IHC vs pathologist

The IHC tumour metric is shown under **three normalizations**, all restricted to the
cells inside the annotation:

```
tumor_over_inside       = (tumour cells inside) / (all cells inside)     -> 0..1
tumor_over_inside_clean = (tumour cells inside) / (cells inside)  -> 0..1
                          with the cell set first cleaned: FlowPath Outlier == TRUE
                          cells and Unknown-phenotype cells dropped from BOTH the
                          numerator and the denominator (a QC-filtered version of
                          tumor_over_inside; if a csv has no `Outlier` column no cell
                          is treated as an outlier)
tumor_over_cd45         = (tumour cells inside) / (CD45+ cells inside)   -> can exceed 1
dens_tumor_per_mm2      = (tumour cells inside) / (annotation area in mm^2)  [geojson patients only]
                          area_mm2 = polygon pixel area * 0.325^2 / 1e6

pathologist (per annotation) = ANNOTATION_k / 100
pathologist (union)          = mean of that patient's ANNOTATION_k / 100 values
```

`tumor_over_inside_clean` answers the same IHC-vs-pathologist question as
`tumor_over_inside` but removes debris/segmentation outliers and unclassified cells
that dilute the denominator, so it is expected to track the pathologist's tumour
fraction more closely.

`dens_tumor_per_mm2` is the same tumour-cell density (cells/mm²) that periphery
reports, computed here inside the whole annotation. It needs the polygon area, so it
is only defined for patients with a geojson (`NA` for the CSV-fallback ones), and —
being a density, not a 0..1 fraction — it is compared to the pathologist by **rank**
(Spearman/Kendall), not the x=y line.

Each normalization is computed both per-annotation (one polygon) and over the union
of a patient's polygons.

```{r tumor-content}
# PRIMARY = sf point-in-polygon on the geojson (fixed /0.325); CSV Out_of_annotation
# flag only for patients with no geojson.
ihc_tumor_per_ann <- ihc_annotation_metrics(ihc_data, annots, scope = "per_annotation")
ihc_tumor_union   <- ihc_annotation_metrics(ihc_data, annots, scope = "union")

path_long <- neoplastic_data |>
  mutate(patient_id = slide_key(SAMPLE)) |>
  pivot_longer(starts_with("ANNOTATION_"),
               names_to = "annotation", values_to = "path_pct") |>
  filter(!is.na(path_pct)) |>
  mutate(path_frac = path_pct / 100)

cc_ann <- ihc_tumor_per_ann |>
  inner_join(path_long, by = c("patient_id", "annotation")) |>
  filter(!is.na(tumor_over_inside), !is.na(path_frac))

path_union <- path_long |>
  group_by(patient_id) |>
  summarise(path_frac = mean(path_frac), .groups = "drop")
cc_union <- ihc_tumor_union |>
  inner_join(path_union, by = "patient_id") |>
  filter(!is.na(tumor_over_inside), !is.na(path_frac))

cat("per-annotation pairs:", nrow(cc_ann), " | union pairs:", nrow(cc_union), "\n")

# Diagnostic: the ACTUAL pairs that enter the correlation, largest gap first. Read
# the top rows to tell WHY the correlation is what it is:
#   - big gap + source == "csv" or tiny n_inside  -> suspect membership, not biology
#   - big gap + healthy sf patient                -> genuine IHC/pathologist disagreement
#   - per-annotation only: verify geojson _a<k> lines up with pathologist ANNOTATION_k
cc_union |>
  transmute(patient_id, source, n_inside,
            area_mm2       = round(area_mm2, 3),
            ihc_tumor_frac = round(tumor_over_inside, 3),
            dens_tumor_mm2 = round(dens_tumor_per_mm2, 1),
            path_frac      = round(path_frac, 3),
            abs_gap        = round(abs(tumor_over_inside - path_frac), 3)) |>
  arrange(desc(abs_gap)) |>
  knitr::kable(caption = "UNION concordance pairs (largest IHC-vs-pathologist gap first).")

cc_ann |>
  transmute(patient_id, annotation, source, n_inside,
            ihc_tumor_frac = round(tumor_over_inside, 3),
            path_frac      = round(path_frac, 3),
            abs_gap        = round(abs(tumor_over_inside - path_frac), 3)) |>
  arrange(desc(abs_gap)) |>
  knitr::kable(caption = "PER-ANNOTATION concordance pairs (largest gap first). Confirm each geojson _a<k> is the pathologist's ANNOTATION_k.")
```

**Concordance statistics.** `paired_cor3(x, y)` reports three correlations between
IHC and pathologist across the shared patients, each `NA` when `n < 3` or either
vector is constant:

```
Pearson r    = linear correlation
Spearman rho = rank correlation (monotonic, robust to outliers)
Kendall tau  = rank concordance (pairwise, best-behaved at small n)
```

```{r tumor-content-plot}
# `metric` selects the IHC normalization; `unit_scale` draws the x=y line and forces
# a 0..1 square (only meaningful for tumor_over_inside — tumor_over_cd45 is unbounded
# and pathologist is 0..1, so their scatter is judged by RANK, not the diagonal).
# Points are coloured by `patient_id` (from cc_ann / cc_union) so a single patient's
# annotations read as one colour — useful in the per-annotation panel where a patient
# contributes several points.
plot_concordance <- function(df, metric, subtitle, xlab, unit_scale = TRUE) {
  d  <- dplyr::filter(df, is.finite(.data[[metric]]), is.finite(path_frac))
  cc <- paired_cor3(d[[metric]], d$path_frac)   # Pearson + Spearman + Kendall
  p <- ggplot(d, aes(.data[[metric]], path_frac)) +
    geom_point(aes(colour = patient_id), alpha = 0.7, size = 2) +
    annotate("text", x = -Inf, y = Inf, hjust = -0.05, vjust = 1.2,
             label = sprintf("Pearson r = %.2f\nSpearman rho = %.2f\nKendall tau = %.2f\n(n = %d)",
                             cc$pearson, cc$spearman, cc$kendall, cc$n)) +
    labs(title = "Neoplastic cellularity: IHC vs pathologist", subtitle = subtitle,
         x = xlab, y = "Pathologist tumour fraction", colour = "Patient") +
    theme_classic(base_size = 12)
  if (unit_scale)
    p <- p + geom_abline(slope = 1, linetype = "dashed", colour = "red") +
             coord_equal(xlim = c(0, 1), ylim = c(0, 1))
  p
}

norm_specs <- list(
  list(metric = "tumor_over_inside",       xlab = "IHC tumour / all cells inside",  unit = TRUE),
  list(metric = "tumor_over_inside_clean", xlab = "IHC tumour / cells inside (excl. outliers + Unknown)", unit = TRUE),
  list(metric = "tumor_over_cd45",         xlab = "IHC tumour / CD45+ cells inside", unit = FALSE),
  list(metric = "dens_tumor_per_mm2",      xlab = "IHC tumour density (cells / mm²) inside", unit = FALSE)
)

for (s in norm_specs) {
  if (nrow(cc_ann) > 0)
    print(plot_concordance(cc_ann, s$metric, paste0(s$metric, " — per annotation"),
                           s$xlab, s$unit))
  if (nrow(cc_union) > 0)
    print(plot_concordance(cc_union, s$metric, paste0(s$metric, " — union"),
                           s$xlab, s$unit))
}
```

**Bland–Altman (union).** Per patient:

```
mean = (IHC + pathologist) / 2
diff = IHC - pathologist

bias   = average of diff over patients
LoA    = 1.96 * standard deviation of diff
limits = bias plus/minus LoA
```

Bland–Altman is strictly interpretable only when both quantities share the 0..1
scale (i.e. `tumor_over_inside`); for `tumor_over_cd45` it is shown for consistency,
but read it as descriptive since the CD45 ratio is unbounded.

```{r tumor-content-bland-altman}
plot_bland_altman <- function(df, metric, title) {
  d <- dplyr::filter(df, is.finite(.data[[metric]]), is.finite(path_frac))
  if (nrow(d) <= 2) return(invisible())
  ba <- d |> transmute(mean_val = (.data[[metric]] + path_frac) / 2,
                       diff_val = .data[[metric]] - path_frac)
  bias <- mean(ba$diff_val); loa <- 1.96 * sd(ba$diff_val)
  print(
    ggplot(ba, aes(mean_val, diff_val)) +
      geom_hline(yintercept = bias, colour = "blue") +
      geom_hline(yintercept = bias + c(-loa, loa), linetype = "dashed", colour = "grey40") +
      geom_point(alpha = 0.7, size = 2) +
      labs(title = title,
           subtitle = sprintf("bias = %.3f, limits of agreement +/- %.3f", bias, loa),
           x = "mean of the two", y = "IHC − pathologist") +
      theme_classic(base_size = 12)
  )
}

for (s in norm_specs)
  plot_bland_altman(cc_union, s$metric,
                    paste0("Bland–Altman (union): ", s$metric, " − pathologist"))
```

## Immune infiltrate — IHC vs clinical immune phenotype

Per patient, over the **union** inside cells:

```
cd45_over_inside    = n_cd45_inside    / n_inside   (all CD45+ cells)
frac_<lineage>      = n_<lineage>      / n_inside   (lineage in CD8T, CD4T, Treg, NK)
cd3cd45_over_inside = n_cd3cd45_inside / n_inside   (marker CD3+CD45+ T cells)
gzmb_nk_over_inside = n_gzmb_nk_inside / n_inside   (GZMB+ NK cells)
```

`cd3cd45` (CD45+ AND CD3+) and `gzmb_nk` (NK lineage AND GZMB+) are **marker-gated**
subsets — defined by `<marker>_sign`, not the phenotype label — so they sit
alongside the phenotype lineages as extra readouts of overall immune (CD45+),
T-cell (CD3+CD45+) and cytotoxic-NK (GZMB+ NK) content.

Clinical references: `HOT score` (continuous) and `Immuno-phenotype` (categorical),
joined on `slide_key(`ID PATIENT`)`.

```{r infiltrate-join}
clin <- clinical_data |>
  mutate(patient_id = slide_key(`ID PATIENT`)) |>
  select(patient_id,
         hot_score  = any_of("HOT score"),
         immuno_phe = any_of("Immuno-phenotype"))

infiltrate <- ihc_tumor_union |>
  select(patient_id, cd45_over_inside, frac_CD8T, frac_CD4T, frac_Treg, frac_NK,
         cd3cd45_over_inside, gzmb_nk_over_inside) |>
  inner_join(clin, by = "patient_id")

cat("patients with IHC infiltrate + clinical immune phenotype:", nrow(infiltrate), "\n")
```

Statistic: Spearman correlation between `cd45_over_inside` and the numeric `HOT score`.

```{r infiltrate-hotscore}
if ("hot_score" %in% names(infiltrate) &&
    sum(is.finite(suppressWarnings(as.numeric(infiltrate$hot_score)))) > 2) {
  df <- infiltrate |> mutate(hot_score = suppressWarnings(as.numeric(hot_score)))
  st <- paired_spearman(df$cd45_over_inside, df$hot_score)
  ggplot(df, aes(cd45_over_inside, hot_score)) +
    geom_point(alpha = 0.7, size = 2) +
    geom_smooth(method = "lm", se = FALSE, colour = "steelblue", formula = y ~ x) +
    annotate("text", x = -Inf, y = Inf, hjust = -0.1, vjust = 1.5,
             label = sprintf("Spearman rho = %.2f (p = %.3g, n = %d)",
                             st$rho, st$p, st$n)) +
    labs(title = "Immune infiltrate vs clinical HOT score",
         x = "IHC CD45+ fraction inside tumour region", y = "clinical HOT score") +
    theme_classic(base_size = 12)
}
```

Grouped by `Immuno-phenotype` (ordered cold -> intermediate -> hot); each box is
the distribution of the IHC fraction within a phenotype group. Metrics now include
the marker-gated CD45+, CD3+CD45+ and GZMB+ NK readouts.

```{r infiltrate-phenotype}
# The metrics carried into every immuno-phenotype comparison (all are 0..1
# fractions of the cells inside the tumour region, so the arcsin panel below can
# transform them all).
infiltrate_metrics <- c("cd45_over_inside", "cd3cd45_over_inside", "gzmb_nk_over_inside",
                        "frac_CD8T", "frac_CD4T", "frac_Treg", "frac_NK")

# One boxplot-by-phenotype figure; `transform` is applied to the fraction so the
# same code draws the raw and the arcsin-sqrt panels (see next chunk).
plot_infiltrate_phenotype <- function(infiltrate, ylab, transform = identity) {
  if (!("immuno_phe" %in% names(infiltrate)) ||
      dplyr::n_distinct(na.omit(infiltrate$immuno_phe)) <= 1) return(invisible())
  df <- infiltrate |>
    pivot_longer(all_of(infiltrate_metrics), names_to = "ihc_metric", values_to = "frac") |>
    filter(!is.na(immuno_phe), is.finite(frac)) |>
    mutate(immuno_phe = hotcold_order(immuno_phe), value = transform(frac)) |>
    filter(is.finite(value))
  ggplot(df, aes(immuno_phe, value, colour = immuno_phe)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.2, alpha = 0.6) +
    facet_wrap(~ ihc_metric, scales = "free_y") +
    scale_colour_manual(values = hotcold_cols(levels(df$immuno_phe)),
                        name = "Immuno-phenotype") +
    labs(title = "IHC infiltrate metrics by clinical immuno-phenotype",
         x = NULL, y = ylab, colour = "Immuno-phenotype") +
    theme_classic(base_size = 11) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
}

plot_infiltrate_phenotype(infiltrate, "fraction inside tumour region")
```

**Arcsin-transformed.** The same metrics under the arcsine square-root transform
(`asin(sqrt(fraction))`), which stabilises the variance of a proportion by
stretching the crowded 0/1 tails — read the boxes as the transformed scale, not raw
fractions.

```{r infiltrate-phenotype-arcsin}
plot_infiltrate_phenotype(infiltrate, "arcsin(sqrt(fraction))", transform = asin_sqrt)
```

### Composition under three denominators

Same population counts, three denominators (each `NA` when its denominator is 0).
Alongside the phenotype lineages, the marker-gated **CD45+**, **CD3+ CD45+** and
**GZMB+ NK** populations are included:

```
frac_inside = n_<population> / n_inside         (of all cells inside)
frac_tumor  = n_<population> / n_tumor_inside   (of tumour cells inside)
frac_cd45   = n_<population> / n_cd45_inside    (of CD45+ cells inside)

populations = CD8T, CD4T, Treg, NK, Immune_other        (phenotype lineages)
            + CD45+, CD3+ CD45+, GZMB+ NK               (marker-gated subsets)
```

Note `frac_cd45` for the CD45+ population is 1 by construction (a self-normalisation
check), and marker-gated subsets are shown under all three denominators for
consistency with the lineages.

```{r infiltrate-composition, fig.height=5}
# Marker-gated populations to add to the phenotype lineages (label -> count column
# from region_ratios). CD45+ = all immune; CD3+CD45+ = marker T cells; GZMB+ NK.
comp_markers <- c("CD45+"      = "n_cd45_inside",
                  "CD3+ CD45+" = "n_cd3cd45_inside",
                  "GZMB+ NK"   = "n_gzmb_nk_inside")

# One composition figure (population x denominator); `transform` applied to the
# fraction so the same code draws the raw and the arcsin-sqrt panels.
plot_composition <- function(region_metrics, ylab, caption, transform = identity) {
  comp <- region_composition(region_metrics, markers = comp_markers) |>
    pivot_longer(c(frac_inside, frac_tumor, frac_cd45),
                 names_to = "normalization", values_to = "frac") |>
    mutate(normalization = factor(
             c(frac_inside = "of all cells", frac_tumor = "of tumour cells",
               frac_cd45 = "of CD45+ cells")[normalization],
             levels = c("of all cells", "of tumour cells", "of CD45+ cells")),
           value = transform(frac)) |>
    filter(is.finite(value))
  if (nrow(comp) == 0) return(invisible())
  ggplot(comp, aes(lineage, value, fill = lineage)) +
    geom_boxplot(outlier.shape = NA, colour = "grey30") +
    geom_jitter(width = .15, alpha = .5, size = 1) +
    facet_wrap(~ normalization, scales = "free_y") +
    scale_fill_manual(values = oi, guide = "none") +
    labs(title = "IHC immune composition inside the tumour region",
         x = NULL, y = ylab, caption = caption) +
    theme_classic(base_size = 11) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
}

plot_composition(ihc_tumor_union, "fraction",
                 "denominators: all cells / tumour cells / CD45+ cells inside")
```

**Arcsin-transformed.** Same composition under the arcsine square-root transform
of each fraction.

```{r infiltrate-composition-arcsin, fig.height=5}
plot_composition(ihc_tumor_union, "arcsin(sqrt(fraction))",
                 "arcsin-sqrt of: all cells / tumour cells / CD45+ cells inside",
                 transform = asin_sqrt)
```

## Biological-validity QC — whole-slide IHC immune readout vs clinical table

These use `ihc_data` (whole slide, every imaged patient — no annotation). Per
patient p, over all of that patient's cells:

```
lineage fraction:  frac(p, lineage) = (cells of p with that lineage) / (all cells of p)
                                       [ihc_lineage_fraction]

marker pos-fraction: <marker>_posfrac(p) =
     (cells of p positive for <marker>) / (cells of p with a non-NA <marker>_sign)
                                       [ihc_marker_fraction; na.rm]
```

```{r whole-slide-immune}
ihc_lin <- ihc_lineage_fraction(ihc_data) |>
  filter(lineage %in% comparable_lineages) |>
  select(patient_id, lineage, frac) |>
  pivot_wider(names_from = lineage, values_from = frac, values_fill = 0)

ihc_mk <- ihc_marker_fraction(ihc_data) |>
  select(patient_id, any_of(c("CD45_posfrac", "CD8_posfrac", "FOXP3_posfrac")))

ihc_immune <- full_join(ihc_lin, ihc_mk, by = "patient_id")

clin_full <- clinical_data |>
  mutate(patient_id = slide_key(`ID PATIENT`))

ihc_clin <- ihc_immune |> inner_join(clin_full, by = "patient_id")
cat("patients with whole-slide IHC immune metrics + clinical row:", nrow(ihc_clin), "\n")

# ihc_immune_score = CD45_posfrac if present, else the CD8T lineage fraction.
score_col <- intersect(c("CD45_posfrac", "CD8T"), names(ihc_clin))[1]
ihc_clin$ihc_immune_score <- if (!is.na(score_col)) ihc_clin[[score_col]] else NA_real_
```

### Tissue vs peripheral-blood immunity

```
NLR(p) = BASAL VALUE OF NEUTROPHILS / BASAL VALUE OF LYMPHOCYTES
```

compared against `ihc_immune_score` by Spearman correlation (also for lymphocytes
and PNI).

```{r blood-immunity}
blood <- ihc_clin |>
  transmute(
    patient_id, ihc_immune_score,
    lymphocytes = suppressWarnings(as.numeric(`BASAL VALUE OF LYMPHOCYTES`)),
    neutrophils = suppressWarnings(as.numeric(`BASAL VALUE OF NEUTROPHILS`)),
    pni         = suppressWarnings(as.numeric(`PROGNOSTIC NUTRITIONAL INDEX`))
  ) |>
  mutate(nlr = neutrophils / lymphocytes)

blood_long <- blood |>
  pivot_longer(c(lymphocytes, nlr, pni), names_to = "blood_metric", values_to = "value") |>
  filter(is.finite(ihc_immune_score), is.finite(value))

if (nrow(blood_long) > 2) {
  labs_tbl <- blood_long |>
    group_by(blood_metric) |>
    group_modify(~ paired_spearman(.x$ihc_immune_score, .x$value)) |>
    mutate(lab = sprintf("rho = %.2f (p = %.3g, n = %d)", rho, p, n))
  ggplot(blood_long, aes(ihc_immune_score, value)) +
    geom_point(alpha = 0.7, size = 2) +
    geom_smooth(method = "lm", se = FALSE, colour = "steelblue", formula = y ~ x) +
    geom_text(data = labs_tbl, aes(x = -Inf, y = Inf, label = lab),
              hjust = -0.05, vjust = 1.4, inherit.aes = FALSE, size = 3) +
    facet_wrap(~ blood_metric, scales = "free_y") +
    labs(title = "IHC immune score vs peripheral-blood immunity",
         x = "IHC immune score (CD45+ fraction)", y = NULL) +
    theme_classic(base_size = 11)
}
```

### Induction-treatment response

`ihc_immune_score` distribution grouped by `RESPONSE` (factor ordered by ascending
class value, 1 -> 3); no summary statistic beyond the group boxplots.

```{r response}
if ("RESPONSE" %in% names(ihc_clin)) {
  df <- ihc_clin |>
    filter(!is.na(RESPONSE), is.finite(ihc_immune_score))
  df$RESPONSE <- factor(as.character(df$RESPONSE),
                        levels = as.character(sort(unique(as.numeric(as.character(df$RESPONSE))))))
  if (dplyr::n_distinct(df$RESPONSE) > 1 && nrow(df) > 3) {
    ggplot(df, aes(RESPONSE, ihc_immune_score, colour = RESPONSE)) +
      geom_boxplot(outlier.shape = NA) +
      geom_jitter(width = 0.2, alpha = 0.6) +
      labs(title = "IHC immune score by induction response",
           x = NULL, y = "IHC immune score (CD45+ fraction)") +
      theme_classic(base_size = 12) +
      theme(axis.text.x = element_text(angle = 30, hjust = 1), legend.position = "none")
  }
}
```

### Survival

```
group(p) = "high" if ihc_immune_score(p) >= median, else "low"   -> Kaplan-Meier
HR       = exp( Cox coefficient of ihc_immune_score )            (univariable coxph)
```

Event coding (`coerce_event`): yes/dead/relapse/1 -> 1; no/alive/censored/0 -> 0.
Time = `OS`/`PFS` (months); event = `FOLLOW UP` (death) / `PERSISTENCE OR RELAPSE?`
(progression).

```{r survival}
surv_cfg <- list(
  OS  = list(time = "OS",  event = "FOLLOW UP"),
  PFS = list(time = "PFS", event = "PERSISTENCE OR RELAPSE?")
)

coerce_event <- function(x) {
  if (is.numeric(x)) return(as.integer(x > 0))
  v <- tolower(trimws(as.character(x)))
  dplyr::case_when(
    v %in% c("1","yes","y","dead","death","deceased","event","relapse",
             "progression","progressed","true") ~ 1L,
    v %in% c("0","no","n","alive","ned","censored","false") ~ 0L,
    TRUE ~ NA_integer_
  )
}

run_survival <- function(cfg, label) {
  if (!requireNamespace("survival", quietly = TRUE)) {
    message("survival package not installed — skipping ", label); return(invisible())
  }
  if (!all(c(cfg$time, cfg$event) %in% names(ihc_clin))) {
    message("missing ", cfg$time, "/", cfg$event, " — skipping ", label); return(invisible())
  }
  d <- ihc_clin |>
    transmute(ihc_immune_score,
              .time  = suppressWarnings(as.numeric(.data[[cfg$time]])),
              .event = coerce_event(.data[[cfg$event]])) |>
    filter(is.finite(.time), !is.na(.event), is.finite(ihc_immune_score))
  if (nrow(d) < 5 || dplyr::n_distinct(d$.event) < 2) {
    message("too few informative rows for ", label, " (n = ", nrow(d), ")"); return(invisible())
  }
  d <- d |> mutate(immune_group = if_else(ihc_immune_score >= median(ihc_immune_score),
                                          "IHC-immune high", "IHC-immune low"))
  fit  <- survival::survfit(survival::Surv(.time, .event) ~ immune_group, data = d)
  cox  <- survival::coxph(survival::Surv(.time, .event) ~ ihc_immune_score, data = d)
  hr   <- exp(stats::coef(cox))[1]; pcox <- summary(cox)$coefficients[1, "Pr(>|z|)"]
  cat(sprintf("%s: Cox HR per unit IHC immune score = %.2f (p = %.3g, n = %d)\n",
              label, hr, pcox, nrow(d)))
  if (requireNamespace("survminer", quietly = TRUE)) {
    print(survminer::ggsurvplot(fit, data = d, pval = TRUE, risk.table = TRUE,
                                title = paste(label, "by IHC immune score (median split)"),
                                legend.title = "")$plot)
  } else {
    plot(fit, col = c("firebrick", "steelblue"), xlab = "time", ylab = "survival",
         main = paste(label, "by IHC immune score"))
    legend("topright", levels(factor(d$immune_group)),
           col = c("firebrick", "steelblue"), lty = 1, bty = "n")
  }
}

invisible(purrr::iwalk(surv_cfg, ~ run_survival(.x, .y)))
```

## Notes

- Membership is **sf point-in-polygon** on the pathologist geojson (centroids
  mapped to the pixel frame by `/0.325`); the FlowPath `Out_of_annotation` flag is
  used only for patients with no geojson. The `source` column marks which per row.
- Whole-slide marker fractions use `na.rm`, so the real denominator is the count of
  cells with a non-`NA` sign, not necessarily every cell of the patient.
- Cohorts are small; every correlation is Spearman with its `n` reported. Read
  associations as descriptive.

```{r export-pdfs, include=FALSE}
# Copy every plot from this analysis (each rendered as a PDF via the
# dev = c("png", "pdf") chunk option) into output/figures/clinical_data/. Runs last so
# that all figures from the chunks above already exist on disk.
export_pdf_figures("clinical_data")
```
