---
title: "deconvolution"
author: "sceriff0"
date: "2026-07-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: needs immunedeconv (multi-method deconvolution) + tidyverse, here.
#   immunedeconv lives on GitHub (omnideconv/immunedeconv), NOT Bioconductor,
#   and pulls many Bioc deps. Install once, then snapshot:
#   Bioconductor repos must be active (immunedeconv deps are Bioc packages):
#     install.packages("BiocManager")
#     options(repos = BiocManager::repositories())   # CRAN + all Bioc sub-repos
#     renv::install("omnideconv/immunedeconv"); renv::snapshot()
#   # HPC without GitHub access? conda alternative: conda install bioconda::r-immunedeconv
#   BayesPrism (optional, single-cell reference) uses the sibling omnideconv pkg:
#     renv::install("omnideconv/omnideconv"); renv::snapshot()   # only if you have a reference
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
})

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6)

invisible(capture.output(suppressMessages(source(here("code", "test.R")))))
source(here("code", "validation_helpers.R"))
```

## Purpose

Validate the IHC immune-cell fractions against **bulk-RNA deconvolution**. Run
every applicable `immunedeconv` method, map each method's cell types onto the IHC
phenotype lineages, correlate per lineage across patients, rank **which method
agrees best with the IHC readout**, and cross-check **how well the methods agree
with each other** (an IHC-independent consistency check).

IHC reference = whole-slide lineage fractions from `phenotype_clean` (CD8T, CD4T,
Treg, NK).

### Scope: why four lineages, and which methods

**Why only four cell types are compared.** The deconvolution methods emit many
cell types (xCell ~64, quanTIseq ~10, CIBERSORT/LM22 22). The *comparison* is
capped by the reference, not the methods: it is anchored to the IHC panel, whose
`phenotype_clean` labels map cleanly onto only four lineages — CD8T, CD4T, Treg,
NK (`comparable_lineages`; the map is `phenotype_lineage` in
`validation_helpers.R`). The panel has **no macrophage or B-cell phenotype**, so
those deconvolution outputs have no phenotype-level IHC counterpart and cannot be
validated here (`bulkRna.Rmd` does a marker-level CD163/CD14 check instead).
`deconv_to_lineage()` collapses each method's many types down to these four; every
other type maps to `NA` and is dropped from the concordance. So "few cell types"
is a property of the validation target, and widening it needs a richer IHC panel,
not more methods.

**Why these methods (and the two that need extra inputs).** Methods differ in what
they need to run:

- **Run on a bulk matrix alone** — quanTIseq, EPIC (absolute fractions, apples-to-
  apples with IHC), MCPcounter, xCell (relative scores), and **ABIS**. All enabled.
- **Need a cancer-type indication per sample** — **TIMER**, **ConsensusTME**.
  Enabled only if `INDICATIONS` is set (this cohort doesn't encode a TCGA code).
- **License-gated signature** — **CIBERSORT / CIBERSORT (abs.)** need `CIBERSORT.R`
  + `LM22.txt` from cibersort.stanford.edu. Scaffolded: run if the files are
  configured, skipped cleanly otherwise.
- **Need a single-cell reference** — **BayesPrism** (and the other second-generation
  methods) are not in `immunedeconv`'s bulk interface; they live in the sibling
  `omnideconv` package and require a cell-type-annotated scRNA-seq reference, which
  this project does not ship. Scaffolded: run if a reference is dropped in,
  skipped cleanly otherwise.

So the original four were simply the subset that runs with zero extra artifacts;
the sections below add the rest behind graceful guards.

## Bulk-RNA expression matrix

`immunedeconv` wants a gene × sample matrix with gene symbols as row names.
`counts_data` is DESeq2-normalized (sample × gene); transpose it.

```{r expr-matrix}
expr_mat <- counts_data |>
  column_to_rownames("Sample") |>
  as.matrix() |>
  t()

storage.mode(expr_mat) <- "double"
cat("expression matrix:", nrow(expr_mat), "genes x", ncol(expr_mat), "samples\n")
```

> Caveat: quanTIseq/EPIC are designed for TPM. DESeq2-normalized counts are used
> here as a pragmatic proxy; absolute fractions are therefore approximate, but the
> *cross-patient ranking* (what the Spearman validation uses) is robust to this.

## Method configuration

```{r method-config}
# First-generation methods that run on the bulk matrix with no extra artifacts.
methods_free <- c("quantiseq", "epic", "mcp_counter", "xcell", "abis")

# TIMER / ConsensusTME need one cancer-type code per SAMPLE (length == ncol(expr_mat),
# same order). Set INDICATIONS to enable them, e.g. rep("READ", ncol(expr_mat)) for a
# single-indication cohort, or a per-sample vector. NULL -> both skipped.
INDICATIONS        <- NULL
methods_indication <- if (!is.null(INDICATIONS)) c("timer", "consensus_tme") else character(0)

# CIBERSORT is license-gated: register at cibersort.stanford.edu, download CIBERSORT.R
# and LM22.txt, and point these at them (env vars keep the paths out of the repo).
# Absent -> CIBERSORT and CIBERSORT (abs.) are skipped.
CIBERSORT_BINARY <- Sys.getenv("CIBERSORT_BINARY", unset = NA)  # path to CIBERSORT.R
CIBERSORT_MAT    <- Sys.getenv("CIBERSORT_MAT",    unset = NA)  # path to LM22.txt

# BayesPrism (second-generation) needs a cell-type-annotated scRNA-seq reference,
# read from here() below. Absent -> BayesPrism is skipped.
SC_REFERENCE <- here("data", "sc_reference.rds")
```

## Run deconvolution methods

```{r deconvolve}
# Bulk RNA samples are in CRF-id space; IHC is slide-id space. Bridge via clinical
# with a base-R named-vector lookup (crf id -> slide id), NOT a join.
crf2slide <- crf_to_slide_map(clinical_data)

deconv_long <- NULL
if (requireNamespace("immunedeconv", quietly = TRUE)) {
  # indication-aware wrapper: TIMER/ConsensusTME need `indications`, the rest don't.
  safe_deconv <- purrr::possibly(function(m) {
    args <- list(gene_expression = expr_mat, method = m)
    if (m %in% c("timer", "consensus_tme")) args$indications <- INDICATIONS
    do.call(immunedeconv::deconvolute, args) |>
      pivot_longer(-cell_type, names_to = "sample", values_to = "score") |>
      mutate(method = m)
  }, otherwise = NULL, quiet = FALSE)

  parts <- purrr::map(c(methods_free, methods_indication), safe_deconv)

  # CIBERSORT (+ abs): only when the license-gated files are configured & present.
  if (!is.na(CIBERSORT_BINARY) && !is.na(CIBERSORT_MAT) &&
      file.exists(CIBERSORT_BINARY) && file.exists(CIBERSORT_MAT)) {
    immunedeconv::set_cibersort_binary(CIBERSORT_BINARY)
    immunedeconv::set_cibersort_mat(CIBERSORT_MAT)
    parts <- c(parts, purrr::map(c("cibersort", "cibersort_abs"), safe_deconv))
  } else {
    message("CIBERSORT skipped — set CIBERSORT_BINARY + CIBERSORT_MAT (LM22.txt) to enable.")
  }

  deconv_long <- parts |>
    purrr::compact() |>
    bind_rows() |>
    mutate(crf_id     = norm_id(sample),
           patient_id = unname(crf2slide[crf_id]),   # crf id -> slide id
           lineage    = deconv_to_lineage(cell_type))

  n_unmapped  <- dplyr::n_distinct(deconv_long$crf_id[is.na(deconv_long$patient_id)])
  deconv_long <- deconv_long |> filter(!is.na(patient_id))

  cat("deconvolution rows:", nrow(deconv_long),
      "across", dplyr::n_distinct(deconv_long$method), "method(s):",
      paste(sort(unique(deconv_long$method)), collapse = ", "), "\n",
      n_unmapped, "RNA sample(s) had no slide crosswalk\n")
} else {
  message("immunedeconv not installed — install it to run this report.")
}
```

## BayesPrism (single-cell reference; scaffolded)

`immunedeconv`'s `deconvolute()` covers only bulk-signature methods. BayesPrism is
a **second-generation** method: it estimates fractions from a cell-type-annotated
scRNA-seq reference, via the sibling **`omnideconv`** package. This project ships
no such reference, so the block below is a ready-to-run scaffold — it activates
only when a reference is present at `data/sc_reference.rds` (an `.rds` holding a
list: `counts` = gene×cell matrix, `labels` = per-cell type, `batch` = per-cell
sample id), and skips cleanly otherwise. Verify the argument names against your
installed `omnideconv` version before trusting the numbers.

```{r bayesprism}
bayes_long <- NULL
if (file.exists(SC_REFERENCE) && requireNamespace("omnideconv", quietly = TRUE)) {
  ref <- readRDS(SC_REFERENCE)
  bayes_long <- tryCatch({
    # build_model(): learn the reference; deconvolute(): apply it to the bulk matrix.
    model <- omnideconv::build_model(
      single_cell_object    = ref$counts,
      cell_type_annotations = ref$labels,
      batch_ids             = ref$batch,
      method                = "bayesprism",
      bulk_gene_expression  = expr_mat)
    frac <- omnideconv::deconvolute(
      bulk_gene_expression  = expr_mat,
      model                 = model,
      single_cell_object    = ref$counts,
      cell_type_annotations = ref$labels,
      batch_ids             = ref$batch,
      method                = "bayesprism")
    as.data.frame(frac) |>
      tibble::rownames_to_column("sample") |>            # omnideconv returns sample x cell_type
      pivot_longer(-sample, names_to = "cell_type", values_to = "score") |>
      mutate(method = "bayesprism")
  }, error = function(e) { message("BayesPrism failed: ", conditionMessage(e)); NULL })
} else {
  message("BayesPrism skipped — needs data/sc_reference.rds + the omnideconv package.")
}

if (!is.null(bayes_long) && !is.null(deconv_long)) {
  bayes_long <- bayes_long |>
    mutate(crf_id     = norm_id(sample),
           patient_id = unname(crf2slide[crf_id]),
           lineage    = deconv_to_lineage(cell_type)) |>
    filter(!is.na(patient_id))
  deconv_long <- bind_rows(deconv_long, bayes_long)
  cat("BayesPrism added:", nrow(bayes_long), "rows\n")
}
```

## IHC lineage fractions (reference)

```{r ihc-lineages}
lin_raw <- ihc_lineage_fraction(ihc_data) |>
  filter(lineage %in% comparable_lineages) |>
  select(patient_id, lineage, ihc_frac = frac)

# Complete the patient x lineage grid: a lineage a patient simply lacks must read
# as 0 (present), not a missing row. Otherwise the inner_join below silently drops
# that patient from that lineage's panel — the "plot never shows all patients" bug.
ihc_lin <- expand.grid(patient_id = unique(slide_key(ihc_data$patient_id)),
                       lineage    = comparable_lineages,
                       stringsAsFactors = FALSE) |>
  left_join(lin_raw, by = c("patient_id", "lineage")) |>
  mutate(ihc_frac = ifelse(is.na(ihc_frac), 0, ihc_frac))

cat("IHC patients with lineage fractions:", dplyr::n_distinct(ihc_lin$patient_id), "\n")
```

## Method vs IHC — per-lineage concordance

For each method, collapse its cell types to the four comparable lineages (sum
where a method splits a lineage, e.g. quanTIseq's M1/M2 do not map here), join to
the IHC fraction on patient + lineage, and correlate across patients.

```{r concordance}
paired <- NULL
if (!is.null(deconv_long) && nrow(deconv_long) > 0) {
  deconv_lin <- deconv_long |>
    filter(!is.na(lineage)) |>
    group_by(method, patient_id, lineage) |>
    summarise(score = sum(score), .groups = "drop")

  paired <- deconv_lin |>
    inner_join(ihc_lin, by = c("patient_id", "lineage")) |>
    filter(is.finite(score), is.finite(ihc_frac))
  cat("paired (method x lineage x patient) rows:", nrow(paired), "\n")
}
```

```{r concordance-table}
if (!is.null(paired) && nrow(paired) > 0) {
  summary_tbl <- paired |>
    group_by(method, lineage) |>
    group_modify(~ paired_spearman(.x$score, .x$ihc_frac)) |>
    ungroup() |>
    arrange(lineage, desc(rho))
  knitr::kable(summary_tbl, digits = 3,
               caption = "Spearman IHC vs deconvolution, per method and lineage")

  # Which method agrees best overall (mean rho across lineages).
  method_rank <- summary_tbl |>
    group_by(method) |>
    summarise(mean_rho = mean(rho, na.rm = TRUE),
              n_lineages = sum(is.finite(rho)), .groups = "drop") |>
    arrange(desc(mean_rho))
  knitr::kable(method_rank, digits = 3,
               caption = "Method ranking by mean Spearman across lineages")
}
```

```{r concordance-plot}
if (!is.null(paired) && nrow(paired) > 0) {
  ggplot(paired, aes(ihc_frac, score)) +
    geom_point(alpha = 0.6, size = 1.8) +
    geom_smooth(method = "lm", se = FALSE, colour = "steelblue", formula = y ~ x) +
    facet_grid(method ~ lineage, scales = "free") +
    labs(title = "IHC lineage fraction vs bulk-RNA deconvolution score",
         x = "IHC fraction of all cells", y = "deconvolution score") +
    theme_bw(base_size = 10)
}
```

## Cross-method agreement (methods vs each other and vs IHC)

Concordance with IHC only tells us which method matches *one* reference. Whether the
methods agree with **each other** is an independent consistency check: cell types on
which several independent algorithms converge are more trustworthy, and a method
that agrees with the others but not IHC (or vice-versa) localises where the
disagreement is. We treat IHC as one more "source" and compute, for every source
pair, the mean Spearman across the four lineages (rank-based, so the methods'
different score scales don't matter).

```{r agreement}
agr <- NULL
if (!is.null(deconv_long) && nrow(deconv_long) > 0) {
  # per (method, patient, lineage) score, summing a method's split sub-types.
  deconv_lin <- deconv_long |>
    filter(!is.na(lineage)) |>
    group_by(method, patient_id, lineage) |>
    summarise(score = sum(score), .groups = "drop")

  agree_src <- bind_rows(
    deconv_lin |> transmute(source = method, patient_id, lineage, value = score),
    ihc_lin    |> transmute(source = "IHC",  patient_id, lineage, value = ihc_frac)
  )
  agr <- pairwise_agreement(agree_src)
}
```

```{r agreement-heatmap, fig.height=6}
if (!is.null(agr) && nrow(agr) > 0) {
  # order sources by mean off-diagonal agreement (most-consensual first); keep IHC visible.
  ord <- agr |>
    filter(a != b) |>
    group_by(a) |>
    summarise(m = mean(mean_rho, na.rm = TRUE), .groups = "drop") |>
    arrange(m) |>
    pull(a)
  agr <- agr |> mutate(a = factor(a, levels = ord), b = factor(b, levels = ord))

  ggplot(agr, aes(a, b, fill = mean_rho)) +
    geom_tile(colour = "white") +
    geom_text(aes(label = ifelse(is.na(mean_rho), "·", sprintf("%.2f", mean_rho))),
              size = 3) +
    scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                         midpoint = 0, limits = c(-1, 1), name = "mean rho",
                         na.value = "grey90") +
    coord_equal() +
    labs(title = "Cross-source agreement (mean Spearman across lineages)",
         subtitle = "Each deconvolution method vs every other and vs IHC; averaged over CD8T/CD4T/Treg/NK",
         x = NULL, y = NULL) +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 40, hjust = 1),
          panel.grid = element_blank())
}
```

```{r agreement-ranks}
if (!is.null(agr) && nrow(agr) > 0) {
  # How much each source agrees with the consensus of the OTHERS (mean off-diagonal),
  # and separately with IHC alone.
  vs_ihc <- agr |> filter(b == "IHC", a != "IHC") |> select(source = a, rho_vs_IHC = mean_rho)
  agr |>
    filter(a != b) |>
    group_by(source = a) |>
    summarise(mean_rho_vs_others = mean(mean_rho, na.rm = TRUE), .groups = "drop") |>
    left_join(vs_ihc, by = "source") |>
    arrange(desc(mean_rho_vs_others)) |>
    knitr::kable(digits = 3,
      caption = "Per-source agreement with the other sources vs with IHC alone. A high vs-others but low vs-IHC source (or the reverse) flags where methods and IHC diverge.")
}
```

## Notes

- Absolute-fraction methods (quanTIseq, EPIC, CIBERSORT abs., ABIS) are the fair
  apples-to-apples comparison; MCPcounter/xCell/CIBERSORT (rel.) scores are
  relative, so only their *rank* correlation is meaningful, not the slope. The
  cross-method agreement panel is Spearman throughout for exactly this reason.
- Only CD8T / CD4T / Treg / NK are compared — the IHC panel has no macrophage or
  B-cell phenotype, so those deconvolution outputs have no phenotype-level IHC
  counterpart (see `bulkRna.Rmd` for a marker-level CD163/CD14 check instead).
  Widening the comparison needs a richer IHC panel, not more methods.
- Method coverage is gated by inputs, not preference: quanTIseq/EPIC/MCPcounter/
  xCell/ABIS run out of the box; TIMER/ConsensusTME need `INDICATIONS`;
  CIBERSORT needs its license-gated `LM22.txt` + `CIBERSORT.R`; BayesPrism needs
  an scRNA-seq reference at `data/sc_reference.rds` via `omnideconv`. Each is
  skipped cleanly (with a message) when its input is absent, so the report always
  knits with whatever is available.
- The CIBERSORT and BayesPrism blocks are scaffolds — verify `set_cibersort_*()`
  and `omnideconv::build_model()/deconvolute()` argument names against your
  installed versions before trusting their output.
- Cross-patient Spearman needs ≥ 3 shared patients per lineage; sparser cells
  return `NA` rather than a spurious correlation. With this cohort's size the
  agreement matrix is descriptive — read it for *pattern* (which methods cluster),
  not significance.
