---
title: "deconvolution"
author: "sceriff0"
date: "2026-07-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: needs immunedeconv (multi-method deconvolution) + tidyverse, here.
#   immunedeconv lives on GitHub (omnideconv/immunedeconv), NOT Bioconductor,
#   and pulls many Bioc deps. Install once, then snapshot:
#   Bioconductor repos must be active (immunedeconv deps are Bioc packages):
#     install.packages("BiocManager")
#     options(repos = BiocManager::repositories())   # CRAN + all Bioc sub-repos
#     renv::install("omnideconv/immunedeconv"); renv::snapshot()
#   # HPC without GitHub access? conda alternative: conda install bioconda::r-immunedeconv
#   BayesPrism (optional, single-cell reference) uses the sibling omnideconv pkg:
#     renv::install("omnideconv/omnideconv"); renv::snapshot()   # only if you have a reference
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
})

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6)

invisible(capture.output(suppressMessages(source(here("code", "test.R")))))
source(here("code", "validation_helpers.R"))
```

## Starting data and notation

Every value below is derived from two inputs, bridged by the clinical table:

- `ihc_data` — FlowPath single cells deduplicated to **one copy per patient**
  (whole slide, no annotation boundary). $C_p$ is the set of all cells of patient
  $p$; $\text{lineage}(c)$ is `phenotype_clean` collapsed via the `phenotype_lineage`
  table.
- `counts_data` — DESeq2-normalized bulk RNA, one row per sample, one column per
  gene, keyed by `Sample` (a CRF id).
- `clinical_data` — supplies the CRF$\leftrightarrow$slide id bridge.

Id spaces and the bridge. Bulk RNA lives in **CRF-id** space; IHC lives in
**slide-id** space. `slide_key()` maps any id to the shared numeric slide space
(strip non-digits, apply the verified typo fix `15879`$\to$`15897`). `norm_id()`
upper-cases and strips non-alphanumerics. `crf_to_slide_map(clinical_data)` returns
a named vector $\texttt{crf2slide}[\,\text{crf}\,]=\text{slide}$, built from
`norm_id(`ID CRF PRESERVE`)`$\to$`slide_key(`ID PATIENT`)`.

Comparison is anchored to $\mathcal{L}=\{$CD8T, CD4T, Treg, NK$\}$
(`comparable_lineages`) — the only IHC phenotype lineages with a clean
deconvolution counterpart. `deconv_to_lineage()` maps each method's cell-type
string to an $\ell\in\mathcal{L}$ (case-insensitive pattern match:
`regulatory|treg`$\to$Treg, `cd8`$\to$CD8T, `cd4`$\to$CD4T,
`^nk|nk cell|natural killer`$\to$NK), everything else to `NA`. All correlations
are Spearman $\rho$ via `paired_spearman` (rank-based; `NA` when $n<3$ shared
patients or either vector is constant).

## Bulk-RNA expression matrix

$\texttt{expr\_mat}$ is `counts_data` (sample $\times$ gene) transposed to
gene $\times$ sample, cast double, with gene symbols as row names:
$\texttt{expr\_mat}[g,s]=\texttt{counts\_data}[s,g]$.

```{r expr-matrix}
expr_mat <- counts_data |>
  column_to_rownames("Sample") |>
  as.matrix() |>
  t()

storage.mode(expr_mat) <- "double"
cat("expression matrix:", nrow(expr_mat), "genes x", ncol(expr_mat), "samples\n")
```

> Caveat: quanTIseq/EPIC are designed for TPM. DESeq2-normalized counts are used
> here as a pragmatic proxy; absolute fractions are therefore approximate, but the
> *cross-patient ranking* (what the Spearman validation uses) is robust to this.

## Method configuration

Each method is gated by the inputs it needs, not preference:

- `methods_free` (quanTIseq, EPIC, MCPcounter, xCell, ABIS) — run on `expr_mat`
  alone. Always enabled.
- `methods_indication` (TIMER, ConsensusTME) — need one cancer-type code per
  sample (`INDICATIONS`, length $=$ `ncol(expr_mat)`). Enabled iff `INDICATIONS`
  is non-`NULL`; here it is `NULL`, so both are skipped.
- CIBERSORT / CIBERSORT (abs.) — license-gated `CIBERSORT.R` + `LM22.txt`, paths
  from env vars `CIBERSORT_BINARY` / `CIBERSORT_MAT`. Run iff both are set and the
  files exist.
- BayesPrism — needs an scRNA-seq reference at `data/sc_reference.rds` via
  `omnideconv`. Run iff present.

```{r method-config}
# First-generation methods that run on the bulk matrix with no extra artifacts.
methods_free <- c("quantiseq", "epic", "mcp_counter", "xcell", "abis")

# TIMER / ConsensusTME need one cancer-type code per SAMPLE (length == ncol(expr_mat),
# same order). Set INDICATIONS to enable them, e.g. rep("READ", ncol(expr_mat)) for a
# single-indication cohort, or a per-sample vector. NULL -> both skipped.
INDICATIONS        <- NULL
methods_indication <- if (!is.null(INDICATIONS)) c("timer", "consensus_tme") else character(0)

# CIBERSORT is license-gated: register at cibersort.stanford.edu, download CIBERSORT.R
# and LM22.txt, and point these at them (env vars keep the paths out of the repo).
# Absent -> CIBERSORT and CIBERSORT (abs.) are skipped.
CIBERSORT_BINARY <- Sys.getenv("CIBERSORT_BINARY", unset = NA)  # path to CIBERSORT.R
CIBERSORT_MAT    <- Sys.getenv("CIBERSORT_MAT",    unset = NA)  # path to LM22.txt

# BayesPrism (second-generation) needs a cell-type-annotated scRNA-seq reference,
# read from here() below. Absent -> BayesPrism is skipped.
SC_REFERENCE <- here("data", "sc_reference.rds")
```

## Run deconvolution methods

Each enabled method $m$ produces a cell-type $\times$ sample score table,
pivoted long to rows $(\text{cell\_type}, \text{sample}, \text{score}, m)$. Then
$\text{crf\_id}=\texttt{norm\_id}(\text{sample})$,
$\text{patient\_id}=\texttt{crf2slide}[\text{crf\_id}]$ (slide space), and
$\text{lineage}=\texttt{deconv\_to\_lineage}(\text{cell\_type})$. Rows with no
slide crosswalk (`patient_id` is `NA`) are dropped.

```{r deconvolve}
# Bulk RNA samples are in CRF-id space; IHC is slide-id space. Bridge via clinical
# with a base-R named-vector lookup (crf id -> slide id), NOT a join.
crf2slide <- crf_to_slide_map(clinical_data)

deconv_long <- NULL
if (requireNamespace("immunedeconv", quietly = TRUE)) {
  # indication-aware wrapper: TIMER/ConsensusTME need `indications`, the rest don't.
  safe_deconv <- purrr::possibly(function(m) {
    args <- list(gene_expression = expr_mat, method = m)
    if (m %in% c("timer", "consensus_tme")) args$indications <- INDICATIONS
    do.call(immunedeconv::deconvolute, args) |>
      pivot_longer(-cell_type, names_to = "sample", values_to = "score") |>
      mutate(method = m)
  }, otherwise = NULL, quiet = FALSE)

  parts <- purrr::map(c(methods_free, methods_indication), safe_deconv)

  # CIBERSORT (+ abs): only when the license-gated files are configured & present.
  if (!is.na(CIBERSORT_BINARY) && !is.na(CIBERSORT_MAT) &&
      file.exists(CIBERSORT_BINARY) && file.exists(CIBERSORT_MAT)) {
    immunedeconv::set_cibersort_binary(CIBERSORT_BINARY)
    immunedeconv::set_cibersort_mat(CIBERSORT_MAT)
    parts <- c(parts, purrr::map(c("cibersort", "cibersort_abs"), safe_deconv))
  } else {
    message("CIBERSORT skipped — set CIBERSORT_BINARY + CIBERSORT_MAT (LM22.txt) to enable.")
  }

  deconv_long <- parts |>
    purrr::compact() |>
    bind_rows() |>
    mutate(crf_id     = norm_id(sample),
           patient_id = unname(crf2slide[crf_id]),   # crf id -> slide id
           lineage    = deconv_to_lineage(cell_type))

  n_unmapped  <- dplyr::n_distinct(deconv_long$crf_id[is.na(deconv_long$patient_id)])
  deconv_long <- deconv_long |> filter(!is.na(patient_id))

  cat("deconvolution rows:", nrow(deconv_long),
      "across", dplyr::n_distinct(deconv_long$method), "method(s):",
      paste(sort(unique(deconv_long$method)), collapse = ", "), "\n",
      n_unmapped, "RNA sample(s) had no slide crosswalk\n")
} else {
  message("immunedeconv not installed — install it to run this report.")
}
```

## BayesPrism (single-cell reference; scaffolded)

`immunedeconv::deconvolute()` covers only bulk-signature methods. BayesPrism is a
second-generation method (fractions from a cell-type-annotated scRNA-seq
reference, via `omnideconv`); this project ships no reference, so the block is a
ready-to-run scaffold that activates only when `data/sc_reference.rds` is present
(an `.rds` list: `counts` = gene$\times$cell, `labels` = per-cell type, `batch` =
per-cell sample id) and skips cleanly otherwise. Its output is mapped to
$(\text{patient\_id}, \text{lineage})$ by the same
$\texttt{norm\_id}\to\texttt{crf2slide}$ / `deconv_to_lineage` pipeline as the
other methods, then row-bound into `deconv_long`. Verify the `omnideconv`
argument names against your installed version before trusting the numbers.

```{r bayesprism}
bayes_long <- NULL
if (file.exists(SC_REFERENCE) && requireNamespace("omnideconv", quietly = TRUE)) {
  ref <- readRDS(SC_REFERENCE)
  bayes_long <- tryCatch({
    # build_model(): learn the reference; deconvolute(): apply it to the bulk matrix.
    model <- omnideconv::build_model(
      single_cell_object    = ref$counts,
      cell_type_annotations = ref$labels,
      batch_ids             = ref$batch,
      method                = "bayesprism",
      bulk_gene_expression  = expr_mat)
    frac <- omnideconv::deconvolute(
      bulk_gene_expression  = expr_mat,
      model                 = model,
      single_cell_object    = ref$counts,
      cell_type_annotations = ref$labels,
      batch_ids             = ref$batch,
      method                = "bayesprism")
    as.data.frame(frac) |>
      tibble::rownames_to_column("sample") |>            # omnideconv returns sample x cell_type
      pivot_longer(-sample, names_to = "cell_type", values_to = "score") |>
      mutate(method = "bayesprism")
  }, error = function(e) { message("BayesPrism failed: ", conditionMessage(e)); NULL })
} else {
  message("BayesPrism skipped — needs data/sc_reference.rds + the omnideconv package.")
}

if (!is.null(bayes_long) && !is.null(deconv_long)) {
  bayes_long <- bayes_long |>
    mutate(crf_id     = norm_id(sample),
           patient_id = unname(crf2slide[crf_id]),
           lineage    = deconv_to_lineage(cell_type)) |>
    filter(!is.na(patient_id))
  deconv_long <- bind_rows(deconv_long, bayes_long)
  cat("BayesPrism added:", nrow(bayes_long), "rows\n")
}
```

## IHC lineage fractions (reference)

`ihc_lineage_fraction(ihc_data)` counts cells per $(p,\ell)$ and normalises by the
patient's total cell count (over **all** lineages, `NA` included), giving the
whole-slide fraction

$$
\text{ihc\_frac}_{p,\ell}=\frac{|\{c\in C_p:\text{lineage}(c)=\ell\}|}{|C_p|}.
$$

Restricted to $\ell\in\mathcal{L}$, then **completed to 0** across the full
$\{p\}\times\mathcal{L}$ grid: a lineage a patient simply lacks reads as $0$
(present), not a missing row — otherwise the `inner_join` below would silently
drop that patient from that lineage's panel.

```{r ihc-lineages}
lin_raw <- ihc_lineage_fraction(ihc_data) |>
  filter(lineage %in% comparable_lineages) |>
  select(patient_id, lineage, ihc_frac = frac)

# Complete the patient x lineage grid: a lineage a patient simply lacks must read
# as 0 (present), not a missing row. Otherwise the inner_join below silently drops
# that patient from that lineage's panel — the "plot never shows all patients" bug.
ihc_lin <- expand.grid(patient_id = unique(slide_key(ihc_data$patient_id)),
                       lineage    = comparable_lineages,
                       stringsAsFactors = FALSE) |>
  left_join(lin_raw, by = c("patient_id", "lineage")) |>
  mutate(ihc_frac = ifelse(is.na(ihc_frac), 0, ihc_frac))

cat("IHC patients with lineage fractions:", dplyr::n_distinct(ihc_lin$patient_id), "\n")
```

## Method vs IHC — per-lineage concordance

Per-$(m,p,\ell)$ score sums a method's sub-types mapping to the same lineage
(e.g. two NK sub-types):

$$
\text{score}_{m,p,\ell}=\sum_{\substack{\text{rows of }m\text{ for }p\\ \texttt{deconv\_to\_lineage}(\text{cell\_type})=\ell}} \text{score}.
$$

This is inner-joined to $\text{ihc\_frac}_{p,\ell}$ on $(p,\ell)$, keeping finite
pairs.

```{r concordance}
paired <- NULL
if (!is.null(deconv_long) && nrow(deconv_long) > 0) {
  deconv_lin <- deconv_long |>
    filter(!is.na(lineage)) |>
    group_by(method, patient_id, lineage) |>
    summarise(score = sum(score), .groups = "drop")

  paired <- deconv_lin |>
    inner_join(ihc_lin, by = c("patient_id", "lineage")) |>
    filter(is.finite(score), is.finite(ihc_frac))
  cat("paired (method x lineage x patient) rows:", nrow(paired), "\n")
}
```

Per $(m,\ell)$, concordance is the cross-patient Spearman
$\rho_{m,\ell}=\texttt{paired\_spearman}(\text{score}_{m,\cdot,\ell},\ \text{ihc\_frac}_{\cdot,\ell})$.
Method ranking is the mean over lineages,
$\overline{\rho}_m=\operatorname{mean}_{\ell}\rho_{m,\ell}$ (`na.rm`), with
$n_\ell=\#\{\ell:\rho_{m,\ell}\text{ finite}\}$.

```{r concordance-table}
if (!is.null(paired) && nrow(paired) > 0) {
  summary_tbl <- paired |>
    group_by(method, lineage) |>
    group_modify(~ paired_spearman(.x$score, .x$ihc_frac)) |>
    ungroup() |>
    arrange(lineage, desc(rho))
  knitr::kable(summary_tbl, digits = 3,
               caption = "Spearman IHC vs deconvolution, per method and lineage")

  # Which method agrees best overall (mean rho across lineages).
  method_rank <- summary_tbl |>
    group_by(method) |>
    summarise(mean_rho = mean(rho, na.rm = TRUE),
              n_lineages = sum(is.finite(rho)), .groups = "drop") |>
    arrange(desc(mean_rho))
  knitr::kable(method_rank, digits = 3,
               caption = "Method ranking by mean Spearman across lineages")
}
```

Scatter of $\text{ihc\_frac}$ ($x$) vs $\text{score}$ ($y$), faceted method
$\times$ lineage.

```{r concordance-plot}
if (!is.null(paired) && nrow(paired) > 0) {
  ggplot(paired, aes(ihc_frac, score)) +
    geom_point(alpha = 0.6, size = 1.8) +
    geom_smooth(method = "lm", se = FALSE, colour = "steelblue", formula = y ~ x) +
    facet_grid(method ~ lineage, scales = "free") +
    labs(title = "IHC lineage fraction vs bulk-RNA deconvolution score",
         x = "IHC fraction of all cells", y = "deconvolution score") +
    theme_bw(base_size = 10)
}
```

## Cross-method agreement (methods vs each other and vs IHC)

Treat IHC as one more "source" alongside each method, giving one value per
$(\text{source}, p, \ell)$: the per-lineage deconvolution score for a method, or
$\text{ihc\_frac}$ for source `IHC`.

```{r agreement}
agr <- NULL
if (!is.null(deconv_long) && nrow(deconv_long) > 0) {
  # per (method, patient, lineage) score, summing a method's split sub-types.
  deconv_lin <- deconv_long |>
    filter(!is.na(lineage)) |>
    group_by(method, patient_id, lineage) |>
    summarise(score = sum(score), .groups = "drop")

  agree_src <- bind_rows(
    deconv_lin |> transmute(source = method, patient_id, lineage, value = score),
    ihc_lin    |> transmute(source = "IHC",  patient_id, lineage, value = ihc_frac)
  )
  agr <- pairwise_agreement(agree_src)
}
```

`pairwise_agreement` returns a symmetric long table over every ordered source
pair $(a,b)$. For $a=b$ the entry is fixed at $\text{mean\_rho}=1$
($\text{n\_lineages}=$ `NA`). For $a\neq b$ it inner-joins the two sources on
$(p,\ell)$ (their shared patients), computes a per-lineage Spearman
$\rho_\ell=\texttt{paired\_spearman}(\text{value}_a,\text{value}_b)$ across those
shared patients, then averages:

$$
\text{mean\_rho}(a,b)=\operatorname{mean}_{\ell}\rho_\ell\ (\text{na.rm}),\qquad
\text{n\_lineages}(a,b)=\#\{\ell:\rho_\ell\text{ finite}\}.
$$

If the join is empty (no shared patients) the entry is $\text{mean\_rho}=$ `NA`,
$\text{n\_lineages}=0$. Rank-based throughout, so the methods' differing score
scales do not matter.

```{r agreement-heatmap, fig.height=6}
if (!is.null(agr) && nrow(agr) > 0) {
  # order sources by mean off-diagonal agreement (most-consensual first); keep IHC visible.
  ord <- agr |>
    filter(a != b) |>
    group_by(a) |>
    summarise(m = mean(mean_rho, na.rm = TRUE), .groups = "drop") |>
    arrange(m) |>
    pull(a)
  agr <- agr |> mutate(a = factor(a, levels = ord), b = factor(b, levels = ord))

  ggplot(agr, aes(a, b, fill = mean_rho)) +
    geom_tile(colour = "white") +
    geom_text(aes(label = ifelse(is.na(mean_rho), "·", sprintf("%.2f", mean_rho))),
              size = 3) +
    scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B",
                         midpoint = 0, limits = c(-1, 1), name = "mean rho",
                         na.value = "grey90") +
    coord_equal() +
    labs(title = "Cross-source agreement (mean Spearman across lineages)",
         subtitle = "Each deconvolution method vs every other and vs IHC; averaged over CD8T/CD4T/Treg/NK",
         x = NULL, y = NULL) +
    theme_minimal(base_size = 11) +
    theme(axis.text.x = element_text(angle = 40, hjust = 1),
          panel.grid = element_blank())
}
```

Per source $a$: $\text{mean\_rho\_vs\_others}=\operatorname{mean}_{b\neq a}\text{mean\_rho}(a,b)$
(agreement with the consensus of the others) and $\text{rho\_vs\_IHC}=\text{mean\_rho}(a,\text{IHC})$
(agreement with IHC alone). A high vs-others but low vs-IHC source (or the reverse)
localises where methods and IHC diverge.

```{r agreement-ranks}
if (!is.null(agr) && nrow(agr) > 0) {
  # How much each source agrees with the consensus of the OTHERS (mean off-diagonal),
  # and separately with IHC alone.
  vs_ihc <- agr |> filter(b == "IHC", a != "IHC") |> select(source = a, rho_vs_IHC = mean_rho)
  agr |>
    filter(a != b) |>
    group_by(source = a) |>
    summarise(mean_rho_vs_others = mean(mean_rho, na.rm = TRUE), .groups = "drop") |>
    left_join(vs_ihc, by = "source") |>
    arrange(desc(mean_rho_vs_others)) |>
    knitr::kable(digits = 3,
      caption = "Per-source agreement with the other sources vs with IHC alone. A high vs-others but low vs-IHC source (or the reverse) flags where methods and IHC diverge.")
}
```

## Notes

- Relative-score methods (MCPcounter, xCell, CIBERSORT rel.) are rank-only; their
  slope against IHC is not meaningful, so every statistic here is Spearman.
  Absolute-fraction methods (quanTIseq, EPIC, CIBERSORT abs., ABIS) are the fair
  apples-to-apples comparison but are still compared by rank for uniformity.
- Only $\mathcal{L}=\{$CD8T, CD4T, Treg, NK$\}$ are compared: the IHC panel has no
  macrophage or B-cell phenotype, so `deconv_to_lineage` returns `NA` for those
  and they are dropped (`bulkRna.Rmd` does a marker-level CD163/CD14 check
  instead).
- Cross-patient Spearman needs $\ge 3$ shared patients (and non-constant vectors);
  sparser $(m,\ell)$ or source pairs return `NA` rather than a spurious
  correlation, and `mean_rho`/`mean_rho` averages over the finite lineages only.
  With this cohort's size the agreement matrix is descriptive — read it for
  *pattern*, not significance.
- The CIBERSORT and BayesPrism blocks are scaffolds — verify `set_cibersort_*()`
  and `omnideconv::build_model()/deconvolute()` argument names against your
  installed versions before trusting their output.
</content>
</invoke>
