---
title: "deconvolution"
author: "sceriff0"
date: "2026-07-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: needs immunedeconv (multi-method deconvolution) + tidyverse, here.
#   immunedeconv lives on GitHub (omnideconv/immunedeconv), NOT Bioconductor,
#   and pulls many Bioc deps. Install once, then snapshot:
#   Bioconductor repos must be active (immunedeconv deps are Bioc packages):
#     install.packages("BiocManager")
#     options(repos = BiocManager::repositories())   # CRAN + all Bioc sub-repos
#     renv::install("omnideconv/immunedeconv"); renv::snapshot()
#   # HPC without GitHub access? conda alternative: conda install bioconda::r-immunedeconv
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
})

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6)

invisible(capture.output(suppressMessages(source(here("code", "test.R")))))
source(here("code", "validation_helpers.R"))
```

## Purpose

Validate the IHC immune-cell fractions against **bulk-RNA deconvolution**. Run
several methods via `immunedeconv` (quanTIseq + EPIC give absolute fractions
directly comparable to IHC; MCPcounter + xCell give relative scores), map each
method's cell types onto the IHC phenotype lineages, correlate per lineage across
patients, and rank **which method agrees best with the IHC readout**.

IHC reference = whole-slide lineage fractions from `phenotype_clean` (CD8T, CD4T,
Treg, NK — the lineages with a clean deconvolution counterpart; the panel has no
macrophage/B-cell phenotype).

## Bulk-RNA expression matrix

`immunedeconv` wants a gene × sample matrix with gene symbols as row names.
`counts_data` is DESeq2-normalized (sample × gene); transpose it.

```{r expr-matrix}
expr_mat <- counts_data |>
  column_to_rownames("Sample") |>
  as.matrix() |>
  t()

storage.mode(expr_mat) <- "double"
cat("expression matrix:", nrow(expr_mat), "genes x", ncol(expr_mat), "samples\n")
```

> Caveat: quanTIseq/EPIC are designed for TPM. DESeq2-normalized counts are used
> here as a pragmatic proxy; absolute fractions are therefore approximate, but the
> *cross-patient ranking* (what the Spearman validation uses) is robust to this.

## Run deconvolution methods

```{r deconvolve}
methods <- c("quantiseq", "epic", "mcp_counter", "xcell")

# Bulk RNA samples are in CRF-id space; IHC is slide-id space. Bridge via clinical
# with a base-R named-vector lookup (crf id -> slide id), NOT a join.
crf2slide <- crf_to_slide_map(clinical_data)

deconv_long <- NULL
if (requireNamespace("immunedeconv", quietly = TRUE)) {
  safe_deconv <- purrr::possibly(function(m) {
    immunedeconv::deconvolute(expr_mat, m) |>
      pivot_longer(-cell_type, names_to = "sample", values_to = "score") |>
      mutate(method = m)
  }, otherwise = NULL, quiet = FALSE)

  deconv_long <- purrr::map(methods, safe_deconv) |>
    purrr::compact() |>
    bind_rows() |>
    mutate(crf_id     = norm_id(sample),
           patient_id = unname(crf2slide[crf_id]),   # crf id -> slide id
           lineage    = deconv_to_lineage(cell_type))

  n_unmapped  <- dplyr::n_distinct(deconv_long$crf_id[is.na(deconv_long$patient_id)])
  deconv_long <- deconv_long |> filter(!is.na(patient_id))

  cat("deconvolution rows:", nrow(deconv_long),
      "across", dplyr::n_distinct(deconv_long$method), "methods;",
      n_unmapped, "RNA sample(s) had no slide crosswalk\n")
} else {
  message("immunedeconv not installed — install it to run this report.")
}
```

## IHC lineage fractions (reference)

```{r ihc-lineages}
lin_raw <- ihc_lineage_fraction(ihc_data) |>
  filter(lineage %in% comparable_lineages) |>
  select(patient_id, lineage, ihc_frac = frac)

# Complete the patient x lineage grid: a lineage a patient simply lacks must read
# as 0 (present), not a missing row. Otherwise the inner_join below silently drops
# that patient from that lineage's panel — the "plot never shows all patients" bug.
ihc_lin <- expand.grid(patient_id = unique(slide_key(ihc_data$patient_id)),
                       lineage    = comparable_lineages,
                       stringsAsFactors = FALSE) |>
  left_join(lin_raw, by = c("patient_id", "lineage")) |>
  mutate(ihc_frac = ifelse(is.na(ihc_frac), 0, ihc_frac))

cat("IHC patients with lineage fractions:", dplyr::n_distinct(ihc_lin$patient_id), "\n")
```

## Method vs IHC — per-lineage concordance

For each method, collapse its cell types to the four comparable lineages (sum
where a method splits a lineage, e.g. quanTIseq's M1/M2 do not map here), join to
the IHC fraction on patient + lineage, and correlate across patients.

```{r concordance}
paired <- NULL
if (!is.null(deconv_long) && nrow(deconv_long) > 0) {
  deconv_lin <- deconv_long |>
    filter(!is.na(lineage)) |>
    group_by(method, patient_id, lineage) |>
    summarise(score = sum(score), .groups = "drop")

  paired <- deconv_lin |>
    inner_join(ihc_lin, by = c("patient_id", "lineage")) |>
    filter(is.finite(score), is.finite(ihc_frac))
  cat("paired (method x lineage x patient) rows:", nrow(paired), "\n")
}
```

```{r concordance-table}
if (!is.null(paired) && nrow(paired) > 0) {
  summary_tbl <- paired |>
    group_by(method, lineage) |>
    group_modify(~ paired_spearman(.x$score, .x$ihc_frac)) |>
    ungroup() |>
    arrange(lineage, desc(rho))
  knitr::kable(summary_tbl, digits = 3,
               caption = "Spearman IHC vs deconvolution, per method and lineage")

  # Which method agrees best overall (mean rho across lineages).
  method_rank <- summary_tbl |>
    group_by(method) |>
    summarise(mean_rho = mean(rho, na.rm = TRUE),
              n_lineages = sum(is.finite(rho)), .groups = "drop") |>
    arrange(desc(mean_rho))
  knitr::kable(method_rank, digits = 3,
               caption = "Method ranking by mean Spearman across lineages")
}
```

```{r concordance-plot}
if (!is.null(paired) && nrow(paired) > 0) {
  ggplot(paired, aes(ihc_frac, score)) +
    geom_point(alpha = 0.6, size = 1.8) +
    geom_smooth(method = "lm", se = FALSE, colour = "steelblue", formula = y ~ x) +
    facet_grid(method ~ lineage, scales = "free") +
    labs(title = "IHC lineage fraction vs bulk-RNA deconvolution score",
         x = "IHC fraction of all cells", y = "deconvolution score") +
    theme_bw(base_size = 10)
}
```

## Notes

- Absolute-fraction methods (quanTIseq, EPIC) are the fair apples-to-apples
  comparison; MCPcounter/xCell scores are relative, so only their *rank*
  correlation with IHC is meaningful, not the slope.
- Only CD8T / CD4T / Treg / NK are compared — the IHC panel has no macrophage or
  B-cell phenotype, so those deconvolution outputs have no phenotype-level IHC
  counterpart (see `bulkRna.Rmd` for a marker-level CD163/CD14 check instead).
- Cross-patient Spearman needs ≥ 3 shared patients per lineage; sparser cells
  return `NA` rather than a spurious correlation.
