---
title: "exclusivity"
author: "sceriff0"
date: "2026-07-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
})

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6)

invisible(capture.output(suppressMessages(source(here("code", "test.R")))))
source(here("code", "validation_helpers.R"))
theme_set(theme_paper)   # house style (mirage benchmarks)
CAP <- "ihc_method internal QC · lineage leakage & exclusivity"
```

## Purpose

**Lineage leakage & exclusivity** — an internal QC (only `ihc_data`) that reads the
**segmentation and staining**: cell lineages biology keeps separate should not
co-stain in the same cell. Excess co-positivity is a direct signal of **optical
spillover between touching cells** or **mis-segmentation** (one mask capturing
two cells). Split out of [benchmarks.html](benchmarks.html) so the phenotyping,
channel and morphology checks stay separate from the spillover checks.

Two views: phenotype-level leakage, then a literature-backed panel of
mutually-exclusive marker pairs — each with its rare biological exception noted so
the QC flags *excess* co-positivity, not the known baseline.

## 1. Phenotype-level leakage

CD45⁺ cells among tumour-typed cells, and PANCK⁺ cells among immune-typed cells —
both should be ~0 if the phenotyping respects the tumour/immune divide.

```{r leakage}
immune_ph <- c("Immune", "T helper", "T cytotoxic", "Activated T cytotoxic",
               "CD4+ Treg", "CD8+ T reg", "Natural Killer", "Activated Natural Killer")

excl <- ihc_data |>
  mutate(patient_id  = slide_key(patient_id),
         is_tumor_ph = str_detect(replace_na(phenotype_clean, ""), "Tumor"),
         is_immune_ph = phenotype_clean %in% immune_ph,
         panck_pos = is_pos(PANCK_sign),
         cd45_pos  = is_pos(CD45_sign))

spill <- excl |>
  group_by(patient_id) |>
  summarise(
    n_cells            = n(),
    cd45pos_in_tumorph = if (any(is_tumor_ph))  mean(cd45_pos[is_tumor_ph],  na.rm = TRUE) else NA_real_,
    panckpos_in_immune = if (any(is_immune_ph)) mean(panck_pos[is_immune_ph], na.rm = TRUE) else NA_real_,
    .groups = "drop"
  )
knitr::kable(spill, digits = 3,
             caption = "Per-patient phenotype-level leakage (want ~0)")
```

```{r leakage-plot, fig.height=4.5}
spill |>
  pivot_longer(c(cd45pos_in_tumorph, panckpos_in_immune),
               names_to = "leak", values_to = "rate") |>
  filter(!is.na(rate)) |>
  mutate(leak = c(cd45pos_in_tumorph = "CD45+ among tumour-typed",
                  panckpos_in_immune = "PANCK+ among immune-typed")[leak]) |>
  ggplot(aes(reorder(patient_id, rate), rate, fill = leak)) +
    geom_col(position = position_dodge(width = .7), width = .65) +
    coord_flip() +
    scale_fill_manual(values = oi[c(1, 2)], name = NULL) +
    labs(title = "Phenotype-level leakage per patient",
         subtitle = "Fraction of a lineage's cells calling the opposite lineage's marker — lower is better.",
         x = "patient", y = "fraction of cells", caption = CAP)
```

## 2. Marker-pair exclusivity panel

Each pair is two markers biology assigns to distinct lineages. `p_b_given_a` /
`p_a_given_b` are the conditional co-positivity rates — the fraction of one
marker's positive cells that also call the other. These should sit near the
"expected co-positivity" column (the rare true double-positive population); values
well above it flag spillover for that pair.

```{r exclusivity-pairs}
# Per-cell marker positivity matrix (one logical column per marker).
sign_pos <- ihc_data |>
  transmute(patient_id = slide_key(patient_id),
            across(all_of(paste0(ihc_markers, "_sign")), is_pos, .names = "{.col}")) |>
  rename_with(~ sub("_sign$", "", .x), ends_with("_sign"))

# Literature-backed mutually-exclusive lineage marker pairs (+ known exception).
exclusivity_pairs <- tibble::tribble(
  ~a,      ~b,      ~pair_label,                              ~expected_copos, ~reference,
  "PANCK", "CD45",  "epithelial/tumour vs leukocyte",         "~0 (rare hybrid CTCs)",     "Nandedkar 1998",
  "CD4",   "CD8",   "helper vs cytotoxic T",                  "~3% peripheral DP-T",       "Overgaard 2023 (Biomedicines)",
  "CD3",   "CD56",  "T cell vs NK cell",                      "rare CD3+CD56+ NKT",        "std NK/NKT gating",
  "CD3",   "CD14",  "lymphoid vs monocyte/macrophage",        "~0 distinct lineages",      "flow-panel convention",
  "CD3",   "CD163", "lymphoid vs M2 macrophage",              "~0 (lymphocytes CD163-)",   "Roberts 2004",
  "CD45",  "SMA",   "leukocyte vs smooth-muscle/myofibroblast","~0 distinct lineages",     "stromal-marker convention"
) |>
  # keep only pairs whose markers are both present
  filter(a %in% names(sign_pos), b %in% names(sign_pos))

pair_stats <- exclusivity_pairs |>
  rowwise() |>
  mutate(
    n_a          = sum(sign_pos[[a]], na.rm = TRUE),
    n_b          = sum(sign_pos[[b]], na.rm = TRUE),
    n_both       = sum(sign_pos[[a]] & sign_pos[[b]], na.rm = TRUE),
    pct_dp_all   = n_both / nrow(sign_pos),
    p_b_given_a  = ifelse(n_a > 0, n_both / n_a, NA_real_),
    p_a_given_b  = ifelse(n_b > 0, n_both / n_b, NA_real_)
  ) |>
  ungroup() |>
  select(pair = pair_label, a, b, expected_copos, reference,
         n_both, pct_dp_all, p_b_given_a, p_a_given_b)

knitr::kable(pair_stats, digits = 4,
             caption = "Mutually-exclusive marker pairs: observed vs expected co-positivity")
```

```{r exclusivity-summary, fig.height=4.5}
# The worst-case conditional co-positivity per pair (max of the two directions).
pair_stats |>
  mutate(max_cond = pmax(p_b_given_a, p_a_given_b, na.rm = TRUE),
         pair = sprintf("%s / %s", a, b)) |>
  filter(is.finite(max_cond)) |>
  ggplot(aes(reorder(pair, max_cond), max_cond)) +
    geom_col(fill = oi[2], width = .65) +
    coord_flip() +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    labs(title = "Worst-direction co-positivity per exclusive pair",
         subtitle = "max( P(B+|A+), P(A+|B+) ) — high bars = the pair most likely showing spillover.",
         x = NULL, y = "conditional co-positivity", caption = CAP)
```

```{r exclusivity-heatmap}
per_pat <- exclusivity_pairs |>
  select(a, b, pair_label) |>
  purrr::pmap_dfr(function(a, b, pair_label) {
    sign_pos |>
      group_by(patient_id) |>
      summarise(dp = mean(.data[[a]] & .data[[b]], na.rm = TRUE), .groups = "drop") |>
      mutate(pair = pair_label)
  })

if (nrow(per_pat) > 0) {
  ggplot(per_pat, aes(patient_id, pair, fill = dp)) +
    geom_tile(colour = "white") +
    scale_fill_viridis_c(option = "magma", direction = -1,
                         labels = scales::percent_format(accuracy = .1),
                         name = "double-pos\nfraction") +
    labs(title = "Marker-pair co-positivity per patient",
         subtitle = "Bright = high double-positive rate = likely spillover in that patient.",
         x = "patient", y = NULL, caption = CAP) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          panel.grid = element_blank())
}
```

**Pairs deliberately excluded** (this panel's phenotypes acknowledge their
overlap, so co-positivity is expected, not an artifact): PANCK/VIMENTIN (the
`VIM+Tumor` EMT phenotype), and CD8/FOXP3 (the `CD8+ T reg` phenotype).

**References for the exclusivity claims:**

- PANCK/CD45 — Nandedkar MA *et al.* (1998), CD45 immunoreactivity as a diagnostic
  pitfall in carcinoma: keratin and CD45 are normally exclusive.
  [PubMed 9872653](https://pubmed.ncbi.nlm.nih.gov/9872653/)
- CD4/CD8 — "Legend or Truth: Mature CD4⁺CD8⁺ Double-Positive T Cells in the
  Periphery," *Biomedicines* 2023: DP-T are a real but ~3% minority.
  [MDPI](https://www.mdpi.com/2227-9059/11/10/2702)
- CD3/CD56 — CD3⁻CD56⁺ = NK, CD3⁺CD56⁺ = rare NKT; standard immunophenotyping.
- CD3/CD163 — Roberts ES *et al.* (2004): lymphocytes are CD163-negative.
  [PMC2258269](https://pmc.ncbi.nlm.nih.gov/articles/PMC2258269/)

## Notes

- Uses only `ihc_data` — no external reference, so it holds for every imaged slide.
- A high co-positivity for an *exclusive* pair points at segmentation/optical
  spillover between neighbouring cells, not biology — cross-check the offending
  patient in the per-patient heatmap and the channel QC in
  [benchmarks.html](benchmarks.html).
