---
title: "exclusivity"
author: "sceriff0"
date: "2026-07-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
})

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6)

invisible(capture.output(suppressMessages(source(here("code", "test.R")))))
source(here("code", "validation_helpers.R"))
theme_set(theme_paper)   # house style (mirage benchmarks)
CAP <- "ihc_method internal QC · lineage leakage & exclusivity"
```

## Starting data and notation

Single input: `ihc_data`, the FlowPath single cells deduplicated to **one copy per
patient** (whole slide). Let $C_p$ be the cells of patient $p$ and
$C=\bigcup_p C_p$ the pooled cohort. All patient ids are mapped to slide space by
`slide_key()` (strip non-digits, apply the typo fix `15879`$\to$`15897`).

For a marker $m$ carrying a `<m>_sign` column, positivity is
$$
m^+(c)=\mathbf{1}\!\left[\texttt{is\_pos}(\texttt{<m>\_sign}(c))\right],\qquad
\texttt{is\_pos}(v)\iff v\in\{+,\text{pos},\text{positive},\text{yes},\text{true},1\}.
$$
`phenotype_clean` (the parenthetical label in `phenotype`) is the per-cell type;
`lineage(c)` collapses it via `phenotype_lineage`. This report defines its two
phenotype membership sets directly from `phenotype_clean`, **not** through
`lineage()`:
$$
\text{tumour-ph}(c)\iff \texttt{phenotype\_clean}(c)\ \text{contains}\ \texttt{"Tumor"},\qquad
\text{immune-ph}(c)\iff \texttt{phenotype\_clean}(c)\in\Phi_\text{imm},
$$
with the literal set $\Phi_\text{imm}=\{$`Immune`, `T helper`, `T cytotoxic`,
`Activated T cytotoxic`, `CD4+ Treg`, `CD8+ T reg`, `Natural Killer`,
`Activated Natural Killer`$\}$ (defined in the chunk below; excludes Stroma /
Unknown / Tumor). Missing `phenotype_clean` is treated as the empty string, so it
matches neither set.

## 1. Phenotype-level leakage

Per patient $p$, the fraction of one lineage's phenotyped cells that call the
opposite lineage's marker:
$$
\text{cd45pos\_in\_tumorph}_p=
\frac{\bigl|\{c\in C_p:\text{tumour-ph}(c)\ \wedge\ \text{CD45}^+(c)\}\bigr|}
     {\bigl|\{c\in C_p:\text{tumour-ph}(c)\}\bigr|},
$$
$$
\text{panckpos\_in\_immune}_p=
\frac{\bigl|\{c\in C_p:\text{immune-ph}(c)\ \wedge\ \text{PANCK}^+(c)\}\bigr|}
     {\bigl|\{c\in C_p:\text{immune-ph}(c)\}\bigr|},
$$
with $n\_\text{cells}_p=|C_p|$. Each ratio is `NA` when its patient has no cell in
the corresponding phenotype set (`any(...)` guard); positivity is never `NA`.

```{r leakage}
immune_ph <- c("Immune", "T helper", "T cytotoxic", "Activated T cytotoxic",
               "CD4+ Treg", "CD8+ T reg", "Natural Killer", "Activated Natural Killer")

excl <- ihc_data |>
  mutate(patient_id  = slide_key(patient_id),
         is_tumor_ph = str_detect(replace_na(phenotype_clean, ""), "Tumor"),
         is_immune_ph = phenotype_clean %in% immune_ph,
         panck_pos = is_pos(PANCK_sign),
         cd45_pos  = is_pos(CD45_sign))

spill <- excl |>
  group_by(patient_id) |>
  summarise(
    n_cells            = n(),
    cd45pos_in_tumorph = if (any(is_tumor_ph))  mean(cd45_pos[is_tumor_ph],  na.rm = TRUE) else NA_real_,
    panckpos_in_immune = if (any(is_immune_ph)) mean(panck_pos[is_immune_ph], na.rm = TRUE) else NA_real_,
    .groups = "drop"
  )
knitr::kable(spill, digits = 3,
             caption = "Per-patient phenotype-level leakage (want ~0)")
```

Bars are the two ratios above, per patient, dropping `NA` entries.

```{r leakage-plot, fig.height=4.5}
spill |>
  pivot_longer(c(cd45pos_in_tumorph, panckpos_in_immune),
               names_to = "leak", values_to = "rate") |>
  filter(!is.na(rate)) |>
  mutate(leak = c(cd45pos_in_tumorph = "CD45+ among tumour-typed",
                  panckpos_in_immune = "PANCK+ among immune-typed")[leak]) |>
  ggplot(aes(reorder(patient_id, rate), rate, fill = leak)) +
    geom_col(position = position_dodge(width = .7), width = .65) +
    coord_flip() +
    scale_fill_manual(values = oi[c(1, 2)], name = NULL) +
    labs(title = "Phenotype-level leakage per patient",
         subtitle = "Fraction of a lineage's cells calling the opposite lineage's marker — lower is better.",
         x = "patient", y = "fraction of cells", caption = CAP)
```

## 2. Marker-pair exclusivity panel

`sign_pos` is the per-cell positivity matrix: one logical column $m^+$ per marker
$m\in$`ihc_markers`, obtained by applying `is_pos` to every `<m>_sign` column and
stripping the `_sign` suffix. Each row is a cell $c\in C$; `patient_id` is
`slide_key`-normalised.

For a candidate pair $(a,b)$ retained only if both columns exist in `sign_pos`,
all counts are **pooled over the whole cohort** $C$:
$$
n_a=\sum_{c\in C}a^+(c),\qquad
n_b=\sum_{c\in C}b^+(c),\qquad
n_\text{both}=\sum_{c\in C}a^+(c)\,b^+(c).
$$
$$
\text{pct\_dp\_all}=\frac{n_\text{both}}{|C|},\qquad
p(b\!\mid\!a)=\frac{n_\text{both}}{n_a},\qquad
p(a\!\mid\!b)=\frac{n_\text{both}}{n_b},
$$
with the two conditionals set to `NA` when their denominator ($n_a$ or $n_b$) is
$0$. `pct_dp_all` uses the all-cell denominator $|C|$; the conditionals use the
single-marker denominators.

```{r exclusivity-pairs}
# Per-cell marker positivity matrix (one logical column per marker).
sign_pos <- ihc_data |>
  transmute(patient_id = slide_key(patient_id),
            across(all_of(paste0(ihc_markers, "_sign")), is_pos, .names = "{.col}")) |>
  rename_with(~ sub("_sign$", "", .x), ends_with("_sign"))

# Literature-backed mutually-exclusive lineage marker pairs (+ known exception).
exclusivity_pairs <- tibble::tribble(
  ~a,      ~b,      ~pair_label,                              ~expected_copos, ~reference,
  "PANCK", "CD45",  "epithelial/tumour vs leukocyte",         "~0 (rare hybrid CTCs)",     "Nandedkar 1998",
  "CD4",   "CD8",   "helper vs cytotoxic T",                  "~3% peripheral DP-T",       "Overgaard 2023 (Biomedicines)",
  "CD3",   "CD56",  "T cell vs NK cell",                      "rare CD3+CD56+ NKT",        "std NK/NKT gating",
  "CD3",   "CD14",  "lymphoid vs monocyte/macrophage",        "~0 distinct lineages",      "flow-panel convention",
  "CD3",   "CD163", "lymphoid vs M2 macrophage",              "~0 (lymphocytes CD163-)",   "Roberts 2004",
  "CD45",  "SMA",   "leukocyte vs smooth-muscle/myofibroblast","~0 distinct lineages",     "stromal-marker convention"
) |>
  # keep only pairs whose markers are both present
  filter(a %in% names(sign_pos), b %in% names(sign_pos))

pair_stats <- exclusivity_pairs |>
  rowwise() |>
  mutate(
    n_a          = sum(sign_pos[[a]], na.rm = TRUE),
    n_b          = sum(sign_pos[[b]], na.rm = TRUE),
    n_both       = sum(sign_pos[[a]] & sign_pos[[b]], na.rm = TRUE),
    pct_dp_all   = n_both / nrow(sign_pos),
    p_b_given_a  = ifelse(n_a > 0, n_both / n_a, NA_real_),
    p_a_given_b  = ifelse(n_b > 0, n_both / n_b, NA_real_)
  ) |>
  ungroup() |>
  select(pair = pair_label, a, b, expected_copos, reference,
         n_both, pct_dp_all, p_b_given_a, p_a_given_b)

knitr::kable(pair_stats, digits = 4,
             caption = "Mutually-exclusive marker pairs: observed vs expected co-positivity")
```

Worst-direction co-positivity per pair:
$$
\text{max\_cond}=\max\bigl(p(b\!\mid\!a),\,p(a\!\mid\!b)\bigr),
$$
plotted for the pairs with a finite value.

```{r exclusivity-summary, fig.height=4.5}
# The worst-case conditional co-positivity per pair (max of the two directions).
pair_stats |>
  mutate(max_cond = pmax(p_b_given_a, p_a_given_b, na.rm = TRUE),
         pair = sprintf("%s / %s", a, b)) |>
  filter(is.finite(max_cond)) |>
  ggplot(aes(reorder(pair, max_cond), max_cond)) +
    geom_col(fill = oi[2], width = .65) +
    coord_flip() +
    scale_y_continuous(labels = scales::percent_format(accuracy = 1)) +
    labs(title = "Worst-direction co-positivity per exclusive pair",
         subtitle = "max( P(B+|A+), P(A+|B+) ) — high bars = the pair most likely showing spillover.",
         x = NULL, y = "conditional co-positivity", caption = CAP)
```

Per-patient double-positive fraction, one value per (patient, pair), over **all**
of that patient's cells:
$$
\text{dp}_{p,(a,b)}=\frac{1}{|C_p|}\sum_{c\in C_p}a^+(c)\,b^+(c).
$$
This denominator is $|C_p|$ (all cells), i.e. the per-patient analogue of
`pct_dp_all`, not of the conditional rates above.

```{r exclusivity-heatmap}
per_pat <- exclusivity_pairs |>
  select(a, b, pair_label) |>
  purrr::pmap_dfr(function(a, b, pair_label) {
    sign_pos |>
      group_by(patient_id) |>
      summarise(dp = mean(.data[[a]] & .data[[b]], na.rm = TRUE), .groups = "drop") |>
      mutate(pair = pair_label)
  })

if (nrow(per_pat) > 0) {
  ggplot(per_pat, aes(patient_id, pair, fill = dp)) +
    geom_tile(colour = "white") +
    scale_fill_viridis_c(option = "magma", direction = -1,
                         labels = scales::percent_format(accuracy = .1),
                         name = "double-pos\nfraction") +
    labs(title = "Marker-pair co-positivity per patient",
         subtitle = "Bright = high double-positive rate = likely spillover in that patient.",
         x = "patient", y = NULL, caption = CAP) +
    theme(axis.text.x = element_text(angle = 45, hjust = 1),
          panel.grid = element_blank())
}
```

Pairs held out of the tribble because their phenotypes are defined to overlap
(co-positivity is expected, not an artifact): PANCK/VIMENTIN (the `VIM+Tumor` EMT
phenotype) and CD8/FOXP3 (the `CD8+ T reg` phenotype).

## Notes

- Single source `ihc_data` (whole slide); every metric is a count ratio, no
  external reference.
- Denominator audit: leakage rates divide by phenotype-set size within $C_p$;
  `pct_dp_all` and heatmap `dp` divide by all cells ($|C|$ / $|C_p|$);
  `p_b_given_a` / `p_a_given_b` divide by the single-marker positive counts
  ($n_a$ / $n_b$). The conditional rates and pair counts are pooled cohort-wide,
  whereas leakage and the heatmap are per patient.
- Leakage membership uses `phenotype_clean` text and the literal $\Phi_\text{imm}$
  list, not the `phenotype_lineage` collapse.
