---
title: "ihc_qc"
author: "sceriff0"
date: "2026-07-05"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: tidyverse + here (in the lockfile). Optional: diptest for Hartigan's dip
# test (a bimodality-coefficient fallback runs if it is absent).
#   renv::install("diptest"); renv::snapshot()
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
})

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6)

invisible(capture.output(suppressMessages(source(here("code", "test.R")))))
source(here("code", "validation_helpers.R"))
theme_set(theme_paper)   # house style (mirage benchmarks)
CAP <- "ihc_method internal QC · FlowPath single-cell IHC"
```

## Starting data and notation

Every value below is derived from a single input, with **no external reference**:

- `ihc_data` — FlowPath single cells, deduplicated to **one copy per patient**
  (whole slide). Each cell carries `phenotype_clean` (the parenthetical label in
  `phenotype`) and, for each marker $m$, the triple
  $(m\texttt{\_raw},\,m\texttt{\_zscore},\,m\texttt{\_sign})$.

Notation used throughout. Let $C_p$ be all cells of patient $p$ in `ihc_data`, and
$C=\bigcup_p C_p$ the whole set of imaged cells. `ihc_marker_long(ihc_data)`
reshapes $C$ to **one row per $(c,m)$ pair** over the markers $m\in M$
(`ihc_markers` = `marker_gene_map$marker`, excluding `wrongL1CAM`/DAPI), carrying
per pair the raw value $r_{c,m}$, the z-score $z_{c,m}$, the sign string
$s_{c,m}$, and the boolean

$$
\text{pos}_{c,m}=\texttt{is\_pos}(s_{c,m})=\mathbf{1}\!\left[\,\operatorname{lower}(\operatorname{trim}(s_{c,m}))\in\{+,\text{pos},\text{positive},\text{yes},\text{true},1\}\,\right].
$$

`phenotype_clean` is the cell type $t(c)$. All patient ids are mapped to the
shared slide space by `slide_key()`. The reports restricts to
**typed** cells: $t(c)\notin\{\texttt{NA},\text{"Unknown"}\}$.

```{r marker-long}
mk_long <- ihc_marker_long(ihc_data) |>
  filter(!is.na(phenotype_clean), phenotype_clean != "Unknown")
cat("cells:", dplyr::n_distinct(mk_long$cell_id),
    "| markers:", dplyr::n_distinct(mk_long$marker),
    "| phenotypes:", dplyr::n_distinct(mk_long$phenotype_clean), "\n")
```

## 1. Phenotype ↔ marker concordance

For each phenotype $t$ and marker $m$, over the typed pairs
$\{c : t(c)=t\}$:

$$
\text{mean\_z}_{t,m}=\operatorname{mean}_{c:\,t(c)=t} z_{c,m},\qquad
\text{pct\_pos}_{t,m}=\operatorname{mean}_{c:\,t(c)=t}\text{pos}_{c,m},
$$

both computed with `na.rm = TRUE`. The heatmap fills each $(m,t)$ tile with
$\text{mean\_z}_{t,m}$ (diverging scale centred at $0$).

```{r concordance-matrix}
conc <- mk_long |>
  group_by(phenotype_clean, marker) |>
  summarise(mean_z = mean(zscore, na.rm = TRUE),
            pct_pos = mean(pos,  na.rm = TRUE), .groups = "drop")

ggplot(conc, aes(marker, phenotype_clean, fill = mean_z)) +
  geom_tile(colour = "white") +
  scale_fill_gradient2(low = "#2166ac", mid = "white", high = "#b2182b",
                       midpoint = 0, name = "mean z") +
  labs(title = "Phenotype vs marker: mean z-score",
       subtitle = "Expect each phenotype to peak on its defining marker(s); off-diagonal signal = mislabelled cells.",
       x = "marker", y = "phenotype (phenotype_clean)", caption = CAP) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        panel.grid = element_blank())
```

**Defining-marker check.** `expected` is a fixed table $D$ giving the defining
marker $d(t)$ for each phenotype. Within each phenotype the markers are ranked by
descending mean z-score,

$$
\text{rank\_z}_{t,m}=\operatorname{rank}\!\big(-\text{mean\_z}_{t,\cdot}\big)_m
\quad(\text{rank }1=\text{largest mean\_z}),
$$

and the row for $m=d(t)$ is kept (`inner_join` on $D$). The reported
`rank_of_defining` is $\text{rank\_z}_{t,d(t)}$ and

$$
\text{pass}_t=\mathbf{1}\!\left[\text{rank\_z}_{t,d(t)}\le 2\right]
$$

(defining marker is among the phenotype's top-2 mean-z signals). `mean_z` is
rounded to 2 dp; rows are sorted by descending `rank_of_defining`.

```{r concordance-expected}
# Encode the expected defining marker(s) per phenotype and check it is the top
# (or near-top) signal. Flags phenotypes whose defining marker is not their max.
expected <- tibble::tribble(
  ~phenotype_clean,           ~marker,
  "PANCK+Tumor",              "PANCK",
  "VIM+Tumor",                "VIMENTIN",
  "T helper",                 "CD4",
  "T cytotoxic",              "CD8",
  "Activated T cytotoxic",    "CD8",
  "CD4+ Treg",                "FOXP3",
  "CD8+ T reg",               "FOXP3",
  "Natural Killer",           "CD56",
  "Activated Natural Killer", "CD56",
  "Immune",                   "CD45",
  "Stroma",                   "SMA"
)

check <- conc |>
  group_by(phenotype_clean) |>
  mutate(rank_z = rank(-mean_z)) |>
  inner_join(expected, by = c("phenotype_clean", "marker")) |>
  transmute(phenotype_clean, defining_marker = marker,
            mean_z = round(mean_z, 2), rank_of_defining = as.integer(rank_z),
            pass = rank_z <= 2) |>
  arrange(desc(rank_of_defining))
knitr::kable(check, caption = "Is each phenotype's defining marker among its top-2 signals?")
```

## 2. Per-channel usability

Per marker $m$, over all cells with finite $z_{c,m}$ (typed and untyped —
`ihc_marker_long(ihc_data)` is re-derived here without the phenotype filter).

**Bimodality.** Sarle's bimodality coefficient on the z-scores $\{z_{c,m}\}$
($n$ finite values, sample mean $\bar z$, sample sd $\sigma$):

$$
g=\frac{1}{n}\sum\Big(\frac{z-\bar z}{\sigma}\Big)^3,\quad
k=\frac{1}{n}\sum\Big(\frac{z-\bar z}{\sigma}\Big)^4-3,\quad
\text{BC}=\frac{g^2+1}{\,k+\dfrac{3(n-1)^2}{(n-2)(n-3)}\,},
$$

returning `NA` if $n<4$ or $\sigma=0$. `dip_p` is Hartigan's dip-test p-value
(`diptest::dip.test`) when the package is installed and $n\ge 4$, else `NA`.

**Sign-vs-zscore separability.** With the positive/negative split from
$\text{pos}_{c,m}$:

$$
\text{neg\_max}_m=\max_{c:\,\lnot\text{pos}} z_{c,m},\qquad
\text{pos\_min}_m=\min_{c:\,\text{pos}} z_{c,m},
$$

$$
\text{separable}_m=\mathbf{1}\!\left[\text{pos\_min}_m>\text{neg\_max}_m\right],\quad
\text{cutoff}_m=\begin{cases}\tfrac{\text{neg\_max}_m+\text{pos\_min}_m}{2}&\text{separable}\\[2pt]\texttt{NA}&\text{else,}\end{cases}
$$

$$
\text{likely\_bimodal}_m=\mathbf{1}\!\left[\text{BC}_m>0.555\ \lor\ (\,\text{dip\_p}_m<0.05\,)\right].
$$

Rows are sorted by descending BC.

```{r channel-qc}
bimodality_coef <- function(x) {                       # Sarle's BC
  x <- x[is.finite(x)]; n <- length(x)
  if (n < 4) return(NA_real_)
  m <- mean(x); s <- sd(x); if (s == 0) return(NA_real_)
  g <- mean(((x - m) / s)^3)                           # skewness
  k <- mean(((x - m) / s)^4) - 3                        # excess kurtosis
  (g^2 + 1) / (k + 3 * (n - 1)^2 / ((n - 2) * (n - 3)))
}
dip_p <- function(x) {
  x <- x[is.finite(x)]
  if (!requireNamespace("diptest", quietly = TRUE) || length(x) < 4) return(NA_real_)
  suppressWarnings(diptest::dip.test(x)$p.value)
}

channel_qc <- ihc_marker_long(ihc_data) |>
  filter(!is.na(zscore)) |>
  group_by(marker) |>
  summarise(
    n        = n(),
    bc       = bimodality_coef(zscore),
    dip_p    = dip_p(zscore),
    neg_max  = suppressWarnings(max(zscore[!pos], na.rm = TRUE)),
    pos_min  = suppressWarnings(min(zscore[pos],  na.rm = TRUE)),
    .groups  = "drop"
  ) |>
  mutate(
    separable = is.finite(neg_max) & is.finite(pos_min) & pos_min > neg_max,
    cutoff    = if_else(separable, (neg_max + pos_min) / 2, NA_real_),
    likely_bimodal = bc > 0.555 | (!is.na(dip_p) & dip_p < 0.05)
  ) |>
  arrange(desc(bc))
knitr::kable(channel_qc, digits = 3,
             caption = "Per-channel bimodality and sign-vs-zscore separability")
```

Density of $z_{c,m}$ per marker (finite $z$ and non-`NA` sign only), split by the
call $\text{pos}_{c,m}$ (`sign +` / `sign -`); facets are free-scaled.

```{r channel-qc-plot, fig.height=8}
ihc_marker_long(ihc_data) |>
  filter(!is.na(zscore), !is.na(sign)) |>
  mutate(call = if_else(pos, "sign +", "sign -")) |>
  ggplot(aes(zscore, fill = call)) +
    geom_density(alpha = 0.55, colour = NA) +
    facet_wrap(~ marker, scales = "free") +
    scale_fill_manual(values = c("sign +" = oi[1], "sign -" = oi[2]), name = NULL) +
    labs(title = "Marker z-score by sign call",
         subtitle = "Clean channels show two separated humps; failed channels overlap or are unimodal.",
         x = "z-score", y = "density", caption = CAP)
```

## 3. Morphology by phenotype

Morphology features are the intersection of
$\{\texttt{area},\texttt{perimeter},\texttt{eccentricity},\texttt{solidity}\}$
with `names(ihc_data)`, plus circularity when both `area` and `perimeter` exist:

$$
\text{circularity}(c)=\frac{4\pi\,\texttt{area}(c)}{\texttt{perimeter}(c)^2}.
$$

Over typed cells, each feature is pivoted long and filtered to finite values; the
plot is a per-phenotype boxplot (phenotypes ordered by median feature value,
outliers hidden, facets free-scaled).

```{r morphology}
morph_feats <- intersect(c("area", "perimeter", "eccentricity", "solidity"),
                         names(ihc_data))

morph <- ihc_data |>
  filter(!is.na(phenotype_clean), phenotype_clean != "Unknown") |>
  mutate(circularity = if (all(c("area", "perimeter") %in% names(ihc_data)))
           4 * pi * area / perimeter^2 else NA_real_) |>
  select(phenotype_clean, any_of(c(morph_feats, "circularity"))) |>
  pivot_longer(-phenotype_clean, names_to = "feature", values_to = "value") |>
  filter(is.finite(value))

ggplot(morph, aes(reorder(phenotype_clean, value, FUN = median), value,
                  fill = phenotype_clean)) +
  geom_boxplot(outlier.shape = NA, width = .6, colour = "grey30") +
  facet_wrap(~ feature, scales = "free") +
  coord_flip() +
  scale_fill_viridis_d(option = "D", guide = "none") +
  labs(title = "Cell morphology by phenotype",
       subtitle = "Shapes should track cell type: tumour largest, lymphocytes small/round, stroma elongated.",
       x = NULL, y = NULL, caption = CAP)
```

## Notes

- All three checks use only `ihc_data`; no external reference is joined.
- `mean` reductions carry `na.rm = TRUE`; `neg_max`/`pos_min` return non-finite
  (`±Inf`) when a marker has no negative (resp. positive) cell, which drives
  `separable = FALSE` via the `is.finite` guards.
- `separable = FALSE` means `_sign` is not a pure threshold on `_zscore` (the
  `_sign+` and `_sign−` z-score ranges overlap — likely thresholded on `_raw`);
  its `cutoff` is then `NA`.
- `likely_bimodal` uses an inclusive-or of BC $>0.555$ and dip-p $<0.05$; if
  `diptest` is absent the dip-p term is `NA` and only BC decides.
