---
title: "Tumour border / invasive-margin immunity"
author: "sceriff0"
date: "2026-07-06"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---

```{r setup, include=FALSE}
# renv: sf + jsonlite for the geojson buffers, plus tidyverse + here.
#   renv::install(c("sf", "jsonlite")); renv::snapshot()
suppressPackageStartupMessages({
  library(tidyverse)
  library(here)
  library(sf)
})

knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 9, fig.height = 6)

# Loaders (dds, clinical_data, neoplastic_data, counts_data, ihc_data) and the
# shared validation helpers. test.R prints colnames() on load; silence it.
invisible(capture.output(suppressMessages(source(here("code", "test.R")))))
source(here("code", "validation_helpers.R"))
```

## Starting data and geometry

Inputs: `ihc_data` (whole-slide FlowPath cells, one copy per patient, with
`centroid_x`/`centroid_y` in **microns**), the pathologist annotation **geojson**
polygons (in **pixels**), and the clinical `HOT score` / `Immuno-phenotype`.

Unlike the inside-the-annotation checks, membership here needs real geometry (a
distance band cannot come from the binary `Out_of_annotation` flag), so this report
keeps `sf`. Cells are placed in the polygon pixel frame by a **fixed** conversion:

```
x_px = centroid_x / 0.325
y_px = centroid_y / 0.325           (0.325 microns per pixel)
```

So 1 pixel = 0.325 microns, which gives the two conversions used below:

```
threshold d microns  ->  d / 0.325 pixels
pixel area A         ->  A * 0.325^2 / 1e6  square millimetres
```

## Margin and core regions

For a core polygon P (the union of a patient's annotations, or each single
annotation) and a half-width `d` microns, with `d_px = d / 0.325`:

```
margin band = (P grown by d_px)  minus  (P shrunk by d_px)
deep core   =  P shrunk by d_px
```

(grow/shrink are `st_buffer` by +/- `d_px`; `margin_regions`.) If `d_px` exceeds
the core's inradius the shrunk core is empty, and the margin band falls back to the
full grown buffer. Thresholds analysed: `d` = 100, 250, 500 microns (500 is the
consensus-Immunoscore invasive-margin half-width).

```{r periphery-metrics}
THRESHOLDS_UM <- c(100, 250, 500)
UM_PER_PX     <- 0.325

annots <- load_annotations(patient_ids = neoplastic_data$SAMPLE)

peri_union   <- ihc_periphery_metrics(ihc_data, annots, scope = "union",
                                       thresholds_um = THRESHOLDS_UM, um_per_px = UM_PER_PX)
peri_per_ann <- ihc_periphery_metrics(ihc_data, annots, scope = "per_annotation",
                                       thresholds_um = THRESHOLDS_UM, um_per_px = UM_PER_PX)

cat("union rows:", nrow(peri_union),
    "| patients:", dplyr::n_distinct(peri_union$patient_id),
    "| per-annotation rows:", nrow(peri_per_ann), "\n")
```

## Per-region quantities

For the cells inside a region (`st_within`), `region_ratios_area()` returns the
membership counts (as in `clinical_data.Rmd`) plus:

```
area_mm2            = region pixel area * 0.325^2 / 1e6
dens_all_per_mm2    = n_inside      / area_mm2
dens_cd45_per_mm2   = n_cd45_inside / area_mm2
dens_<lineage>_per_mm2 = n_<lineage> / area_mm2    (CD8T, CD4T, Treg, NK)

cd45_over_inside    = n_cd45_inside / n_inside
```

Densities are `NA` when the region is empty.

```{r periphery-coverage, echo=FALSE}
peri_union |>
  filter(threshold_um == 500) |>
  select(patient_id, region, source, um_per_unit, area_mm2,
         n_inside, n_cd45_inside, cd45_over_inside) |>
  arrange(patient_id, region) |>
  knitr::kable(digits = 3,
    caption = "Cells and area per region at d = 500 µm (union). n_inside = 0 ⇒ empty region.")
```

## Margin immunity vs clinical HOT score

Statistic: Spearman correlation (per threshold) between the margin-band value and
the numeric `HOT score`, for both:

```
fraction = cd45_over_inside   = n_cd45_inside / n_inside
density  = dens_cd45_per_mm2  = n_cd45_inside / area_mm2
```

```{r hot-join}
clin <- clinical_data |>
  mutate(patient_id = slide_key(`ID PATIENT`)) |>
  select(patient_id,
         hot_score  = any_of("HOT score"),
         immuno_phe = any_of("Immuno-phenotype")) |>
  mutate(hot_score = suppressWarnings(as.numeric(hot_score)))

margin_union <- peri_union |>
  filter(region == "margin") |>
  inner_join(clin, by = "patient_id")

cat("patients with margin metrics + clinical immune phenotype:",
    dplyr::n_distinct(margin_union$patient_id), "\n")
```

```{r hot-score-fraction, fig.height=4}
df <- margin_union |>
  filter(is.finite(cd45_over_inside), is.finite(hot_score))
if (nrow(df) > 2) {
  labs_tbl <- df |>
    group_by(threshold_um) |>
    group_modify(~ paired_spearman(.x$cd45_over_inside, .x$hot_score)) |>
    mutate(lab = sprintf("rho = %.2f\np = %.3g (n = %d)", rho, p, n))
  ggplot(df, aes(cd45_over_inside, hot_score)) +
    geom_point(alpha = 0.7, size = 2) +
    geom_smooth(method = "lm", se = FALSE, colour = "steelblue", formula = y ~ x) +
    geom_text(data = labs_tbl, aes(x = -Inf, y = Inf, label = lab),
              hjust = -0.05, vjust = 1.2, inherit.aes = FALSE, size = 3) +
    facet_wrap(~ threshold_um, labeller = labeller(threshold_um = ~ paste0("d = ", .x, " µm"))) +
    labs(title = "Invasive-margin immune fraction vs clinical HOT score",
         x = "CD45+ fraction in margin band", y = "clinical HOT score") +
    theme_classic(base_size = 11)
}
```

```{r hot-score-density, fig.height=4}
df <- margin_union |>
  filter(is.finite(dens_cd45_per_mm2), is.finite(hot_score))
if (nrow(df) > 2) {
  labs_tbl <- df |>
    group_by(threshold_um) |>
    group_modify(~ paired_spearman(.x$dens_cd45_per_mm2, .x$hot_score)) |>
    mutate(lab = sprintf("rho = %.2f\np = %.3g (n = %d)", rho, p, n))
  ggplot(df, aes(dens_cd45_per_mm2, hot_score)) +
    geom_point(alpha = 0.7, size = 2) +
    geom_smooth(method = "lm", se = FALSE, colour = "steelblue", formula = y ~ x) +
    geom_text(data = labs_tbl, aes(x = -Inf, y = Inf, label = lab),
              hjust = -0.05, vjust = 1.2, inherit.aes = FALSE, size = 3) +
    facet_wrap(~ threshold_um, labeller = labeller(threshold_um = ~ paste0("d = ", .x, " µm"))) +
    labs(title = "Invasive-margin immune density vs clinical HOT score",
         x = "CD45+ per mm² in margin band", y = "clinical HOT score") +
    theme_classic(base_size = 11)
}
```

## Margin immunity by clinical immuno-phenotype

Distribution of the margin-band `cd45_over_inside` within each `Immuno-phenotype`
group (ordered cold -> intermediate -> hot), faceted by threshold.

```{r phenotype-boxplot, fig.height=4}
df <- margin_union |>
  filter(!is.na(immuno_phe), is.finite(cd45_over_inside)) |>
  mutate(immuno_phe = hotcold_order(immuno_phe))
if (dplyr::n_distinct(na.omit(df$immuno_phe)) > 1 && nrow(df) > 3) {
  ggplot(df, aes(immuno_phe, cd45_over_inside, colour = immuno_phe)) +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.2, alpha = 0.6) +
    facet_wrap(~ threshold_um, labeller = labeller(threshold_um = ~ paste0("d = ", .x, " µm"))) +
    scale_colour_manual(values = hotcold_cols(levels(df$immuno_phe)),
                        name = "Immuno-phenotype") +
    labs(title = "Invasive-margin immune fraction by clinical immuno-phenotype",
         x = NULL, y = "CD45+ fraction in margin band") +
    theme_classic(base_size = 11) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
}
```

## Margin-vs-core exclusion score

Per patient at `d = 500` microns:

```
exclusion_frac = cd45_over_inside (margin)  -  cd45_over_inside (core)
exclusion_dens = dens_cd45_per_mm2 (margin) -  dens_cd45_per_mm2 (core)
```

Positive means immunity is concentrated at the border.

```{r margin-core}
mc <- peri_union |>
  filter(threshold_um == 500) |>
  select(patient_id, region, cd45_over_inside,
         dens_cd45_per_mm2, dens_CD8T_per_mm2, n_inside) |>
  pivot_wider(names_from = region,
              values_from = c(cd45_over_inside, dens_cd45_per_mm2, dens_CD8T_per_mm2, n_inside)) |>
  mutate(
    exclusion_frac = cd45_over_inside_margin - cd45_over_inside_core,
    exclusion_dens = dens_cd45_per_mm2_margin - dens_cd45_per_mm2_core
  ) |>
  inner_join(clin, by = "patient_id")

mc |>
  select(patient_id, immuno_phe,
         margin_cd45 = cd45_over_inside_margin, core_cd45 = cd45_over_inside_core,
         exclusion_frac, hot_score) |>
  arrange(desc(exclusion_frac)) |>
  knitr::kable(digits = 3,
    caption = "Margin-vs-core CD45+ fraction at d = 500 µm. Positive exclusion_frac ⇒ border-concentrated immunity.")
```

```{r margin-core-scatter, fig.height=4.5}
sc <- mc |> filter(is.finite(cd45_over_inside_margin), is.finite(cd45_over_inside_core))
if (nrow(sc) > 1) {
  aes_col <- if ("immuno_phe" %in% names(sc) && any(!is.na(sc$immuno_phe)))
    aes(cd45_over_inside_core, cd45_over_inside_margin, colour = hotcold_order(immuno_phe))
  else aes(cd45_over_inside_core, cd45_over_inside_margin)
  p <- ggplot(sc, aes_col) +
    geom_abline(slope = 1, linetype = "dashed", colour = "grey50") +
    geom_point(size = 3, alpha = 0.85) +
    coord_equal() +
    labs(title = "Margin vs core CD45+ fraction (d = 500 µm)",
         x = "CD45+ fraction, deep core", y = "CD45+ fraction, invasive margin",
         colour = "Immuno-phenotype") +
    theme_classic(base_size = 11)
  if ("colour" %in% names(sc) || any(!is.na(sc$immuno_phe)))
    p <- p + scale_colour_manual(values = hotcold_cols(levels(hotcold_order(sc$immuno_phe))),
                                 name = "Immuno-phenotype", na.value = "grey70")
  print(p)
}
```

```{r exclusion-by-phenotype, fig.height=4}
df <- mc |>
  filter(!is.na(immuno_phe), is.finite(exclusion_frac)) |>
  mutate(immuno_phe = hotcold_order(immuno_phe))
if (dplyr::n_distinct(na.omit(df$immuno_phe)) > 1 && nrow(df) > 3) {
  ggplot(df, aes(immuno_phe, exclusion_frac, colour = immuno_phe)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey50") +
    geom_boxplot(outlier.shape = NA) +
    geom_jitter(width = 0.2, alpha = 0.6) +
    scale_colour_manual(values = hotcold_cols(levels(df$immuno_phe)), name = "Immuno-phenotype") +
    labs(title = "Margin-minus-core exclusion score by clinical phenotype (d = 500 µm)",
         x = NULL, y = "exclusion score (margin − core CD45+ fraction)") +
    theme_classic(base_size = 11) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
}
```

## Single annotations vs union

Per-annotation margin fraction (each single-annotation core) vs the patient's union
margin fraction, at `d = 500` microns — a stability check across a patient's
separately drawn regions.

```{r per-ann-vs-union, fig.height=4.5}
pa <- peri_per_ann |>
  filter(region == "margin", threshold_um == 500) |>
  select(patient_id, annotation, cd45_over_inside)
un <- peri_union |>
  filter(region == "margin", threshold_um == 500) |>
  select(patient_id, union_cd45 = cd45_over_inside)
cmp <- inner_join(pa, un, by = "patient_id") |>
  filter(is.finite(cd45_over_inside), is.finite(union_cd45))
if (nrow(cmp) > 1) {
  ggplot(cmp, aes(patient_id, cd45_over_inside)) +
    geom_line(aes(y = union_cd45, group = 1), colour = "grey60") +
    geom_point(aes(y = union_cd45), colour = "grey40", size = 3, shape = 18) +
    geom_jitter(aes(colour = annotation), width = 0.05, size = 2, alpha = 0.85) +
    labs(title = "Margin immune fraction: single annotations vs union (d = 500 µm)",
         x = "patient", y = "CD45+ fraction in margin band", colour = "annotation") +
    theme_classic(base_size = 11) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
}
```

## Margin composition (three denominators)

The same three-denominator composition used inside the tumour, applied to the
margin band at `d = 500` microns:

```
frac_inside = n_<lineage> / n_inside         (of all cells in band)
frac_tumor  = n_<lineage> / n_tumor_inside   (of tumour cells in band)
frac_cd45   = n_<lineage> / n_cd45_inside    (of CD45+ cells in band)
```

```{r margin-composition, fig.height=5}
comp <- peri_union |>
  filter(region == "margin", threshold_um == 500) |>
  region_composition() |>
  pivot_longer(c(frac_inside, frac_tumor, frac_cd45),
               names_to = "normalization", values_to = "frac") |>
  mutate(normalization = factor(
           c(frac_inside = "of all cells", frac_tumor = "of tumour cells",
             frac_cd45 = "of CD45+ cells")[normalization],
           levels = c("of all cells", "of tumour cells", "of CD45+ cells"))) |>
  filter(is.finite(frac))

if (nrow(comp) > 0) {
  ggplot(comp, aes(lineage, frac, fill = lineage)) +
    geom_boxplot(outlier.shape = NA, colour = "grey30") +
    geom_jitter(width = .15, alpha = .5, size = 1) +
    facet_wrap(~ normalization, scales = "free_y") +
    scale_fill_manual(values = oi, guide = "none") +
    labs(title = "Immune composition in the invasive-margin band (d = 500 µm)",
         x = NULL, y = "fraction") +
    theme_classic(base_size = 11) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
}
```

## Margin cell-type fractions: hot vs cold

Same margin-band composition (d = 500 µm, union), now split by the clinical
`Immuno-phenotype`: for each lineage, the boxes contrast **hot** vs **cold** tumours.

```
group(patient) = hot / intermediate / cold   (from clinical Immuno-phenotype)
compared per lineage and per denominator (of all cells / tumour cells / CD45+ cells)
```

```{r margin-composition-hotcold, fig.height=5}
comp_hc <- peri_union |>
  filter(region == "margin", threshold_um == 500) |>
  region_composition() |>
  inner_join(clin, by = "patient_id") |>
  filter(!is.na(immuno_phe)) |>
  mutate(immuno_phe = hotcold_order(immuno_phe)) |>   # cold -> intermediate -> hot
  pivot_longer(c(frac_inside, frac_tumor, frac_cd45),
               names_to = "normalization", values_to = "frac") |>
  mutate(normalization = factor(
           c(frac_inside = "of all cells", frac_tumor = "of tumour cells",
             frac_cd45 = "of CD45+ cells")[normalization],
           levels = c("of all cells", "of tumour cells", "of CD45+ cells"))) |>
  filter(is.finite(frac))

if (nrow(comp_hc) > 0 && dplyr::n_distinct(na.omit(comp_hc$immuno_phe)) > 1) {
  ggplot(comp_hc, aes(lineage, frac, colour = immuno_phe)) +
    geom_boxplot(outlier.shape = NA, position = position_dodge(width = .75)) +
    geom_point(position = position_jitterdodge(jitter.width = .12, dodge.width = .75),
               alpha = .6, size = 1) +
    facet_wrap(~ normalization, scales = "free_y") +
    scale_colour_manual(values = hotcold_cols(levels(comp_hc$immuno_phe)),
                        name = "Immuno-phenotype") +   # hot = red, cold = light blue
    labs(title = "Invasive-margin cell-type fractions: hot vs cold (d = 500 µm)",
         subtitle = sprintf("%d patients with a margin band and an immuno-phenotype",
                            dplyr::n_distinct(comp_hc$patient_id)),
         x = NULL, y = "fraction in margin band") +
    theme_classic(base_size = 11) +
    theme(axis.text.x = element_text(angle = 30, hjust = 1))
}
```

## Notes

- Band width is physical: each polygon is buffered by `d / 0.325` pixel units; the
  `um_per_unit` column echoes the 0.325 µm/px conversion used.
- Geojson-only: a margin band needs a real boundary, so only patients with a
  pathologist geojson are covered (no `Out_of_annotation` fallback).
- Fraction vs density: a thin band can read immune-rich by fraction yet sparse by
  density (per mm²) — the Immunoscore unit.
- Small cohort; Spearman with reported `n` throughout.
